---
title: "Getting started with classydms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with classydms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette walks through the full `classydms` pipeline at a high level: load raw GC-DMS data, preprocess, generate synthetic data, pretrain the encoder, classify with elastic net, and produce a spatial saliency map for biomarker localization.

```{r setup}
library(classydms)
library(purrr)   # for map(), used in some downstream steps
```

## 1. Load raw data from a directory

The expected file layout is one file per subject, with disease and control samples in separate folders. Each file should be tab-delimited with the GC-DMS layout described in `?read_dms_file` (line 1 = "Vc", line 2 = tab + CV values, line 3 = header, lines 4+ = time + intensities).

```{r load}
# One-shot loader: reads all files matching the POS-channel pattern,
# skips broken ones, names each sample after its file basename (so
# results stay traceable to subjects), and returns a (samples, y) pair.
disease_dir = here::here("sampledata/influenza")
control_dir = here::here("sampledata/controls")

loaded <- load_dms_directory(
  disease_dir = disease_dir,
  control_dir = control_dir,
  pattern     = "POS.*\\.(xls|txt|tsv|csv)$"  # POS channel only — see note
)

samples <- loaded$samples   # named list, each element has $path, $time, $cv, $Z
y       <- loaded$y         # named integer vector: 2 = disease, 1 = control

# Sample names map to file basenames without extensions, so you can:
samples[["241115-S1994-S1-DMS1_POS"]]    # index by subject
names(samples)                            # see the full list of subjects

# Sanity check
stopifnot(is.list(samples), length(samples) > 0,
          is.matrix(samples[[1]]$Z),
          all(nzchar(names(samples))))   # no empty-string names
```

**Channel choice**: classifiers in this package train on the **positive ion channel only** (POS files). Negative-channel data is not used. The default `pattern` enforces this — pass an explicit `pattern` argument only if you need a different channel or filename convention.

If you have a non-standard file format, write your own loader that produces the same list shape — each element must have `$path`, `$time`, `$cv`, and `$Z`, and the outer list should have meaningful (non-empty) names. As long as that contract is met, the rest of the pipeline works unchanged.

## 2. Preprocess

```{r preprocess}
# Savitzky-Golay smoothing + AsLS baseline correction + clip below zero
samples_proc <- lapply(samples, process_one_sample,
                        sg_p = 3, sg_n = 21,
                        asls_lambda = 1e6, asls_p = 0.01, asls_niter = 10)

# Dust threshold each sample to a fixed floor (e.g. global value, or
# per-sample via robust_eps). For a fixed floor:
dust_thr <- 0.005
samples_dust <- lapply(samples_proc, baseline_basement,
                        basement_thr = dust_thr)

# Cohort-wide trim bounds
b_list <- lapply(samples_dust,
                  function(s) trim_bounds_from_occupancy(s$Z))
R0 <- max(sapply(samples_dust, function(s) nrow(s$Z)))
C0 <- max(sapply(samples_dust, function(s) ncol(s$Z)))
global_b <- pooled_trim_bounds(b_list, R = R0, C = C0,
                                q_lo = 0.05, q_hi = 0.95)
samples_trimmed <- lapply(samples_dust, trim_sample, bounds = global_b)

```

## 3. Quality control: flag and exclude junk samples

Before estimating peak parameters or pretraining, run per-sample QC on the trimmed samples. This catches samples with too little signal, suspiciously high or low summary statistics, or unusual morphology relative to the rest of the cohort.

```{r qc}
# Compute per-sample QC features. compute_cohort_correlation = TRUE adds
# a multivariate similarity check (slower, but catches morphological
# outliers that don't trigger univariate IQR).
qc <- qc_features(samples_trimmed, compute_cohort_correlation = TRUE)
print(qc)

# Apply default exclusion criteria (low signal, IQR outliers, low
# cohort correlation). Pass NA or Inf for any threshold to disable.
qc_flagged <- flag_junk_samples(qc,
                                 min_prop_nonzero       = 0.10,
                                 iqr_mult               = 3,
                                 min_cohort_correlation = 0.30)

# Visualize: each panel shows one QC feature with flagged samples in red.
# Look for outliers and adjust thresholds if the defaults are too strict
# or too lenient for your cohort.
plot_qc(qc_flagged)

# How many samples were flagged?
message("Flagged: ", sum(qc_flagged$flagged), " of ", nrow(qc_flagged))
```

If you're satisfied with the QC decisions, exclude flagged samples from all downstream steps:

```{r qc-exclude}
keep <- !qc_flagged$flagged
samples_kept <- samples_trimmed[keep]
y_kept       <- y[keep]
```

If any of the QC criteria seem too aggressive for your cohort (e.g., you want to keep moderate outliers), re-run `flag_junk_samples()` with adjusted thresholds:

```{r qc-relax, eval = FALSE}
# More permissive: IQR outlier threshold of 5 instead of 3
qc_flagged_loose <- flag_junk_samples(qc, iqr_mult = 5)
```

## 4. Normalize and pad (post-QC)

After QC exclusion, finish preprocessing by log-quantile normalizing and padding to a common size for the encoder.

```{r normalize-pad}
Z_norm   <- map(samples_kept, ~normalize_sample(.x$Z))
pad_dims <- compute_pad_targets(Z_norm)
Z_padded <- map(Z_norm, ~pad_to_target(.x, pad_dims$H, pad_dims$W))
```

## 5. Estimate peak parameters from real data (label-agnostic)

```{r peak-params}
Z_raw     <- map(samples_kept, ~.x$Z)
Z_pretrim <- map(samples_kept,
                  ~if (!is.null(.x$Z_pretrim)) .x$Z_pretrim else .x$Z)

peak_params <- estimate_peak_params(Z_raw,
                                     Z_pretrim_list = Z_pretrim,
                                     top_k = 2500,
                                     frac_height = 0.5)
```

`top_k = 2500` (rather than a smaller value like 150) gives `estimate_peak_params` enough capacity to characterize the full intensity distribution including dim peaks. Lower values bias the intensity log-normal toward the brightest peaks and make synthetic samples systematically too bright.

## 6. Pretrain the encoder

For most users, the online variant is recommended because it removes the RAM ceiling and lets the model see hundreds of thousands of synthetic samples:

```{r pretrain}
pretrain_result <- pretrain_denoising_online(
  peak_params,
  H = pad_dims$H, W = pad_dims$W,
  steps_per_epoch  = 500L,
  epochs           = 32L,
  batch_size       = 64L,
  num_workers      = 8L,            # parallel R-side data generation
  dust_threshold   = 0.005,         # match real-data dust threshold
  size_jitter      = 0.15,          # peak size variation; tune via QC
  device           = "mps",         # "cpu" or "cuda" on non-Apple systems
  save_path        = "encoder.pt",
  seed             = 42L
)
```

This writes `encoder.pt`, `encoder_autoencoder.pt`, and `encoder_manifest.Rdata` to the working directory, plus periodic checkpoints. The manifest contains the full hyperparameter configuration and loss curves for reproducibility.

See `vignette("pretraining", package = "classydms")` for the rationale behind these specific hyperparameters and how to tune them for your cohort.

Alternatively, skip pretraining entirely by downloading an encoder a collaborator has already trained:

```{r load-pretrained}
encoder <- load_pretrained_encoder("flu_v1")
```

## 7. Reconstruction QC: does the autoencoder generalize to real data?

A critical check before classification: feed real samples through the **full autoencoder** (encoder + decoder) and visually compare the reconstructions to the originals. If reconstructions look similar to the originals, the encoder learned features that transfer from synthetic to real. If reconstructions are blank or smeared, the synthetic distribution didn't match real closely enough during pretraining.

```{r reconstruct}
# Reconstruct a handful of real samples through the trained autoencoder
recon <- reconstruct_real_samples(
  autoencoder    = pretrain_result$autoencoder,
  Z_list         = Z_padded,
  sample_indices = 1:6
)

# Side-by-side heatmaps: row 1 = real, row 2 = reconstruction
par(mfrow = c(2, 6), mar = c(1, 1, 2, 1))
for (r in recon) {
  image(t(r$original)[, nrow(r$original):1],
        col = hcl.colors(100, "YlOrRd", rev = TRUE),
        axes = FALSE,
        main = paste0("real #", r$index))
}
for (r in recon) {
  image(t(r$reconstructed)[, nrow(r$reconstructed):1],
        col = hcl.colors(100, "YlOrRd", rev = TRUE),
        axes = FALSE,
        main = sprintf("recon (MSE=%.3f)", r$mse))
}
par(mfrow = c(1, 1))

# Aggregate MSE across all samples
all_recon <- reconstruct_real_samples(pretrain_result$autoencoder,
                                       Z_padded,
                                       sample_indices = seq_along(Z_padded))
mse_per_sample <- sapply(all_recon, function(r) r$mse)
summary(mse_per_sample)
```

**What healthy reconstructions look like:**

- Peak *locations* in the reconstruction match peak locations in the original. Bright spots appear in the right (RT, CV) coordinates, even if intensities are dampened.
- The overall *layout* — where signal is concentrated vs. where it's sparse — is preserved.
- Mean MSE across all samples is low (typically < 0.05 for normalized data).

**Warning signs:**

- Reconstructions are uniformly blank or near-zero everywhere → encoder collapsed to outputting an average image
- Reconstructions have peaks in completely wrong locations → encoder didn't learn spatial structure
- A few specific samples have much higher MSE than the rest → those samples are morphological outliers (often worth flagging in QC)

If reconstructions look poor, **do not proceed to classification** — the encoder features won't be meaningful. Common causes and fixes:

- Synthetic-real distribution gap: re-check with `synthetic_quality_check()` (see `vignette("pretraining")`), adjust `dust_threshold` and `size_jitter`
- Insufficient training: increase `steps_per_epoch * epochs` and retrain
- Encoder mode collapse: check that encoder feature variance is meaningful (most channels should vary across samples)

## 8. Classify with elastic net on encoder features

```{r classify}
elnet <- cv_encoder_elastic_net(pretrain_result$encoder,
                                 Z_list = Z_padded, y = y_kept,
                                 k = 5, alpha = 0.5)
elnet$auc
elnet$coefs
```

## 9. Saliency maps

```{r saliency}
disease_idx <- which(y_kept == 2)
orig_dims <- lapply(samples_kept,
                     function(s) c(H = nrow(s$Z), W = ncol(s$Z)))

sal_disease <- aggregate_saliency_masked(pretrain_result$encoder,
                                          Z_padded, coefs = elnet$coefs,
                                          sample_indices = disease_idx,
                                          orig_dims = orig_dims)
```

The result is a 2-D matrix in padded image coordinates. Map back to physical (RT in seconds, CV in volts) using `samples_kept[[1]]$time` and `samples_kept[[1]]$cv`. Saliency hotspots correspond to (RT, CV) locations whose encoder activation, weighted by elastic net coefficients, most strongly drives the disease prediction. These coordinates are the candidate biomarker compounds.

## Troubleshooting

### "attempt to use zero-length variable name"

This error means a list somewhere in your pipeline has an **empty-string** name (`""`), and a downstream operation is trying to use that name. Two common causes:

1. The `samples` list was built via `purrr::map()` + `purrr::keep()` (or `c()` of filtered lists) and inherited empty-string names. **Fix**: replace any empty names with meaningful ones, or drop them entirely. `load_dms_directory()` names samples after their file basenames and explicitly verifies no name is empty.
2. A sample's `Z` matrix has empty-string column or row names. **Fix**: `dimnames(Z) <- NULL`. The `read_dms_file()` function does this for you.

If you wrote a custom loader, give each sample a meaningful name (e.g. file basename, subject ID) and make sure the `Z` matrices don't carry empty dimnames. Keeping non-empty names is important — you'll want them for tracing classification results and saliency maps back to specific subjects.

