---
title: "Case/control simulation — end-to-end pipeline validation with ground truth"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7.5, fig.height = 4, dpi = 96)
library(classydms)
library(ggplot2)
library(dplyr)
library(tibble)
library(purrr)

# Consistent heatmap helper (log = TRUE compresses the dynamic range so
# dim structure is visible next to bright peaks)
plot_mat <- function(mat, title = "", low = "white", high = "#B2182B",
                     limits = NULL, log = FALSE) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  scale <- if (log) {
    scale_fill_gradient(low = low, high = high, limits = limits,
                        trans = "log1p", name = "log(1+intensity)")
  } else {
    scale_fill_gradient(low = low, high = high, limits = limits,
                        name = "intensity")
  }
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale +
    scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 11) +
    theme(panel.grid = element_blank())
}

# Diverging-scale helper for difference maps
plot_mat_diff <- function(mat, title = "", limits = NULL,
                          low = "#2166AC", high = "#B2182B") {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient2(low = low, mid = "white", high = high,
                          midpoint = 0, limits = limits,
                          name = "case - control") +
    scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 11) +
    theme(panel.grid = element_blank())
}
```

# Summary

This walkthrough validates the full pipeline end-to-end by injecting known biomarker peaks into a synthetic case group and measuring both classification and spatial recovery. Sections:

1. **Purpose** — Motivates the ground-truth simulation and states the AUC and biomarker-recovery acceptance gates.
2. **Load a base cohort's `peak_params`** — Runs the preprocessing chain on real flu samples and estimates the empirical peak pool.
3. **Hold out biomarkers for Level 1** — Picks one pool peak as the biomarker and strips its neighborhood from the background pool.
4. **Simulate the cohort** — Simulates 30 cases + 30 controls and reports per-biomarker firing rates by class.
5. **Visualize a case and a control side by side** — Plots log-scale case/control panels and a case-minus-control diff to make the injected peak visible.
6. **Preprocess and pad the simulated cohort** — Normalizes and pads simulated samples to a common encoder input size.
7. **Pretraining and classification (heavy — flip `eval = TRUE` to run)** — Optional pretraining, 5-fold CV AUC, and biomarker-recovery AUC from aggregated saliency.
8. **Fake-saliency demo — always runs** — Sanity-checks `evaluate_biomarker_recovery` on a hand-built Gaussian-blob saliency and a uniform-noise baseline.
9. **Next steps — Levels 2 and 3** — Lists expected metric ranges for progressively harder `biomarkers_level2/3()` scenarios.

# Purpose

Every step of the classydms pipeline is validated in isolation (report on real cohorts, tutorial on toy params). But there's no ground truth for whether the **full pipeline** — preprocess → params → pretrain → classify → saliency — can recover a signal we *know* is there.

> **This tutorial uses the peak_params flow** (each sample independently draws random peak positions from a common pool). A newer alternative — the **catalog flow** — matches real GC-DMS structure more closely (peak positions are conserved across samples, only prevalence varies). If you're building a new analysis today, prefer the catalog flow: see `catalog_pipeline_sanity_check` and `catalog_synthesis_walkthrough`. The peak_params flow documented here is still valid and useful as a reference.

This tutorial injects known "biomarker" peaks into a synthetic case group and asks two questions:

1. **Classification**: can the pipeline distinguish cases from controls? (Metric: cross-validated AUC.)
2. **Biomarker recovery**: does the spatial saliency map localize the injected biomarkers? (Metric: per-pixel AUC of saliency values against a ground-truth mask around each biomarker.)

If Level 1 (single strong biomarker) can't hit AUC ≥ 0.80 on either metric, the pipeline has a bug and we should stop. If it can, Levels 2 and 3 stress-test performance under progressively more realistic conditions.

# 1. Load a base cohort's `peak_params`

The base peak distribution should come from a real cohort so backgrounds look realistic. Here we use the small `sampledata/influenza` slice that ships with the package (adjust paths for your own cohort).

Adjust `disease_dir` / `control_dir` to point at any real GC-DMS cohort you have on disk; the pooled `peak_params` from those samples defines "background" for the simulation.

```{r load-base}
# here::here() resolves paths relative to the project root so the tutorial
# knits regardless of the current working directory.
disease_dir <- if (requireNamespace("here", quietly = TRUE)) {
  here::here("sampledata", "influenza")
} else "sampledata/influenza"
control_dir <- if (requireNamespace("here", quietly = TRUE)) {
  here::here("sampledata", "influenza_controls")
} else "sampledata/influenza_controls"

loaded <- load_dms_directory(
  disease_dir = disease_dir,
  control_dir = control_dir,
  pattern     = "POS.*\\.(xls|txt|tsv|csv)$"
)
samples_raw <- loaded$samples
message("Loaded ", length(samples_raw), " samples")

samples_proc <- map(samples_raw, ~process_one_sample(.x,
  sg_p = 3, sg_n = 21, asls_lambda = 1e6, asls_p = 0.01, asls_niter = 10))
samples_dust <- map(samples_proc, ~baseline_basement(.x, basement_thr = 0.005))
b_list <- map(samples_dust, ~trim_bounds_from_occupancy(.x$Z))
R0 <- max(sapply(samples_dust, function(s) nrow(s$Z)))
C0 <- max(sapply(samples_dust, function(s) ncol(s$Z)))
global_b <- pooled_trim_bounds(b_list, R = R0, C = C0,
                                q_lo = 0.05, q_hi = 0.95)
samples_trim <- map(samples_dust, ~trim_sample(.x, bounds = global_b))
Z_raw     <- map(samples_trim, ~.x$Z)
Z_pretrim <- map(samples_trim, ~if (!is.null(.x$Z_pretrim)) .x$Z_pretrim else .x$Z)

pp <- estimate_peak_params(Z_raw, Z_pretrim_list = Z_pretrim,
                            top_k = 500, frac_height = 0.5,
                            verbose = FALSE)
H_native <- nrow(Z_raw[[1]]); W_native <- ncol(Z_raw[[1]])
sprintf("Base pool: %d peaks; native image %d x %d",
        pp$n_peaks_detected, H_native, W_native)
```

# 2. Hold out biomarkers for Level 1

`biomarkers_level1()` picks 1 random peak from the pool as the biomarker template, sets case prevalence to 95% / control 5%, and case intensity multiplier to ×2.0. The peak *and any pool peaks within a protected radius* are stripped from the returned `background_peak_params`, so backgrounds sampled from it never land near the biomarker.

```{r holdout}
scenario <- biomarkers_level1(pp, H = H_native, W = W_native, seed = 42L)
scenario$biomarkers
sprintf("Pool: %d peaks -> %d after holding out biomarker + protected radius",
        pp$n_peaks_detected,
        scenario$background_peak_params$n_peaks_detected)
```

# 3. Simulate the cohort

30 cases + 30 controls should be enough to see a Level 1 signal without slow pretraining.

```{r simulate}
sim <- simulate_case_control_cohort(
  scenario$background_peak_params,
  scenario$biomarkers,
  n_cases    = 30L,
  n_controls = 30L,
  H = H_native, W = W_native,
  seed = 7L
)
table(sim$y, dnn = "class (2 = case, 1 = control)")

# Per-biomarker firing rate by class — sanity check that prevalences match design
firing <- as.data.frame(sim$per_sample_firing)
colnames(firing) <- paste0("biomarker_", seq_len(ncol(sim$per_sample_firing)))
firing$class <- factor(sim$y, levels = c(1, 2), labels = c("control", "case"))
firing %>%
  tidyr::pivot_longer(cols = starts_with("biomarker_"),
                      names_to = "biomarker", values_to = "fired") %>%
  group_by(class, biomarker) %>%
  summarise(prevalence = mean(fired), .groups = "drop")
```

# 4. Visualize a case and a control side by side

The biomarker location for Level 1:

```{r plot-biomarker-loc}
biom <- scenario$biomarkers
b_r <- biom$rt_loc * H_native
b_c <- biom$cv_loc * W_native
sprintf("Biomarker at RT pixel %.0f, CV pixel %.0f", b_r, b_c)
```

Compare a case with the biomarker firing against a control where it doesn't. The absolute intensity of a single injected peak is small relative to the brightest background peaks, so the linear-scale view often makes the biomarker invisible. We plot three views to make the signal visible:

- **Log-scale case and control panels** — compress the dynamic range so dim structure and the biomarker are both readable.
- **Case − control difference panel** — a per-sample subtraction on a diverging color scale (red = case brighter, blue = control brighter). Injected biomarkers show up as a red blob at their location regardless of absolute intensity.

```{r visualize-samples, fig.height = 6.5}
# Find a case and a control that fired appropriately
case_idx    <- which(sim$y == 2L & sim$per_sample_firing[, 1])[1]
control_idx <- which(sim$y == 1L & !sim$per_sample_firing[, 1])[1]

Z_case    <- sim$samples[[case_idx]]$Z
Z_control <- sim$samples[[control_idx]]$Z
Z_diff    <- Z_case - Z_control

# Overlay a marker at the biomarker location
mark_biomarker <- function(p, r, c) {
  p + annotate("point", x = c, y = r, shape = 21, color = "cyan",
               fill = NA, size = 8, stroke = 1.2)
}

lim_log <- c(0, max(c(Z_case, Z_control)))
lim_diff <- c(-1, 1) * max(abs(Z_diff))

gridExtra::grid.arrange(
  mark_biomarker(plot_mat(Z_case, "Case (biomarker fired) — log scale",
                          limits = lim_log, log = TRUE),
                 b_r, b_c),
  mark_biomarker(plot_mat(Z_control, "Control (biomarker absent) — log scale",
                          limits = lim_log, log = TRUE),
                 b_r, b_c),
  mark_biomarker(plot_mat_diff(Z_diff, "Case − Control (diverging)",
                                limits = lim_diff),
                 b_r, b_c),
  ncol = 3
)
```

The cyan circle marks the biomarker location. On the diff panel, the injected biomarker shows up as a red blob (case has extra signal there); the blue speckle elsewhere is the random per-sample variation in which background peaks were drawn — that's noise as far as the classifier is concerned, and should average out across many samples.

Averaging the diff over ALL cases and controls makes the systematic signal even clearer:

```{r average-diff, fig.height = 4}
case_indices    <- which(sim$y == 2L)
control_indices <- which(sim$y == 1L)
Z_case_mean    <- Reduce(`+`, lapply(sim$samples[case_indices],    `[[`, "Z")) / length(case_indices)
Z_control_mean <- Reduce(`+`, lapply(sim$samples[control_indices], `[[`, "Z")) / length(control_indices)
Z_diff_mean    <- Z_case_mean - Z_control_mean

mark_biomarker(
  plot_mat_diff(Z_diff_mean,
                sprintf("mean(case) − mean(control), N=%d each",
                        length(case_indices)),
                limits = c(-1, 1) * max(abs(Z_diff_mean))),
  b_r, b_c)
```

If the pipeline is correctly injecting the biomarker, the averaged diff will show a red blob centered on the biomarker (cyan circle) and near-zero elsewhere — because random per-sample background variation averages out but the systematic biomarker signal does not.

# 5. Preprocess and pad the simulated cohort

The simulated samples were produced clean — no baseline drift, no dust threshold — so preprocessing largely passes them through. We still normalize and pad to make sure they enter the encoder at the same size as any real cohort would.

```{r preprocess-sim}
sim_Z_list <- map(sim$samples, ~.x$Z)
Z_norm  <- map(sim_Z_list, normalize_sample)
pad_dims <- compute_pad_targets(Z_norm)
Z_padded <- map(Z_norm, ~pad_to_target(.x, pad_dims$H, pad_dims$W))
sprintf("Padded image size: %d x %d", pad_dims$H, pad_dims$W)
```

# 6. Pretraining and classification (heavy — flip `eval = TRUE` to run)

These chunks are `eval = FALSE` by default because pretraining takes ~5–15 minutes even on a small cohort. Set `PRETRAIN <- TRUE` to run everything downstream.

```{r pretrain-flag}
PRETRAIN <- FALSE   # flip to TRUE to actually run pretraining + eval
```

```{r pretrain, eval = PRETRAIN}
# Peak params estimated on the *simulated* cohort — this is what a real
# analysis would do, treating the simulation as if it were unknown data.
sim_pp <- estimate_peak_params(sim_Z_list,
                                top_k = 2500, frac_height = 0.5,
                                verbose = FALSE)

pretrain_result <- pretrain_denoising_online(
  sim_pp,
  H = pad_dims$H, W = pad_dims$W,
  steps_per_epoch  = 100L,       # short for tutorial speed
  epochs           = 10L,        # short
  batch_size       = 32L,
  num_workers      = 1L,
  dust_threshold   = 0.005,
  size_jitter      = 0.15,
  device           = "cpu",
  seed             = 1L
)
```

## 6a. Classification: case vs control AUC

```{r classify, eval = PRETRAIN}
elnet <- cv_encoder_elastic_net(pretrain_result$encoder,
                                 Z_list = Z_padded, y = sim$y,
                                 k = 5, alpha = 0.5)
sprintf("Case/control AUC (5-fold CV): %.3f", elnet$auc)
```

**Level 1 acceptance target: AUC ≥ 0.80.** Below that, stop and investigate — a Level 1 signal that strong should be detectable. Between 0.80 and 0.90 is acceptable and moves us to Level 2. Above 0.95 is expected but not required.

## 6b. Biomarker recovery: saliency + spatial AUC

```{r saliency, eval = PRETRAIN}
case_idx_all <- which(sim$y == 2L)
orig_dims <- lapply(sim$samples,
                     function(s) c(H = nrow(s$Z), W = ncol(s$Z)))

sal <- aggregate_saliency_masked(pretrain_result$encoder,
                                  Z_padded, coefs = elnet$coefs,
                                  sample_indices = case_idx_all,
                                  orig_dims = orig_dims)
str(sal, max.level = 1)

# The saliency is in *native* image coordinates (pre-padding). The
# biomarker rt_loc / cv_loc are also fractions of the native H_native,
# W_native, so we evaluate in native coordinates directly.
recovery <- evaluate_biomarker_recovery(
  sal, scenario$biomarkers,
  H = H_native, W = W_native,
  match_radius_rt = 8L, match_radius_cv = 2L,
  top_k = 5L
)
sprintf("Biomarker recovery — per-pixel AUC: %.3f", recovery$auc_per_pixel)
sprintf("Biomarker recovery — recall @ top-5: %.3f", recovery$recall_at_k)
sprintf("Biomarker recovery — precision @ top-5: %.3f", recovery$precision_at_k)
```

# 7. Fake-saliency demo — always runs

Since pretraining is expensive, the eval below uses a *synthesized* saliency map as a placeholder — a Gaussian blob at each biomarker location plus uniform noise. This confirms that `evaluate_biomarker_recovery()` returns sensible metrics.

```{r fake-saliency, fig.height = 4}
biom <- scenario$biomarkers
fake_sal <- matrix(0, nrow = H_native, ncol = W_native)
for (b in seq_len(nrow(biom))) {
  mu_r <- biom$rt_loc[b] * H_native
  mu_c <- biom$cv_loc[b] * W_native
  sig_r <- biom$sigma_rt[b]; sig_c <- biom$sigma_cv[b]
  g_r <- exp(-0.5 * ((seq_len(H_native) - mu_r) / sig_r)^2)
  g_c <- exp(-0.5 * ((seq_len(W_native) - mu_c) / sig_c)^2)
  fake_sal <- fake_sal + outer(g_r, g_c)
}
set.seed(1)
fake_sal <- fake_sal + matrix(runif(H_native * W_native, 0, 0.05),
                               H_native, W_native)

rec <- evaluate_biomarker_recovery(fake_sal, biom,
                                    H = H_native, W = W_native,
                                    top_k = 5L)
sprintf("[fake saliency] per-pixel AUC: %.3f", rec$auc_per_pixel)
sprintf("[fake saliency] recall @ top-5: %.3f", rec$recall_at_k)

plot_mat(fake_sal, title = "Fake saliency (Gaussian blob at biomarker + uniform noise)") +
  annotate("point", x = b_c, y = b_r, shape = 21, color = "cyan",
           fill = NA, size = 8, stroke = 1.2)
```

Random-baseline sanity: uniform noise saliency should return AUC ≈ 0.5.

```{r fake-saliency-chance}
set.seed(2)
noise_sal <- matrix(runif(H_native * W_native), H_native, W_native)
rec_chance <- evaluate_biomarker_recovery(noise_sal, biom,
                                           H = H_native, W = W_native,
                                           top_k = 5L)
sprintf("[chance] per-pixel AUC: %.3f  (should be near 0.5)",
        rec_chance$auc_per_pixel)
```

# 8. Next steps — Levels 2 and 3

Once Level 1 passes the AUC ≥ 0.80 gates on both metrics, run the same pipeline with the harder scenarios by replacing `biomarkers_level1()` with `biomarkers_level2()` or `biomarkers_level3()`. Expected results:

- **Level 2** (3 biomarkers, 80/20 prevalence, ×1.5 intensity): case/control AUC ≥ 0.85, biomarker recovery AUC ≥ 0.75
- **Level 3** (6 biomarkers, 60/40 prevalence, ×1.2 intensity): case/control AUC ≥ 0.70, biomarker recovery AUC ≥ 0.65

If Level 3 achieves those numbers, the pipeline is validated on realistic subtle-signal scenarios. If not, focus attention on the encoder pretraining and elastic-net regularization — the classification signal may be present but the interpretability layer isn't localizing it well.
