---
title: "Comparing and merging catalogs across cohorts"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 96)
library(classydms)
library(ggplot2)
library(dplyr)
library(tibble)
library(purrr)
```

# Purpose

Once you have peak catalogs from multiple cohorts (per-condition, per-batch, or split by any grouping), two natural questions:

1. **Which compounds are shared vs unique across cohorts?** Shared compounds are the "universal volatile signature" — reference points that appear in every human. Compounds only in one cohort might be biomarker candidates or artifacts of that cohort.
2. **Can we merge cohort-specific catalogs into a universal catalog** for cross-cohort encoder pretraining?

Two functions cover both:

- **`compare_catalogs(a, b)`** — pairwise comparison. Reports shared / only-A / only-B compounds and per-catalog prevalence for shared ones.
- **`merge_catalogs(list)`** — union merge. Overlapping compounds combine into one entry with per-source prevalence tracking.

This tutorial demonstrates both on the small flu subset by splitting into disease vs control catalogs.

# 1. Setup — two catalogs from the same cohort split by class

```{r setup-catalogs, cache = TRUE}
disease_dir <- if (requireNamespace("here", quietly = TRUE)) {
  here::here("sampledata", "influenza")
} else "sampledata/influenza"
control_dir <- if (requireNamespace("here", quietly = TRUE)) {
  here::here("sampledata", "influenza_controls")
} else "sampledata/influenza_controls"

loaded <- load_dms_directory(disease_dir, control_dir,
  pattern = "POS.*\\.(xls|txt|tsv|csv)$")
samples_proc <- map(loaded$samples, ~process_one_sample(.x,
  sg_p = 3, sg_n = 21, asls_lambda = 1e6,
  asls_p = 0.01, asls_niter = 10))
samples_dust <- map(samples_proc,
                     ~baseline_basement(.x, basement_thr = 0.005))
b_list <- map(samples_dust, ~trim_bounds_from_occupancy(.x$Z))
R0 <- max(sapply(samples_dust, function(s) nrow(s$Z)))
C0 <- max(sapply(samples_dust, function(s) ncol(s$Z)))
gb <- pooled_trim_bounds(b_list, R = R0, C = C0, q_lo = 0.05, q_hi = 0.95)
samples_trim <- map(samples_dust, ~trim_sample(.x, bounds = gb))
Z_raw <- map(samples_trim, ~.x$Z)
Z_pretrim <- map(samples_trim,
                  ~if (!is.null(.x$Z_pretrim)) .x$Z_pretrim else .x$Z)

pp <- estimate_peak_params(Z_raw, Z_pretrim_list = Z_pretrim,
                            top_k = 500, verbose = FALSE)

# Split by class using subset_peak_params (no need to re-run estimate).
y <- loaded$y  # 2 = disease, 1 = control
pp_disease <- subset_peak_params(pp, which(y == 2L))
pp_control <- subset_peak_params(pp, which(y == 1L))

cat_disease <- build_peak_catalog(pp_disease,
                                    eps_rt = 0.003, eps_cv = 0.015)
cat_control <- build_peak_catalog(pp_control,
                                    eps_rt = 0.003, eps_cv = 0.015)

sprintf("Disease catalog: %d compounds", nrow(cat_disease$compounds))
sprintf("Control catalog: %d compounds", nrow(cat_control$compounds))
```

# 2. Compare the two catalogs

```{r compare}
comp <- compare_catalogs(cat_disease, cat_control,
  name_a = "disease", name_b = "control")
comp$summary
```

The counts show:

- **`n_shared`** — compounds present in both catalogs at similar (RT, CV). These are the shared human volatile signature.
- **`n_only_a`** — compounds only in the disease catalog. Could be disease-specific biomarkers, or compounds that happened to only get detected in the disease subgroup at this small scale.
- **`n_only_b`** — same for controls.
- **`frac_shared_a/b`** — fraction of each catalog that's shared. If both are high (say > 0.8), the two cohorts have very similar compound structures.

## Prevalence-delta for shared compounds

For each shared compound, `prevalence_delta = prevalence_control - prevalence_disease`. Positive delta = compound more prevalent in controls; negative = more prevalent in disease.

```{r prevalence-delta, fig.height = 3.5}
ggplot(comp$shared, aes(x = prevalence_delta)) +
  geom_histogram(bins = 30, fill = "#B2182B") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "prevalence delta (control − disease)",
       y = "n shared compounds",
       title = "Cross-cohort prevalence delta among shared compounds") +
  theme_minimal()

sprintf("Compounds with |delta| > 0.20: %d",
        sum(abs(comp$shared$prevalence_delta) > 0.20))
```

Compounds with large `|prevalence_delta|` are the interesting ones — same (RT, CV) location but very different fire rate between cohorts. **These are biomarker candidates** — the same compound expressed at very different rates in the two groups.

## Cohort-unique compounds

```{r unique-viz, fig.height = 5}
biom_only_disease <- head(comp$only_a[order(-comp$only_a$prevalence), ], 10)
biom_only_control <- head(comp$only_b[order(-comp$only_b$prevalence), ], 10)

cat("Top 10 disease-only compounds by prevalence:\n")
print(biom_only_disease)
cat("\nTop 10 control-only compounds by prevalence:\n")
print(biom_only_control)
```

On a real cohort you'd want to check these against your peak-detection parameters — very-high-prevalence "cohort-only" compounds might reflect a real biological difference OR a systematic difference in how peaks were detected between subcohorts. Cross-reference against the other cohort's raw data at those coordinates.

## Spatial overlay of shared vs unique

```{r spatial-overlay, fig.height = 6}
ggplot() +
  geom_point(data = comp$shared,
              aes(x = cv_loc_disease, y = rt_loc_disease),
              color = "#666666", size = 1, alpha = 0.4) +
  geom_point(data = comp$only_a,
              aes(x = cv_loc, y = rt_loc),
              color = "#B2182B", size = 2.5) +
  geom_point(data = comp$only_b,
              aes(x = cv_loc, y = rt_loc),
              color = "#2166AC", size = 2.5) +
  scale_y_reverse() +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = sprintf(
         "Shared (%d, grey), only-disease (%d, red), only-control (%d, blue)",
         nrow(comp$shared), nrow(comp$only_a), nrow(comp$only_b))) +
  theme_minimal()
```

# 3. Merge into a universal catalog

```{r merge}
universal <- merge_catalogs(
  list(cat_disease, cat_control),
  names = c("disease", "control")
)
sprintf("Universal catalog: %d compounds (from %d + %d source compounds)",
        nrow(universal$compounds),
        nrow(cat_disease$compounds), nrow(cat_control$compounds))
```

Overlapping compounds from both source catalogs get combined into single entries. Compounds unique to either source pass through.

## Per-source prevalence tracking

Each universal-catalog compound carries a `per_source_prevalence` list-column with per-source contributions:

```{r per-source}
universal$compounds |>
  mutate(disease_prev = sapply(per_source_prevalence, `[`, 1),
         control_prev = sapply(per_source_prevalence, `[`, 2)) |>
  select(compound_id, rt_loc, cv_loc, prevalence,
         disease_prev, control_prev) |>
  head(10) |>
  knitr::kable(digits = 3)
```

You can filter universal-catalog compounds by their source composition:

```{r filter-universal}
per_source_df <- do.call(rbind, universal$compounds$per_source_prevalence) |>
  as.data.frame()
colnames(per_source_df) <- universal$parameters$source_names
per_source_df$compound_id <- universal$compounds$compound_id
sprintf("Compounds present in both sources: %d",
        sum(per_source_df$disease > 0 & per_source_df$control > 0))
sprintf("Compounds only in disease source: %d",
        sum(per_source_df$disease > 0 & per_source_df$control == 0))
sprintf("Compounds only in control source: %d",
        sum(per_source_df$disease == 0 & per_source_df$control > 0))
```

# 4. Practical uses

**Cross-condition biomarker discovery**: `compare_catalogs(cat_flu, cat_aki)`. Compounds present in one but not the other are disease-specific candidates. Compounds shared but with different prevalence are also candidates via delta.

**Universal encoder pretraining**: `merge_catalogs(list(cat_flu, cat_aki, cat_chf))` gives one catalog capturing all compounds observed across your diseases. Pretrain the encoder on this universal catalog and it will have learned "peakiness" from a broader distribution — potentially more generalizable across new diseases.

**Diagnostic**: high `frac_shared_a` and `frac_shared_b` together with small `|prevalence_delta|` on shared compounds tells you the two cohorts are almost identical at the volatile level — biology is very similar. Low shared fractions or big deltas suggest real biological difference.

# What comes next

- **Multi-cohort catalogs**: build one per disease, then `merge_catalogs(list(cat_a, cat_b, cat_c, ...))`.
- **Biomarker table**: rank shared compounds by `|prevalence_delta|` for a biomarker candidate list.
- **Test the transferability**: pretrain encoder on the universal catalog, classify each specific cohort. Does one encoder outperform per-cohort encoders?
