---
title: "Catalog pipeline sanity check — end-to-end with shared backgrounds"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 3.5, dpi = 96)
library(classydms)
library(ggplot2)
library(dplyr)
library(tibble)
library(pROC)

plot_mat <- function(mat, title = "", low = "white",
                     high = "#B2182B", limits = NULL) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient(low = low, high = high, limits = limits,
                        name = "intensity", oob = scales::squish) +
    scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank())
}
plot_div <- function(mat, title = "", limits = NULL) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  if (is.null(limits)) limits <- c(-1, 1) * max(abs(mat))
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                          midpoint = 0, limits = limits,
                          oob = scales::squish, name = "saliency") +
    scale_y_reverse() + coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank())
}
```

# Summary

This walkthrough runs the whole catalog-based pipeline (synthesize -> pretrain -> classify -> interpret) at toy scale against sanity-max acceptance targets. Sections:

1. **Purpose** — States the sanity-max acceptance targets for the catalog pipeline.
2. **Toy setup** — Hand-builds a `peak_params` with fixed compound centers plus sample-specific peaks, then runs `build_peak_catalog`.
3. **Sanity-max biomarker spec** — Picks 5 catalog compounds via `pick_catalog_biomarkers` and builds a maximum-signal biomarker spec.
4. **Simulate the cohort** — Runs `simulate_case_control_from_catalog` for 40+40 samples and checks biomarker firing rates.
5. **Case vs control visualization** — Plots one case, one control, and their diff with biomarker circles overlaid.
6. **Pretrain encoder on the catalog** — Runs `pretrain_denoising_online` with the catalog as input.
7. **Encoder-channel sanity check** — Per-channel Wilcoxon with BH adjustment plots channel significance.
8. **Classify** — Runs `cv_encoder_elastic_net` and reports 5-fold CV AUC and non-zero coefficient count.
9. **Saliency + biomarker recovery** — Aggregates saliency over cases and evaluates per-pixel recovery AUC.
10. **Summary vs acceptance targets** — Tabulates observed metrics against the sanity-max targets.
11. **What comes next** — Suggests harder scenarios, cross-cohort transfer, and universal catalog merges.

# Purpose

The [peak_params pipeline_sanity_check](pipeline_sanity_check.html) validates the pipeline end-to-end using per-sample-random peak positions. This tutorial does the same for the **catalog** flow — where every synthetic sample draws from a shared compound catalog. Catalog-based synthetic data structurally matches real GC-DMS, so the encoder learns the right inductive bias during pretraining.

Acceptance targets on sanity-max catalog scenario:

- Case/control AUC ≥ 0.95
- Biomarker recovery per-pixel AUC ≥ 0.85
- Encoder channels significant at BH q < 0.05: ≥ 30 / 64

If all three pass at toy scale, the catalog machinery works end-to-end.

# 1. Toy setup

```{r toy-setup}
# Build a toy peak_params that simulates real compound structure:
# ~20 "compounds" at fixed positions, each appearing in most samples
# with small position jitter. Plus a smaller number of sample-specific
# peaks to create low-prevalence catalog entries.
set.seed(1)
n_samples <- 20L
n_compounds <- 20L
comp_centers <- tibble(
  rt = runif(n_compounds, 0.15, 0.85),
  cv = runif(n_compounds, 0.20, 0.80),
  base_sigma_rt = exp(rnorm(n_compounds, log(4), 0.3)),
  base_sigma_cv = exp(rnorm(n_compounds, log(1.5), 0.2)),
  base_intensity = exp(rnorm(n_compounds, log(0.8), 0.5)),
  prevalence = runif(n_compounds, 0.4, 1.0)   # each compound has intrinsic prevalence
)
# Emit per-sample peaks: for each compound, roll Bernoulli(prevalence),
# place a peak near center with small jitter and morphology variation
rt_loc_raw <- c(); cv_loc_raw <- c()
sigma_rt_raw <- c(); sigma_cv_raw <- c(); intensity_raw <- c()
sample_idx_raw <- c()
for (i in seq_len(n_samples)) {
  for (k in seq_len(n_compounds)) {
    if (runif(1) >= comp_centers$prevalence[k]) next
    rt_loc_raw <- c(rt_loc_raw,
                    comp_centers$rt[k] + rnorm(1, 0, 0.005))
    cv_loc_raw <- c(cv_loc_raw,
                    comp_centers$cv[k] + rnorm(1, 0, 0.005))
    sigma_rt_raw <- c(sigma_rt_raw,
                      comp_centers$base_sigma_rt[k] * exp(rnorm(1, 0, 0.1)))
    sigma_cv_raw <- c(sigma_cv_raw,
                      comp_centers$base_sigma_cv[k] * exp(rnorm(1, 0, 0.1)))
    intensity_raw <- c(intensity_raw,
                      comp_centers$base_intensity[k] * exp(rnorm(1, 0, 0.15)))
    sample_idx_raw <- c(sample_idx_raw, i)
  }
  # Add a few sample-specific peaks
  n_extra <- rpois(1, 3)
  for (e in seq_len(n_extra)) {
    rt_loc_raw <- c(rt_loc_raw, runif(1, 0.1, 0.9))
    cv_loc_raw <- c(cv_loc_raw, runif(1, 0.15, 0.85))
    sigma_rt_raw <- c(sigma_rt_raw, exp(rnorm(1, log(4), 0.3)))
    sigma_cv_raw <- c(sigma_cv_raw, exp(rnorm(1, log(1.5), 0.2)))
    intensity_raw <- c(intensity_raw, exp(rnorm(1, log(0.5), 0.5)))
    sample_idx_raw <- c(sample_idx_raw, i)
  }
}
n_peaks_per_sample <- as.integer(table(factor(sample_idx_raw,
                                                levels = seq_len(n_samples))))
toy_params <- list(
  n_peaks_detected = length(rt_loc_raw), n_samples = n_samples,
  n_peaks = list(mean = mean(n_peaks_per_sample),
                 sd   = sd(n_peaks_per_sample),
                 values = n_peaks_per_sample),
  rt_loc = list(mean = mean(rt_loc_raw), sd = sd(rt_loc_raw)),
  cv_loc = list(mean = mean(cv_loc_raw), sd = sd(cv_loc_raw)),
  sigma_rt = list(meanlog = mean(log(sigma_rt_raw)),
                   sdlog = sd(log(sigma_rt_raw))),
  sigma_cv = list(meanlog = mean(log(sigma_cv_raw)),
                   sdlog = sd(log(sigma_cv_raw))),
  intensity = list(meanlog = mean(log(intensity_raw)),
                    sdlog = sd(log(intensity_raw))),
  noise = list(mean = 0.005, sd = 0.005),
  rt_loc_raw     = rt_loc_raw,
  cv_loc_raw     = cv_loc_raw,
  sigma_rt_raw   = sigma_rt_raw,
  sigma_cv_raw   = sigma_cv_raw,
  intensity_raw  = intensity_raw,
  sample_idx_raw = sample_idx_raw
)

catalog <- build_peak_catalog(toy_params,
                               eps_rt = 0.015, eps_cv = 0.04,
                               min_support_frac = 0.10)
H <- 150L; W <- 30L
sprintf("Catalog: %d compounds (from %d pool peaks, %d toy samples)",
        nrow(catalog$compounds), length(rt_loc_raw), toy_params$n_samples)
```

# 2. Sanity-max biomarker spec

```{r biom-spec}
# On this small toy catalog the top-5% intensity filter can leave too
# few candidates; relax to top 30% for the tutorial. Real-scale
# catalogs (thousands of compounds) work fine with the preset default.
biom_ids <- pick_catalog_biomarkers(catalog,
                                     n_biomarkers = 5L,
                                     min_prevalence = 0.3,
                                     min_intensity_quantile = 0.70,
                                     seed = 42L)
biom_spec <- build_biomarker_spec_catalog(biom_ids,
  case_prevalence = 1.0, control_prevalence = 0.0,
  case_intensity_mult = 10.0)
biom_spec
```

# 3. Simulate the cohort

```{r simulate, cache = TRUE}
sim <- simulate_case_control_from_catalog(
  catalog, biomarkers = biom_spec,
  n_cases = 40L, n_controls = 40L,
  H = H, W = W, seed = 7L
)
Z_list <- lapply(sim$samples, `[[`, "Z")
y <- sim$y
table(y, dnn = "class")
```

Sanity of the simulation: biomarker firing rates by class

```{r firing-check}
firing <- as.data.frame(sim$per_sample_firing)
colnames(firing) <- paste0("bio_", biom_spec$compound_id)
firing$class <- factor(y, levels = c(1, 2), labels = c("control", "case"))
firing |>
  tidyr::pivot_longer(-class, names_to = "biomarker", values_to = "fired") |>
  group_by(class, biomarker) |>
  summarise(prevalence = mean(fired), .groups = "drop")
```

Cases should fire biomarkers at 100%, controls at 0%.

# 4. Case vs control visualization

```{r viz, fig.height = 4}
case_i    <- which(y == 2)[1]; ctrl_i    <- which(y == 1)[1]
Z_case    <- Z_list[[case_i]]; Z_control <- Z_list[[ctrl_i]]
Z_diff    <- Z_case - Z_control

biom_locs <- catalog$compounds[
  match(biom_spec$compound_id, catalog$compounds$compound_id),
  c("rt_loc", "cv_loc")]
b_r <- biom_locs$rt_loc * H; b_c <- biom_locs$cv_loc * W

mark <- function(p) p + annotate("point", x = b_c, y = b_r,
  shape = 21, color = "cyan", fill = NA, size = 6, stroke = 1.2)

lim <- c(0, max(c(Z_case, Z_control)))
gridExtra::grid.arrange(
  mark(plot_mat(Z_case,    "case",    limits = lim)),
  mark(plot_mat(Z_control, "control", limits = lim)),
  mark(plot_div(Z_diff, "case - control")),
  ncol = 3
)
```

Non-biomarker positions should look nearly identical between case and control (that's the shared background). The circles (biomarker positions) should be bright in case, absent in control.

# 5. Pretrain encoder on the catalog

```{r pretrain, cache = TRUE}
pt <- pretrain_denoising_online(
  peak_params = NULL, catalog = catalog,
  H = H, W = W,
  steps_per_epoch = 60L, epochs = 8L, batch_size = 16L,
  size_jitter = 0.15, noise_scale = 3.0, val_n = 32L,
  device = "cpu", seed = 1L, verbose = FALSE
)
encoder <- pt$encoder
```

# 6. Encoder-channel sanity check

```{r encoder-check, fig.height = 3.5}
feats <- extract_encoder_features(encoder, Z_list, device = "cpu")
p_raw <- sapply(seq_len(64L), function(c)
  suppressWarnings(wilcox.test(feats[y == 2L, c],
                                feats[y == 1L, c])$p.value))
p_bh <- p.adjust(p_raw, method = "BH")
n_sig_05  <- sum(p_bh < 0.05, na.rm = TRUE)
n_sig_001 <- sum(p_bh < 0.001, na.rm = TRUE)
sprintf("Encoder channels significant at BH q < 0.05:  %d / 64", n_sig_05)
sprintf("Encoder channels significant at BH q < 0.001: %d / 64", n_sig_001)

ggplot(tibble(channel = seq_len(64L), p_bh = p_bh),
       aes(x = channel, y = -log10(p_bh))) +
  geom_col(fill = "#B2182B") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
              color = "grey40") +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed",
              color = "grey60") +
  labs(x = "encoder channel", y = "-log10(BH q)",
       title = "Per-channel Wilcoxon by class") +
  theme_minimal()
```

# 7. Classify

```{r classify}
elnet <- cv_encoder_elastic_net(encoder, Z_list, y,
                                  k = 5L, alpha = 0.5, seed = 1L)
sprintf("Case/control AUC (5-fold CV): %.3f", elnet$auc)
sprintf("Non-zero elastic-net coefs: %d / 64", sum(elnet$coefs != 0))
```

# 8. Saliency + biomarker recovery

```{r saliency, fig.height = 4}
case_indices <- which(y == 2L)
orig_dims <- lapply(sim$samples,
                     function(s) c(H = nrow(s$Z), W = ncol(s$Z)))
sal <- aggregate_saliency_masked(encoder, Z_list, coefs = elnet$coefs,
                                   sample_indices = case_indices,
                                   orig_dims = orig_dims)

biom_locs <- catalog$compounds[
  match(biom_spec$compound_id, catalog$compounds$compound_id),
  c("rt_loc", "cv_loc")]
recovery <- evaluate_biomarker_recovery(sal, biom_locs,
                                         H = H, W = W,
                                         match_radius_rt = 8L,
                                         match_radius_cv = 2L,
                                         top_k = nrow(biom_spec))
sprintf("Biomarker recovery per-pixel AUC: %.3f", recovery$auc_per_pixel)
sprintf("Recall @ top-%d: %.3f", nrow(biom_spec), recovery$recall_at_k)

plot_div(sal, "aggregated saliency: cases")
```

# 9. Summary vs acceptance targets

```{r summary}
tibble(
  metric = c("Case/control AUC",
             "Biomarker recovery AUC",
             "Encoder channels (BH q < 0.05)"),
  observed = c(sprintf("%.3f", elnet$auc),
               sprintf("%.3f", recovery$auc_per_pixel),
               sprintf("%d / 64", n_sig_05)),
  target = c(">= 0.95", ">= 0.85", ">= 30 / 64"),
  pass = c(elnet$auc >= 0.95,
           recovery$auc_per_pixel >= 0.85,
           n_sig_05 >= 30)
) |> knitr::kable()
```

If all three pass, the catalog pipeline works end-to-end. Ready to run at real scale via `inst/scripts/run_catalog_sanity_max.R` on your actual data.

# What comes next

- Try harder scenarios: realistic-intensity (`case_intensity_mult = 1.0`) with imperfect prevalence (`case_prev = 0.80, control_prev = 0.20`) — this is the honest biomarker discovery difficulty
- Cross-cohort transfer: pretrain on one disease's catalog, classify on another. Uses `split_catalog` or per-condition catalogs.
- Universal catalog: `merge_catalogs(list(cat_aki, cat_chf, cat_flu))` to build one encoder that works across diseases.
