---
title: "Catalog-transfer experiment — does encoder pretraining generalize?"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 96)
library(classydms)
library(ggplot2)
library(dplyr)
library(tibble)
library(purrr)
library(pROC)
```

# Summary

This walkthrough tests whether an encoder pretrained on one compound subset transfers to classification on a disjoint subset, by splitting one catalog and comparing regimes. Sections:

1. **Purpose** — Frames the encoder-transfer question and outlines the split-catalog design.
2. **Load flu data and build catalog** — Preprocesses flu samples and builds one catalog.
3. **Split the catalog** — Uses `split_catalog` to create disjoint compound halves A and B.
4. **Simulate a case/control cohort from split B** — Picks biomarkers from split B and simulates 20+20 samples.
5. **Two pretraining regimes to compare** — Pretrains encoders on split A (transfer), split B (matched), and split A with wider `size_jitter`.
6. **Classify and compare AUC** — Runs `cv_encoder_elastic_net` per encoder and tabulates AUCs and non-zero coefficient counts.
7. **Sanity check: encoder channel differentiation** — Per-channel Wilcoxon by class for each regime.
8. **What to do with the results** — Interprets the AUC gaps to answer the one-encoder-vs-per-disease question and the jitter question.
9. **What to run next** — Lists cross-cohort transfer, wider jitter sweeps, merged-catalog pretraining, and scale-up.

# Purpose

The key transfer-learning question: **can one encoder pretrained on "generic peakiness" from one distribution be applied to case/control classification on a *different* distribution?**

If yes → we can train one big encoder from many diseases combined (or synthetic-from-scratch) and reuse it. If no → we need per-disease pretraining, which changes the deployment story significantly.

Design:

1. Build a peak catalog from a real cohort (this tutorial: 6-sample flu subset).
2. Split the catalog into two disjoint halves — split A (for pretraining) and split B (for classification).
3. Pretrain an encoder on synthetic data from split A. **The encoder never sees split B's compounds.**
4. Simulate a case/control cohort from split B, injecting biomarkers into split B's compounds.
5. Classify with elastic net; report AUC.
6. Compare against a "matched" baseline: pretrain and classify on the same split B — reveals the transfer penalty.

Also comparing `size_jitter` values to test whether wider training-time morphology jitter improves transfer.

# 1. Load flu data and build catalog

```{r load, cache = TRUE}
disease_dir <- if (requireNamespace("here", quietly = TRUE)) {
  here::here("sampledata", "influenza")
} else "sampledata/influenza"
control_dir <- if (requireNamespace("here", quietly = TRUE)) {
  here::here("sampledata", "influenza_controls")
} else "sampledata/influenza_controls"

loaded <- load_dms_directory(disease_dir, control_dir,
  pattern = "POS.*\\.(xls|txt|tsv|csv)$")
samples_proc <- map(loaded$samples, ~process_one_sample(.x,
  sg_p = 3, sg_n = 21, asls_lambda = 1e6, asls_p = 0.01, asls_niter = 10))
samples_dust <- map(samples_proc, ~baseline_basement(.x, basement_thr = 0.005))
b_list <- map(samples_dust, ~trim_bounds_from_occupancy(.x$Z))
R0 <- max(sapply(samples_dust, function(s) nrow(s$Z)))
C0 <- max(sapply(samples_dust, function(s) ncol(s$Z)))
gb <- pooled_trim_bounds(b_list, R = R0, C = C0, q_lo = 0.05, q_hi = 0.95)
samples_trim <- map(samples_dust, ~trim_sample(.x, bounds = gb))
Z_raw <- map(samples_trim, ~.x$Z)
Z_pretrim <- map(samples_trim,
                  ~if (!is.null(.x$Z_pretrim)) .x$Z_pretrim else .x$Z)
pp <- estimate_peak_params(Z_raw, Z_pretrim_list = Z_pretrim,
                            top_k = 500, verbose = FALSE)

catalog <- build_peak_catalog(pp, eps_rt = 0.003, eps_cv = 0.015)
sprintf("Full catalog: %d compounds", nrow(catalog$compounds))
```

At the smoke-test scale the compound count is small; on your real AKI-scale cohort you'd have ~2000-3000 compounds and the transfer test is more meaningful.

# 2. Split the catalog

```{r split}
splits <- split_catalog(catalog, frac = 0.5, seed = 1L)
sprintf("Split A (pretraining): %d compounds", nrow(splits$a$compounds))
sprintf("Split B (classification): %d compounds", nrow(splits$b$compounds))
```

The two splits are guaranteed to have **disjoint compound identities**. Any compound in A cannot appear in B and vice versa.

# 3. Simulate a case/control cohort from split B

Biomarkers are drawn from split B (independent of split A's pretraining data).

```{r simulate, cache = TRUE}
H <- nrow(Z_raw[[1]]); W <- ncol(Z_raw[[1]])
biom_spec <- catalog_biomarkers_sanity_max(splits$b,
  min_base_prevalence = 0.3, seed = 42L)
sim <- simulate_case_control_from_catalog(splits$b,
  biomarkers = biom_spec,
  n_cases = 20L, n_controls = 20L,
  H = H, W = W, seed = 7L)
sprintf("Cohort: %d samples", length(sim$samples))
```

# 4. Two pretraining regimes to compare

## 4a. Transfer regime — pretrain on split A

The encoder sees only compounds from split A during pretraining. It never sees any biomarker or background compound from split B until inference time.

```{r pretrain-transfer, cache = TRUE}
pt_transfer <- pretrain_denoising_online(
  peak_params = NULL,
  catalog = splits$a,
  H = H, W = W,
  steps_per_epoch = 40L, epochs = 5L, batch_size = 8L,
  size_jitter = 0.15,
  noise_scale = 3.0, val_n = 16L,
  device = "cpu", seed = 1L,
  verbose = FALSE
)
```

## 4b. Matched regime — pretrain on split B (upper bound)

Same encoder architecture, same hyperparameters, but pretraining sees split B's compounds. This is the "encoder was trained on the target distribution" upper bound.

```{r pretrain-matched, cache = TRUE}
pt_matched <- pretrain_denoising_online(
  peak_params = NULL,
  catalog = splits$b,
  H = H, W = W,
  steps_per_epoch = 40L, epochs = 5L, batch_size = 8L,
  size_jitter = 0.15,
  noise_scale = 3.0, val_n = 16L,
  device = "cpu", seed = 1L,
  verbose = FALSE
)
```

## 4c. Wider-jitter transfer regime — larger `size_jitter`

Test whether forcing the encoder to see more morphological variation during pretraining improves transfer.

```{r pretrain-transfer-wide, cache = TRUE}
pt_transfer_wide <- pretrain_denoising_online(
  peak_params = NULL,
  catalog = splits$a,
  H = H, W = W,
  steps_per_epoch = 40L, epochs = 5L, batch_size = 8L,
  size_jitter = 0.5,   # wider
  noise_scale = 3.0, val_n = 16L,
  device = "cpu", seed = 1L,
  verbose = FALSE
)
```

# 5. Classify and compare AUC

```{r classify}
Z_list <- lapply(sim$samples, `[[`, "Z")

fit_and_report <- function(encoder, label) {
  set.seed(1L)
  elnet <- cv_encoder_elastic_net(encoder, Z_list, sim$y,
                                    k = 5L, alpha = 0.5, seed = 1L)
  tibble(regime = label, auc = elnet$auc,
         n_nonzero_coefs = sum(elnet$coefs != 0))
}
results <- bind_rows(
  fit_and_report(pt_transfer$encoder,       "transfer (split A, jitter=0.15)"),
  fit_and_report(pt_matched$encoder,        "matched  (split B, jitter=0.15)"),
  fit_and_report(pt_transfer_wide$encoder,  "transfer (split A, jitter=0.5)")
)
results |> knitr::kable(digits = 3)
```

**How to read this:**

- **`matched` AUC** is the upper bound — encoder was trained on the target distribution's compound positions.
- **`transfer (jitter=0.15)` AUC**: how much do we lose by pretraining on an independent compound subset?
- **`transfer (jitter=0.5)` AUC**: does wider morphology jitter improve transfer?

Interpretation guide:

- **Transfer ≈ matched**: the encoder learned "generic peakiness" that generalizes across compound sets. One big encoder across diseases is viable.
- **Transfer << matched**: encoder memorized specific compound positions during pretraining. Per-disease encoders needed.
- **Wider jitter > tight jitter transfer**: wider training-time jitter helps generalization. Use it in production.
- **Wider jitter < tight jitter transfer**: too much jitter degrades useful signal. Stick with tight.

# 6. Sanity check: encoder channel differentiation

For each regime, is the encoder actually representing class-relevant content?

```{r encoder-channels}
per_channel_wilcox <- function(encoder, Z_list, y) {
  feats <- extract_encoder_features(encoder, Z_list)
  p_bh <- p.adjust(sapply(seq_len(64L), function(c)
    suppressWarnings(wilcox.test(feats[y == 2L, c],
                                  feats[y == 1L, c])$p.value)),
    method = "BH")
  tibble(n_sig_05 = sum(p_bh < 0.05),
         n_sig_001 = sum(p_bh < 0.001))
}
enc_stats <- bind_rows(
  cbind(regime = "transfer",       per_channel_wilcox(pt_transfer$encoder,       Z_list, sim$y)),
  cbind(regime = "matched",        per_channel_wilcox(pt_matched$encoder,        Z_list, sim$y)),
  cbind(regime = "transfer wide",  per_channel_wilcox(pt_transfer_wide$encoder,  Z_list, sim$y))
)
enc_stats |> knitr::kable()
```

# 7. What to do with the results

Answers the two questions from the beginning:

**Q1: Should we train one big encoder or per-disease?**

- If `transfer` AUC is within ~0.05 of `matched` AUC → one big encoder works. Pretrain from ANY GC-DMS distribution (or a merged catalog across diseases) and apply everywhere.
- If `transfer` AUC is well below (>0.10 gap) → the encoder is over-fitting compound positions. Either need per-disease encoders, or need to modify pretraining to be more generic (wider jitter, more diverse augmentation).

**Q2: Should synthetic data be varied or close to empirical?**

- If `transfer (wide jitter)` > `transfer (tight jitter)` → more variation during pretraining helps. Consider `size_jitter = 0.3–0.5` for production.
- If `transfer (tight jitter)` > `transfer (wide jitter)` → tight-to-empirical is better. Keep `size_jitter = 0.15`.

# 8. What to run next

- **Cross-cohort transfer**: use a catalog from a *different disease* entirely (e.g. flu-catalog → AKI-classification) for stronger transfer test.
- **Sweep more `size_jitter` values**: 0.1, 0.2, 0.3, 0.5, 0.8 to find the sweet spot.
- **Pretrain on merged catalog**: `merge_catalogs(list(cat_aki, cat_chf, cat_flu))` — a universal encoder from multiple diseases.
- **Retrain the winning configuration at larger scale** — the tutorial uses tiny pretraining (5 epochs × 40 steps × batch 8 = 1600 samples seen). Production would be ~30 epochs × 500 steps × batch 32 = ~500,000 samples.
