---
title: "Pipeline sanity check — does the full stack work at all?"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 3.5, dpi = 96)
library(classydms)
library(ggplot2)
library(dplyr)
library(tibble)
library(pROC)

plot_mat <- function(mat, title = "", low = "white", high = "#B2182B",
                     limits = NULL) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient(low = low, high = high, limits = limits,
                        name = "intensity", oob = scales::squish) +
    scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank())
}
plot_diverging <- function(mat, title = "", limits = NULL) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  if (is.null(limits)) limits <- c(-1, 1) * max(abs(mat))
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                          midpoint = 0, limits = limits, name = "saliency",
                          oob = scales::squish) +
    scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank())
}
```

# TL;DR

This walkthrough runs the full pipeline (synthesize -> pretrain -> classify -> interpret) under maximally favorable conditions to check that everything works end-to-end. Sections:

1. **Purpose** — Describes the whole-pipeline check under sanity-max conditions and the pipeline-health questions it answers.
2. **The sanity-max scenario** — Explains the `biomarkers_sanity_max()` levers and builds a toy 200-peak pool.
3. **Simulate the cohort** — Simulates 60+60 samples and visualizes one case, one control, and their diff.
4. **Pretrain the encoder** — Runs `pretrain_denoising_online` on the sparse background peak_params.
5. **Encoder-channel sanity check** — Per-channel Wilcoxon with BH adjustment to confirm the encoder sees class content.
6. **Classify — case/control AUC** — Runs `cv_encoder_elastic_net` and states acceptance bands.
7. **Saliency + biomarker recovery** — Aggregates case and control saliency and evaluates per-pixel recovery.
8. **Cohort size sweep — does bigger help?** — Fixes the encoder and re-runs classification on random subsamples of increasing size.
9. **Summary — what to conclude** — Blank table of the four acceptance targets to fill in with observed results.
10. **Realistic-intensity scenario** — Rebuilds the scenario without the intensity multiplier or top-quantile bias, retrains encoder, re-classifies, and re-runs saliency plus a comparison table.
11. **Future experiments to plug in here** — Lists broader-pretraining, encoder-cohort sweep, architecture, and intermediate-difficulty ideas.

# Purpose

Every previous tutorial builds intuition for *one* function. This one runs the **entire pipeline** — synthesize → pretrain → classify → interpret — under maximally-favorable conditions and asks:

- **Does the classifier detect an injected signal we know is there?**
- **Do the encoder's features actually vary with class?**
- **Does the saliency map put its hotspot at the injected biomarker?**
- **How much do these results depend on cohort size?**

If Level-1 case/control AUC on your real cohort is coming back near chance (0.50–0.60), something is likely broken. The right diagnostic move is to run this scenario, which is designed so that if AUC isn't ≥ 0.95 there's a pipeline bug, not just a hard signal.

The tutorial uses small `150 × 30` images so a full pretrain + classify + saliency run finishes in about a minute on CPU.

# 1. The sanity-max scenario

`biomarkers_sanity_max()` combines every "make the signal easy" lever we have:

| Lever | Value | Rationale |
|---|---|---|
| # biomarkers | 5 | Collective signal, not one lone peak |
| Intensity quantile pool | top 5% | Bright observed peaks |
| Case prevalence | 1.00 | Present in every case |
| Control prevalence | 0.00 | Never in controls |
| Case intensity multiplier | 10× | Very bright injection |
| Background `n_peaks$mean` | 30 | Sparse — biomarkers dominate |

Compared to Level 1 (~1300 peaks background, 1 biomarker, 5× mult), sanity-max shrinks background by ~40× and grows signal by ~50×. Signal fraction of total image intensity jumps from ~3% to well over 50%.

```{r toy-params}
# A procedurally-built toy pool of ~200 peaks spread across the image
# so top-5% quantile filtering has enough candidates. Real cohorts have
# many thousands of pool peaks.
set.seed(1)
N_POOL <- 200L
peaks_obs <- tibble(
  rt_loc    = runif(N_POOL, 0.10, 0.90),
  cv_loc    = runif(N_POOL, 0.15, 0.85),
  sigma_rt  = exp(rnorm(N_POOL, log(4),   0.4)),
  sigma_cv  = exp(rnorm(N_POOL, log(1.5), 0.3)),
  intensity = exp(rnorm(N_POOL, log(0.5), 0.7))
)
toy_params <- list(
  n_peaks_detected = nrow(peaks_obs),
  n_peaks   = list(mean = 20, sd = 5),
  rt_loc    = list(mean = mean(peaks_obs$rt_loc), sd = sd(peaks_obs$rt_loc)),
  cv_loc    = list(mean = mean(peaks_obs$cv_loc), sd = sd(peaks_obs$cv_loc)),
  sigma_rt  = list(meanlog = mean(log(peaks_obs$sigma_rt)),
                   sdlog   = sd(log(peaks_obs$sigma_rt))),
  sigma_cv  = list(meanlog = mean(log(peaks_obs$sigma_cv)),
                   sdlog   = sd(log(peaks_obs$sigma_cv))),
  intensity = list(meanlog = mean(log(peaks_obs$intensity)),
                   sdlog   = sd(log(peaks_obs$intensity))),
  noise     = list(mean = 0.005, sd = 0.005),
  rt_loc_raw    = peaks_obs$rt_loc,
  cv_loc_raw    = peaks_obs$cv_loc,
  sigma_rt_raw  = peaks_obs$sigma_rt,
  sigma_cv_raw  = peaks_obs$sigma_cv,
  intensity_raw = peaks_obs$intensity
)
H <- 150L; W <- 30L
cat(sprintf("Pool: %d peaks. Intensity quantiles (25/50/95/100%%):\n",
            N_POOL))
print(round(quantile(peaks_obs$intensity, c(0.25, 0.50, 0.95, 1.0)), 3))
```

```{r scenario}
scenario <- biomarkers_sanity_max(toy_params, H = H, W = W, seed = 42L)
cat("Biomarker templates:\n")
print(scenario$biomarkers)
cat("\nBackground n_peaks setting (should be 30):",
    scenario$background_peak_params$n_peaks$mean, "\n")
```

# 2. Simulate the cohort

Use a moderate-size cohort (60 case + 60 control) — enough for stable 5-fold CV.

```{r simulate, cache = TRUE}
sim <- simulate_case_control_cohort(
  scenario$background_peak_params, scenario$biomarkers,
  n_cases = 60L, n_controls = 60L,
  H = H, W = W, seed = 7L
)
Z_list <- lapply(sim$samples, `[[`, "Z")
y <- sim$y
table(y, dnn = "class (2 = case, 1 = control)")

# Sanity: every case should have every biomarker fired
sprintf("Biomarker firing per class (should be 1.00 for cases, 0.00 for controls):")
firing <- as.data.frame(sim$per_sample_firing)
firing$class <- factor(y, levels = c(1, 2), labels = c("control", "case"))
firing |>
  tidyr::pivot_longer(-class, names_to = "biomarker", values_to = "fired") |>
  group_by(class) |> summarise(prevalence = mean(fired), .groups = "drop")
```

Visualize one case vs one control on log scale + diff:

```{r sim-viz, fig.height = 4}
Z_case    <- Z_list[[which(y == 2L)[1]]]
Z_control <- Z_list[[which(y == 1L)[1]]]
Z_diff    <- Z_case - Z_control

biom <- scenario$biomarkers
b_r <- biom$rt_loc * H
b_c <- biom$cv_loc * W
mark_biomarkers <- function(p) {
  p + annotate("point", x = b_c, y = b_r, shape = 21, color = "cyan",
               fill = NA, size = 6, stroke = 1.2)
}

lim <- c(0, max(c(Z_case, Z_control)))
gridExtra::grid.arrange(
  mark_biomarkers(plot_mat(Z_case, "case sample", limits = lim)),
  mark_biomarkers(plot_mat(Z_control, "control sample", limits = lim)),
  mark_biomarkers(plot_diverging(Z_diff, "case - control")),
  ncol = 3
)
```

Every cyan circle in the case panel should have a red blob (biomarker fired), and none of them in the control panel. The diff panel should be dominated by red at biomarker locations.

# 3. Pretrain the encoder

Fast pretraining directly from the (sparse) `background_peak_params`. The encoder is class-agnostic — it just learns the general "peak on background" structure of the sparse-background distribution.

```{r pretrain, cache = TRUE}
pretrain_result <- pretrain_denoising_online(
  scenario$background_peak_params,
  H = H, W = W,
  steps_per_epoch = 50L, epochs = 8L, batch_size = 16L,
  noise_scale = 3.0, val_n = 32L,
  device = "cpu", seed = 1L
)
encoder <- pretrain_result$encoder
```

# 4. Encoder-channel sanity check

**Before touching the classifier**, ask: do any of the 64 encoder channels actually vary with class? If none do, the encoder isn't seeing the class-associated content and there's nothing for the elastic net to work with. If some do, the elastic net should be able to use them.

```{r encoder-channel-check, fig.height = 3.5}
feats <- extract_encoder_features(encoder, Z_list, device = "cpu")
dim(feats)

# Per-channel Wilcoxon by class
p_raw <- sapply(seq_len(64L), function(c)
  suppressWarnings(wilcox.test(feats[y == 2L, c], feats[y == 1L, c])$p.value))
p_bh <- p.adjust(p_raw, method = "BH")

sig_alpha_05  <- sum(p_bh < 0.05, na.rm = TRUE)
sig_alpha_001 <- sum(p_bh < 0.001, na.rm = TRUE)
sprintf("Encoder channels with BH-FDR-significant class difference:")
sprintf("  q < 0.05:  %d / 64", sig_alpha_05)
sprintf("  q < 0.001: %d / 64", sig_alpha_001)

ch_df <- tibble(channel = seq_len(64L), p_raw = p_raw, p_bh = p_bh) |>
  arrange(p_bh)
ggplot(ch_df, aes(x = seq_len(64), y = -log10(p_bh))) +
  geom_col(fill = "#B2182B") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "grey60") +
  labs(x = "encoder channel (sorted by significance)",
       y = "-log10(BH-adjusted p)",
       title = "Per-channel Wilcoxon between cases and controls") +
  theme_minimal()
```

**How to read this:**

- **Many significant channels** = encoder is representing class content well. Elastic net will have signal to work with.
- **Very few** = encoder doesn't distinguish cases from controls at the feature level. Elastic net can't recover what isn't in its input. Fix upstream (more pretraining, different `noise_scale`, more diverse pretraining distribution).
- **None** = pipeline is broken somewhere before the classifier.

# 5. Classify — case/control AUC

```{r classify}
elnet <- cv_encoder_elastic_net(
  encoder, Z_list = Z_list, y = y,
  k = 5L, alpha = 0.5, seed = 1L
)
sprintf("Case/control AUC (5-fold CV): %.3f", elnet$auc)
sprintf("Non-zero elastic-net coefficients: %d / 64", sum(elnet$coefs != 0))
```

**Acceptance targets for sanity-max:**

- AUC ≥ 0.95: pipeline is healthy, move on to progressively harder scenarios.
- AUC 0.75–0.95: pipeline works but there's something to improve. Likely: encoder pretraining under-fits (increase epochs) or noise_scale too high.
- AUC < 0.75 despite the aggressive sanity-max settings: something is broken. Check the encoder-channel Wilcoxon in step 4 — if no channels are significant, look at the encoder (dead channels? bad pretraining loss?). If channels ARE significant but elastic net misses them, look at the classifier (regularization too aggressive? fold size too small?).

# 6. Saliency + biomarker recovery

```{r saliency, fig.height = 4}
case_indices    <- which(y == 2L)
control_indices <- which(y == 1L)
orig_dims <- lapply(sim$samples, function(s) c(H = nrow(s$Z), W = ncol(s$Z)))

sal_cases    <- aggregate_saliency_masked(encoder, Z_list,
                                            coefs = elnet$coefs,
                                            sample_indices = case_indices,
                                            orig_dims = orig_dims)
sal_controls <- aggregate_saliency_masked(encoder, Z_list,
                                            coefs = elnet$coefs,
                                            sample_indices = control_indices,
                                            orig_dims = orig_dims)

lim <- c(-1, 1) * max(abs(c(sal_cases, sal_controls)))
gridExtra::grid.arrange(
  mark_biomarkers(plot_diverging(sal_cases,    "saliency: cases", limits = lim)),
  mark_biomarkers(plot_diverging(sal_controls, "saliency: controls", limits = lim)),
  ncol = 2
)

recovery <- evaluate_biomarker_recovery(
  sal_cases, scenario$biomarkers,
  H = H, W = W,
  match_radius_rt = 8L, match_radius_cv = 2L,
  top_k = 5L
)
sprintf("Per-pixel biomarker recovery AUC: %.3f",  recovery$auc_per_pixel)
sprintf("Recall @ top-5: %.3f", recovery$recall_at_k)
sprintf("Precision @ top-5: %.3f", recovery$precision_at_k)
```

**Acceptance targets:**

- Per-pixel AUC ≥ 0.85 on sanity-max
- Recall @ top-5 ≥ 0.60 (at least 3 of 5 biomarkers land in the top-5 saliency hotspots)

# 7. Cohort size sweep — does bigger help?

Fix the encoder, vary the classifier cohort. Subsample the simulated cohort to different sizes and re-run classification only. This isolates cohort-size effect on the elastic net from cohort-size effect on encoder training.

```{r size-sweep, fig.height = 3.5, cache = TRUE}
sizes <- c(10L, 20L, 40L, 60L)
n_reps <- 5L  # different random subsamples per size for a variance estimate

sweep <- lapply(sizes, function(n) {
  aucs <- sapply(seq_len(n_reps), function(rep) {
    set.seed(100L + n * 10L + rep)
    case_sub <- sample(case_indices, n)
    ctrl_sub <- sample(control_indices, n)
    idx <- c(case_sub, ctrl_sub)
    fit <- cv_encoder_elastic_net(encoder, Z_list[idx], y[idx],
                                   k = 5L, alpha = 0.5, seed = 1L)
    fit$auc
  })
  tibble(n_per_class = n,
         auc_mean = mean(aucs), auc_sd = sd(aucs),
         auc_lo = quantile(aucs, 0.25), auc_hi = quantile(aucs, 0.75))
})
sweep_df <- bind_rows(sweep)
sweep_df

ggplot(sweep_df, aes(x = n_per_class, y = auc_mean)) +
  geom_ribbon(aes(ymin = auc_lo, ymax = auc_hi), alpha = 0.2, fill = "#B2182B") +
  geom_line(color = "#B2182B", size = 1.1) +
  geom_point(color = "#B2182B", size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey60") +
  scale_x_continuous(breaks = sizes) +
  labs(x = "samples per class", y = "case/control AUC",
       title = sprintf("Effect of cohort size (%d reps per size)", n_reps),
       subtitle = "Encoder fixed; only the classifier cohort varies") +
  ylim(0.4, 1.0) +
  theme_minimal()
```

**How to read this:**

- **Monotone increase, flattening after N = 40–60** = you have enough samples per class to make the classifier stable. Adding more is diminishing returns.
- **Steep increase across all sizes with no plateau** = you need more samples. Scale up cohort.
- **Flat and low across all sizes** = the signal is too weak for the classifier to detect regardless of cohort size. The problem is upstream (encoder, pretraining, biomarker strength), not sample count.
- **High variance (wide ribbon)** = classifier is unstable in your regime; consider larger `k` in CV or averaging across seeds.

# Summary — what to conclude

Compare your results against these acceptance targets:

| Metric | Sanity-max target | Your result | Verdict |
|---|---|---|---|
| Case/control AUC | ≥ 0.95 | (fill in) | pass / borderline / fail |
| Encoder-channel Wilcoxon (BH q < 0.05) | ≥ 20 / 64 | (fill in) | pass / borderline / fail |
| Biomarker recovery per-pixel AUC | ≥ 0.85 | (fill in) | pass / borderline / fail |
| Cohort-size curve | monotone, plateau by N=60 | (fill in) | pass / borderline / fail |

If all four are green, the pipeline mechanics are sound and you can climb the difficulty ladder (Level 1 → Level 2 → Level 3, then real cohorts). If any are red, investigate that specific piece before running longer / bigger.

# 8. Realistic-intensity scenario

Sanity-max multiplied biomarker intensity by 10× so the injected signal was visually obvious. In real GC-DMS biomarker studies the differential compounds are often *at the same intensity scale* as background peaks — the signal comes from **which peaks are present**, not from any one peak being dramatically brighter.

This section keeps the sanity property (perfect presence in cases, complete absence in controls) but removes both the intensity multiplier and the top-quantile bias, so biomarkers are drawn from the same distribution as background peaks. Visually indistinguishable from background peaks in a single sample; only the systematic presence/absence across samples reveals them.

```{r realistic-scenario}
# Build a realistic-intensity scenario manually rather than through a
# level preset. Uniform quantile means biomarkers can come from anywhere
# in the pool; case_intensity_mult = 1.0 means no boost.
ho_r <- holdout_biomarkers(toy_params, n_biomarkers = 5L,
                            H = H, W = W,
                            min_intensity_quantile = NULL,
                            seed = 42L)
bg_r <- ho_r$background_peak_params
bg_r$n_peaks$mean <- 30
bg_r$n_peaks$sd   <- 5

biom_r <- build_biomarker_spec(
  ho_r$biomarker_templates,
  case_prevalence        = 1.0,
  control_prevalence     = 0.0,
  case_intensity_mult    = 1.0,    # <-- no boost
  control_intensity_mult = 1.0
)

# Compare biomarker intensity ranges between scenarios
smx_min <- min(scenario$biomarkers$intensity)
smx_max <- max(scenario$biomarkers$intensity)
smx_mlt <- scenario$biomarkers$case_intensity_mult[1]
rea_min <- min(biom_r$intensity)
rea_max <- max(biom_r$intensity)

tibble(
  scenario = c("sanity-max", "realistic"),
  biomarker_intensity_range = c(
    paste0(round(smx_min, 2), " - ", round(smx_max, 2)),
    paste0(round(rea_min, 2), " - ", round(rea_max, 2))
  ),
  case_intensity_mult = c(smx_mlt, 1.0),
  effective_case_amp_range = c(
    paste0(round(smx_min * smx_mlt, 2), " - ",
           round(smx_max * smx_mlt, 2)),
    paste0(round(rea_min, 2), " - ", round(rea_max, 2))
  )
) |> knitr::kable(align = "lrrr")
```

## Simulate the realistic cohort

```{r realistic-simulate, cache = TRUE}
sim_r <- simulate_case_control_cohort(
  bg_r, biom_r,
  n_cases = 60L, n_controls = 60L,
  H = H, W = W, seed = 11L
)
Z_r <- lapply(sim_r$samples, `[[`, "Z")
y_r <- sim_r$y
b_r_pos <- biom_r$rt_loc * H
b_c_pos <- biom_r$cv_loc * W
```

Visualize a case and a control on their **own** intensity scale (so the reader can see that biomarker peaks look just like background peaks):

```{r realistic-viz, fig.height = 4}
Z_case_r    <- Z_r[[which(y_r == 2L)[1]]]
Z_ctrl_r    <- Z_r[[which(y_r == 1L)[1]]]
Z_diff_r    <- Z_case_r - Z_ctrl_r
mark_r <- function(p) {
  p + annotate("point", x = b_c_pos, y = b_r_pos, shape = 21,
               color = "cyan", fill = NA, size = 6, stroke = 1.2)
}
lim_r <- c(0, max(c(Z_case_r, Z_ctrl_r)))
gridExtra::grid.arrange(
  mark_r(plot_mat(Z_case_r, "case", limits = lim_r)),
  mark_r(plot_mat(Z_ctrl_r, "control", limits = lim_r)),
  mark_r(plot_diverging(Z_diff_r, "case - control")),
  ncol = 3
)
```

The biomarker circles in the case panel should be filled in with peaks that are **the same brightness** as any other peak on the canvas — no visual distinction. The control panel should have empty circles (no peak at those exact locations). The diff panel shows the systematic signal despite that visual overlap.

## Retrain the encoder on the realistic background

Principled evaluation requires a fresh encoder trained without leakage from the biomarker regions in *this* scenario. Even though the peak morphology distribution is similar to sanity-max, the specific peaks held out differ, so retraining is the right move.

```{r realistic-pretrain, cache = TRUE}
pretrain_r <- pretrain_denoising_online(
  bg_r,
  H = H, W = W,
  steps_per_epoch = 50L, epochs = 8L, batch_size = 16L,
  noise_scale = 3.0, val_n = 32L,
  device = "cpu", seed = 1L
)
encoder_r <- pretrain_r$encoder
```

## Encoder-channel sanity check on the realistic scenario

```{r realistic-encoder-check, fig.height = 3.5}
feats_r <- extract_encoder_features(encoder_r, Z_r, device = "cpu")
p_raw_r <- sapply(seq_len(64L), function(c)
  suppressWarnings(wilcox.test(feats_r[y_r == 2L, c],
                                feats_r[y_r == 1L, c])$p.value))
p_bh_r  <- p.adjust(p_raw_r, method = "BH")

sig_r_05  <- sum(p_bh_r < 0.05, na.rm = TRUE)
sig_r_001 <- sum(p_bh_r < 0.001, na.rm = TRUE)
sprintf("Realistic — Encoder channels significant at BH q < 0.05:  %d / 64", sig_r_05)
sprintf("Realistic — Encoder channels significant at BH q < 0.001: %d / 64", sig_r_001)

ggplot(tibble(channel = seq_len(64L), p_bh = p_bh_r),
       aes(x = channel, y = -log10(p_bh))) +
  geom_col(fill = "#B2182B") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "grey60") +
  labs(x = "encoder channel", y = "-log10(BH q)",
       title = "Realistic scenario: per-channel Wilcoxon by class") +
  theme_minimal()
```

## Classification and biomarker recovery

```{r realistic-classify}
elnet_r <- cv_encoder_elastic_net(
  encoder_r, Z_list = Z_r, y = y_r,
  k = 5L, alpha = 0.5, seed = 1L
)
sprintf("Realistic — case/control AUC (5-fold CV): %.3f", elnet_r$auc)
sprintf("Realistic — non-zero elastic-net coefs: %d / 64",
        sum(elnet_r$coefs != 0))
```

```{r realistic-saliency, fig.height = 4}
case_idx_r <- which(y_r == 2L); control_idx_r <- which(y_r == 1L)
orig_dims_r <- lapply(sim_r$samples,
                       function(s) c(H = nrow(s$Z), W = ncol(s$Z)))

sal_r <- aggregate_saliency_masked(encoder_r, Z_r,
                                     coefs = elnet_r$coefs,
                                     sample_indices = case_idx_r,
                                     orig_dims = orig_dims_r)
sal_ctrl_r <- aggregate_saliency_masked(encoder_r, Z_r,
                                          coefs = elnet_r$coefs,
                                          sample_indices = control_idx_r,
                                          orig_dims = orig_dims_r)
lim_s <- c(-1, 1) * max(abs(c(sal_r, sal_ctrl_r)))
gridExtra::grid.arrange(
  mark_r(plot_diverging(sal_r,      "saliency: cases", limits = lim_s)),
  mark_r(plot_diverging(sal_ctrl_r, "saliency: controls", limits = lim_s)),
  ncol = 2
)

recovery_r <- evaluate_biomarker_recovery(
  sal_r, biom_r, H = H, W = W,
  match_radius_rt = 8L, match_radius_cv = 2L, top_k = 5L
)
sprintf("Realistic — biomarker recovery per-pixel AUC: %.3f",
        recovery_r$auc_per_pixel)
sprintf("Realistic — recall @ top-5: %.3f",  recovery_r$recall_at_k)
sprintf("Realistic — precision @ top-5: %.3f", recovery_r$precision_at_k)
```

## Comparison: sanity-max vs realistic-intensity

```{r comparison}
tibble(
  metric = c("case/control AUC",
             "encoder channels (BH q < 0.05)",
             "encoder channels (BH q < 0.001)",
             "biomarker recovery per-pixel AUC",
             "recall @ top-5",
             "precision @ top-5"),
  sanity_max = c(sprintf("%.3f", elnet$auc),
                 sprintf("%d / 64", sig_alpha_05),
                 sprintf("%d / 64", sig_alpha_001),
                 sprintf("%.3f", recovery$auc_per_pixel),
                 sprintf("%.3f", recovery$recall_at_k),
                 sprintf("%.3f", recovery$precision_at_k)),
  realistic = c(sprintf("%.3f", elnet_r$auc),
                sprintf("%d / 64", sig_r_05),
                sprintf("%d / 64", sig_r_001),
                sprintf("%.3f", recovery_r$auc_per_pixel),
                sprintf("%.3f", recovery_r$recall_at_k),
                sprintf("%.3f", recovery_r$precision_at_k))
) |> knitr::kable(align = "lrr")
```

**What to read from this comparison:**

- **Case/control AUC drops but should stay well above chance** if the pipeline is honestly leveraging the presence/absence pattern rather than the intensity shift. Below ~0.85 on the realistic scenario means the classifier was primarily riding the intensity multiplier in sanity-max rather than learning "biomarker present at these coords".
- **Fewer significant encoder channels** is expected — signal in feature space is less pronounced without the intensity boost. But some should still be significant if the encoder is representing presence-vs-absence rather than intensity.
- **Biomarker recovery AUC** tells you whether saliency still points at the right spatial regions. If classification AUC is good but recovery AUC drops sharply, the classifier is finding the signal through some non-spatial shortcut — worth investigating.

If AUC on the realistic scenario stays above ~0.85, the pipeline is genuinely learning the presence/absence structure and not just riding a big intensity difference. That's the more realistic capability for actual biomarker discovery.

# Future experiments to plug in here

Ideas noted for later — not implemented yet:

- **Pretrain on a broader distribution than the simulation.** Currently the encoder is trained on `scenario$background_peak_params`, which is derived from the same base distribution as the simulation. Inflate `n_peaks$mean`, widen `size_jitter`, or pretrain from a different real cohort's params to test whether a more diverse pretraining distribution improves generalization.
- **Vary encoder pretraining cohort size.** Same as the classifier-cohort sweep in section 7, but retrain the encoder at each size. Isolates encoder-training-cohort effect from classifier-cohort effect.
- **Vary encoder architecture.** Bottleneck dimension (currently 64), stem strides, number of ResNet blocks. These are architecture changes so they need code edits, not just arg changes.
- **Intermediate intensity levels.** Sweep `case_intensity_mult` between 1.0 and 10.0 to find where classification AUC starts to fall off. Maps the "difficulty vs multiplier" curve.
- **Weaker presence/absence.** Combine the realistic-intensity setup with imperfect prevalence (e.g. `case_prev = 0.85, control_prev = 0.15`) to approach real-cohort conditions.
