---
title: "Walkthrough: pretrain_denoising_online step by step"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 3.5, dpi = 96)
library(classydms)
library(ggplot2)
library(dplyr)
library(tibble)

plot_mat <- function(mat, title = "", low = "white", high = "#B2182B",
                     limits = NULL, log = FALSE) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  scale <- if (log) {
    scale_fill_gradient(low = low, high = high, limits = limits,
                        trans = "log1p", name = "log(1+I)")
  } else {
    scale_fill_gradient(low = low, high = high, limits = limits,
                        name = "intensity")
  }
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() + scale + scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank())
}
```

# Summary

This walkthrough opens up `pretrain_denoising_online` step by step: what the training pair looks like, how a short run is configured, and what the returned object contains. Sections:

1. **What `pretrain_denoising_online` does** — Describes the self-supervised denoising task and its role as an encoder for downstream classification.
2. **Parameter cheat sheet** — Table of all arguments (sizing, optimization, generation, architecture, runtime) with defaults and rationale.
3. **Toy setup** — Hand-builds a `peak_params` for a fast 100x30 training run.
4. **One training pair — what the model actually sees** — Generates and plots a `(clean, noisy)` pair at elevated `noise_scale`.
5. **Train — a fast 5-epoch run** — Runs `pretrain_denoising_online` with tiny settings and shows the returned object structure.
6. **Loss curves** — Plots train and validation MSE per epoch with diagnostic guidance.
7. **Reconstruction on a fresh synthetic sample** — Feeds a new synthetic sample through the trained autoencoder and plots noisy/clean/reconstruction.
8. **The manifest** — Lists fields recorded in the on-disk manifest for reproducibility and resume.
9. **Pretraining from a peak catalog** — Shows how to swap `peak_params = NULL, catalog = my_catalog` to use the catalog generator.
10. **Practical tips** — Advice on `val_real`, `save_path`, `num_workers`, and which knobs to tune first.

# What `pretrain_denoising_online` does

Self-supervised training of a convolutional encoder on synthetic GC-DMS images. The encoder isn't given any class labels — it only sees pairs of `(noisy, clean)` synthetic samples and is asked to reconstruct `clean` from `noisy`. The point isn't the reconstruction itself; it's that this task forces the encoder to learn features that describe *the structure of GC-DMS peaks in general* (Vincent et al. 2008). Those features then transfer to downstream classification on real labeled data — even a very small labeled cohort — via a lightweight elastic net on the 64-dim encoder output.

"Online" means synthetic training batches are generated on the fly per step rather than pre-generated to disk. This lets the model see effectively unlimited unique synthetic samples with a constant memory footprint.

# Parameter cheat sheet

| Param | Default | Role |
|---|---|---|
| `peak_params` | — (required) | Output of `estimate_peak_params()` — the empirical distribution the generator samples from |
| `H`, `W` | — (required) | Input image dimensions in pixels (padded to a common size beforehand) |
| **Sizing** | | |
| `steps_per_epoch` | `500L` | Number of mini-batches per epoch |
| `epochs` | `30L` | Total epochs |
| `batch_size` | `32L` | Samples per gradient step |
| `val_n` | `200L` | Size of a *fixed* synthetic validation set for tracking |
| `val_real` | `NULL` | Optional list of real Z matrices for tracking real-data reconstruction MSE — critical when tuning `noise_scale` |
| **Optimization** | | |
| `lr` | `1e-3` | Adam learning rate |
| `weight_decay` | `1e-4` | L2 regularization on encoder weights |
| `grad_clip` | `1.0` | Clip global gradient norm to this; stability guard |
| `norm_clamp` | `10.0` | Clamp the batch loss to this; prevents a bad synthetic batch from tanking training |
| **Data generation** (all forwarded to `generate_one_synthetic`) | | |
| `add_noise` | `TRUE` | Whether to add background noise. `FALSE` collapses to a pure reconstruction AE (typically much less useful) |
| `noise_scale` | `1.0` | Multiplier on `params$noise$sd`. `1.0` = real-cohort noise; larger = harder denoising task = stronger regularization |
| `size_jitter` | `0.15` | SD of per-peak log-Normal size multiplier |
| `location_mode` | `"empirical"` | Bootstrap observed `(rt, cv)` (default) vs marginal Normal draws |
| `location_jitter_rt`, `location_jitter_cv` | `2`, `1` | SD of per-peak location jitter (pixels) |
| `attribute_mode` | `"joint"` | Bootstrap observed morphology triple (default) vs independent marginals |
| `dust_threshold` | `0` | Zero pixels below this on synthetic samples before normalization; match real-data `basement_thr` |
| **Architecture** | | |
| `stem_stride_rt`, `stem_stride_cv` | `4L`, `1L` | Encoder-stem downsampling per axis. RT gets aggressive downsampling because RT pixel count is much larger than CV |
| **Runtime** | | |
| `device` | auto | `"cpu"` / `"cuda"` / `"mps"` |
| `num_workers` | `1L` | Parallel R-side data generation via `mclapply` on Unix. Speed only |
| `num_threads` | `NULL` | Override torch's thread count. Apple Silicon benefits from setting to number of performance cores |
| `save_path` | `NULL` | Write encoder + autoencoder + manifest to disk each epoch. Set for any real run — a crash mid-training means recoverable state |
| `resume_from` | `NULL` | Path to a previously-saved encoder to resume from. Same-session use; hyperparameters must match the original |
| `checkpoint_every` | `10L` | Rotate a numbered checkpoint every N epochs |
| `loss_diagnostics` | `TRUE` | Print median / p95 / max batch loss each epoch |
| `seed` | `42L` | RNG seed |

**Why the defaults look the way they do:**

- **`lr = 1e-3` with Adam** — standard well-behaved LR for ResNet-style encoders in this size range. If loss plateaus early, halve; if unstable, halve.
- **`weight_decay = 1e-4`** — mild L2 regularization typical for small vision networks.
- **`grad_clip = 1.0` and `norm_clamp = 10.0`** — belt-and-suspenders against a single unlucky synthetic batch producing huge losses / gradients and destabilizing the model. Rarely trigger after epoch 1.
- **`size_jitter = 0.15`** was tuned down from a former default of `0.6` because `0.6` produced synthetic peaks noticeably larger or smaller than any real observed peak (not biologically plausible). `0.15` gives ~×1.16 multiplicative variation which matches typical GC-DMS peak-size variability.
- **`stem_stride_rt = 4L`, `stem_stride_cv = 1L`** — asymmetric downsampling reflects the asymmetric pixel dimensions (RT ~1400px vs CV ~40px). Uniform stride would either over-shrink CV or under-shrink RT.

# 1. Toy setup

To keep this fast we build a small `peak_params` by hand and train on a `100 × 30` canvas.

```{r toy-setup}
peaks_obs <- tibble(
  rt_loc    = c(0.2, 0.35, 0.5, 0.5, 0.7, 0.8, 0.85),
  cv_loc    = c(0.3, 0.5,  0.6, 0.4, 0.7, 0.4, 0.55),
  sigma_rt  = c(3,   6,    2,   4,   5,   3,   7),
  sigma_cv  = c(1.5, 2,    1,   1.5, 2,   1,   2.5),
  intensity = c(0.5, 1.2,  0.3, 0.8, 1.5, 0.4, 1.8)
)
toy_params <- list(
  n_peaks_detected = nrow(peaks_obs),
  n_peaks   = list(mean = 8, sd = 2),
  rt_loc    = list(mean = mean(peaks_obs$rt_loc), sd = sd(peaks_obs$rt_loc)),
  cv_loc    = list(mean = mean(peaks_obs$cv_loc), sd = sd(peaks_obs$cv_loc)),
  sigma_rt  = list(meanlog = mean(log(peaks_obs$sigma_rt)),
                   sdlog   = sd(log(peaks_obs$sigma_rt))),
  sigma_cv  = list(meanlog = mean(log(peaks_obs$sigma_cv)),
                   sdlog   = sd(log(peaks_obs$sigma_cv))),
  intensity = list(meanlog = mean(log(peaks_obs$intensity)),
                   sdlog   = sd(log(peaks_obs$intensity))),
  noise     = list(mean = 0.005, sd = 0.005),
  rt_loc_raw    = peaks_obs$rt_loc,
  cv_loc_raw    = peaks_obs$cv_loc,
  sigma_rt_raw  = peaks_obs$sigma_rt,
  sigma_cv_raw  = peaks_obs$sigma_cv,
  intensity_raw = peaks_obs$intensity
)
H <- 100L; W <- 30L
```

# 2. One training pair — what the model actually sees

Every training step is a batch of `(noisy, clean)` pairs. The encoder-decoder gets `noisy` as input and is trained to output `clean`. This is the "denoising" task.

```{r one-pair, fig.height = 3}
set.seed(1)
pair <- generate_one_synthetic(toy_params, H = H, W = W,
                                add_noise = TRUE, noise_scale = 3.0)
gridExtra::grid.arrange(
  plot_mat(pair$clean, "clean (training target)",
           limits = c(0, max(pair$noisy))),
  plot_mat(pair$noisy, "noisy (encoder input)",
           limits = c(0, max(pair$noisy))),
  ncol = 2
)
```

I set `noise_scale = 3.0` here (vs the default `1.0`) so the noise is visible on the plot. In a real run, sweep `c(1, 3, 5, 10)` and pick the highest value where `val_real` MSE stays flat.

# 3. Train — a fast 5-epoch run

```{r train, cache = TRUE}
result <- pretrain_denoising_online(
  toy_params,
  H = H, W = W,
  steps_per_epoch  = 30L,   # tiny for tutorial; production ~500
  epochs           = 5L,    # tiny; production ~30
  batch_size       = 16L,
  lr               = 1e-3,
  weight_decay     = 1e-4,
  add_noise        = TRUE,
  noise_scale      = 3.0,
  size_jitter      = 0.15,
  val_n            = 32L,
  device           = "cpu",
  seed             = 1L
)
```

The returned object:

```{r result-structure}
str(result, max.level = 1)
```

# 4. Loss curves

Two things to look at:

```{r loss-curves, fig.height = 3.5}
loss_df <- tibble(
  epoch = seq_along(result$loss_history),
  train_loss = result$loss_history,
  val_loss   = if (!is.null(result$val_loss_history))
                 result$val_loss_history else rep(NA, length(result$loss_history))
)
loss_long <- tidyr::pivot_longer(loss_df, cols = -epoch,
                                  names_to = "split", values_to = "mse")
ggplot(loss_long, aes(x = epoch, y = mse, color = split)) +
  geom_line(size = 1.1) + geom_point(size = 2) +
  scale_y_log10() +
  labs(y = "MSE (log scale)",
       title = "Training and validation MSE per epoch") +
  theme_minimal()
```

**What to look for in a real run:**

- **Both curves decreasing then plateauing** → healthy training. Stop around the plateau.
- **Training keeps dropping while validation flattens or rises** → overfitting the synthetic distribution. Reduce `epochs` or `steps_per_epoch`, or increase `weight_decay`.
- **Training and validation both flat and high** → the model isn't learning anything. Check that `noise_scale` isn't so high the input is uninformative, that `lr` isn't too small, and that the synthetic data isn't degenerate.
- **Loss spikes** → increase `grad_clip` (tighten) or `norm_clamp`.

At this tiny scale on 5 epochs the loss barely moves, so this plot is illustrative not diagnostic.

# 5. Reconstruction on a fresh synthetic sample

Feed a NEW synthetic sample through the trained autoencoder:

```{r reconstruction, fig.height = 3.5}
set.seed(99)
test_pair <- generate_one_synthetic(toy_params, H = H, W = W,
                                     add_noise = TRUE, noise_scale = 3.0)
x <- torch::torch_tensor(test_pair$noisy, dtype = torch::torch_float())$
  unsqueeze(1)$unsqueeze(1)
torch::with_no_grad({ recon <- result$autoencoder(x) })
recon_mat <- as.matrix(recon$squeeze()$to(device = "cpu"))

lim <- c(0, max(c(test_pair$noisy, recon_mat)))
gridExtra::grid.arrange(
  plot_mat(test_pair$noisy, "noisy input", limits = lim),
  plot_mat(test_pair$clean, "clean target", limits = lim),
  plot_mat(recon_mat,       "reconstruction", limits = lim),
  ncol = 3
)
```

Reconstruction is imperfect after only 5 epochs on a tiny dataset — the point of the plot is to show *the flow*. In a real run you'd want the reconstruction visibly similar to the clean target, especially at peak locations.

For real-data reconstruction QC (does the encoder generalize from synthetic to real?) use `reconstruct_real_samples()` with real preprocessed samples.

# 6. The manifest

If you passed `save_path`, three files are written each epoch: the encoder (`*.pt`), the full autoencoder (`*_autoencoder.pt`), and a manifest (`*_manifest.Rdata`). The manifest records the full hyperparameter configuration so a run is reproducible from disk:

```{r manifest-fields}
# str of what the manifest would contain (fields recorded per run)
manifest_fields <- c(
  "peak_params (the empirical distribution used)",
  "hyperparams (H, W, epochs, batch_size, lr, weight_decay, add_noise,",
  "  size_jitter, grad_clip, norm_clamp, val_n, dust_threshold,",
  "  location_mode, location_jitter_rt/cv, attribute_mode, noise_scale,",
  "  val_real_n, checkpoint_every, stem_stride_rt/cv, device, seed)",
  "loss_history (train MSE per epoch)",
  "val_loss_history (synthetic validation MSE per epoch)",
  "real_val_loss_history (if val_real was passed)",
  "batch_loss_stats (median/p95/max loss diagnostics per epoch)",
  "last_epoch_completed (for resume_from)"
)
cat(manifest_fields, sep = "\n")
```

# Pretraining from a peak catalog

An alternative to `peak_params`-based pretraining: pass `catalog = my_catalog` and set `peak_params = NULL`. The batch generator switches to `generate_one_synthetic_from_catalog` (per-compound Bernoulli firing) instead of `generate_one_synthetic` (per-sample random peaks from pool).

Why you'd prefer catalog-based pretraining: real GC-DMS samples share the same underlying compounds across subjects; the catalog preserves that shared-backbone structure so the encoder learns the correct inductive bias. See `catalog_pipeline_sanity_check` for an end-to-end example.

```{r catalog-pretrain, eval = FALSE}
# Assuming you already have a catalog:
pretrain_result <- pretrain_denoising_online(
  peak_params = NULL,
  catalog     = my_catalog,
  H = pad_dims$H, W = pad_dims$W,
  steps_per_epoch = 500L, epochs = 30L, batch_size = 32L,
  size_jitter = 0.15,     # still applies to per-compound morphology jitter
  noise_scale = 3.0,      # applies to catalog$noise$sd
  device = "mps",
  save_path = "encoder.pt"
)
```

Arguments that apply *only* to the peak_params flow (ignored when `catalog` is set): `location_mode`, `attribute_mode`.

# Practical tips

- **Always pass `val_real`** in real runs — it's the single most useful signal for whether synthetic pretraining generalizes.
- **Set `save_path`** — a crash mid-epoch loses hours otherwise.
- **`num_workers > 1`** helps a lot on GPU/MPS backends where the GPU otherwise sits idle waiting for R to prepare the next batch. A reasonable choice on M-series is `num_workers = N_performance_cores - 2`.
- **For a first run on a new cohort**: leave everything at defaults except `H`, `W`, `save_path`, and `val_real`. Only tune `noise_scale` and `size_jitter` if the encoder underperforms on classification.
