---
title: "Walkthrough: aggregate_saliency_masked step by step"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 3.5, dpi = 96)
library(classydms)
library(ggplot2)
library(dplyr)
library(tibble)

plot_mat <- function(mat, title = "", low = "white", high = "#B2182B",
                     limits = NULL) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient(low = low, high = high, limits = limits,
                        name = "intensity") +
    scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank())
}
plot_mat_diverging <- function(mat, title = "", limits = NULL) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  if (is.null(limits)) limits <- c(-1, 1) * max(abs(mat))
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                          midpoint = 0, limits = limits,
                          name = "saliency") +
    scale_y_reverse() +
    coord_fixed(ratio = 0.5) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 10) +
    theme(panel.grid = element_blank())
}
```

# What `aggregate_saliency_masked` does

Converts the elastic-net classifier's decision into a spatial `(RT, CV)` map showing *where in the image* the classifier is finding its signal. Answers the biomarker-discovery question: given that the model says a sample looks like a case, which coordinates drove the prediction?

The construction is a variant of **Class Activation Mapping (CAM)** adapted for this pipeline:

1. For one sample, get the encoder's per-channel spatial activation map `A[c, r, c']` (64 channels × H_enc × W_enc pixels at encoder resolution).
2. Weight each channel by its elastic-net coefficient `w[c]` — the same coefficients from `cv_encoder_elastic_net$coefs`. Positive coefs push toward class 2 (case), negative toward class 1.
3. Sum across channels: `CAM[r, c'] = Σ_c w[c] * A[c, r, c']`. This is a *linear combination* of encoder activations, weighted by the classifier — regions of the image where the case-pushing channels light up get high positive CAM values.
4. Upsample bilinearly from encoder resolution back to full image resolution.
5. **Aggregate across all cases**, ignoring padded regions, so per-sample noise averages out and the systematic class-associated regions show up.

Why CAM works: the classifier is *linear* in the global-average-pooled encoder features (elastic net is a linear model). Reversing the average — pushing the same weights back onto the pre-pool spatial map — gives a mathematically equivalent decomposition of the classifier score into per-pixel contributions.

# Parameter cheat sheet

| Param | Default | Role |
|---|---|---|
| `encoder` | — (required) | Pretrained `dms_encoder`. Same one used to build `coefs` |
| `Z_list` | — (required) | Named list of preprocessed, padded intensity matrices |
| `coefs` | — (required) | Elastic-net coefficients (length 64) — take from `cv_encoder_elastic_net$coefs` |
| `sample_indices` | — (required) | Which samples to aggregate over. Usually the case group; sometimes the whole cohort or specific subgroups |
| `orig_dims` | — (required) | List of `c(H = ..., W = ...)` pairs giving each sample's pre-padding dimensions. Used for masking — pixels that were padding in a given sample don't count toward its saliency |
| `device` | `"cpu"` | `"cpu"` / `"cuda"` / `"mps"` — where the encoder forward pass runs |

**Why the parameter design looks this way:**

- **`coefs` is separate from `encoder`** — the encoder is class-agnostic, learned in pretraining. `coefs` come from the labeled-data classifier. Keeping them separate lets you compute *different saliency maps for different classification tasks* using the same encoder (e.g., disease A vs control, then disease B vs control).
- **`sample_indices` is explicit** — you might want case-only, control-only (as a sanity check that it looks different), or all samples. The classifier signal comes primarily from cases; averaging over cases only usually gives the cleanest map.
- **`orig_dims` is required** — most cohorts pad samples to a common size, so a "sample" is actually valid signal in a subregion and zeros in the padded region. Averaging saliency over pixels that are padding for some samples would artificially deflate those regions. The mask fixes that.

# 1. Setup: build a mini cohort with a real signal

We need an encoder AND classifier coefs to demonstrate. The setup is the same as the elastic-net walkthrough — small toy params, hold out a bright biomarker, simulate a small case/control cohort, pretrain briefly, run the classifier.

```{r setup-cohort, cache = TRUE}
set.seed(1)
peaks_obs <- tibble(
  rt_loc    = c(0.2, 0.35, 0.5, 0.5, 0.7, 0.8, 0.85),
  cv_loc    = c(0.3, 0.5,  0.6, 0.4, 0.7, 0.4, 0.55),
  sigma_rt  = c(3,   6,    2,   4,   5,   3,   7),
  sigma_cv  = c(1.5, 2,    1,   1.5, 2,   1,   2.5),
  intensity = c(0.5, 1.2,  0.3, 0.8, 1.5, 0.4, 1.8)
)
toy_params <- list(
  n_peaks_detected = nrow(peaks_obs),
  n_peaks   = list(mean = 8, sd = 2),
  rt_loc    = list(mean = mean(peaks_obs$rt_loc), sd = sd(peaks_obs$rt_loc)),
  cv_loc    = list(mean = mean(peaks_obs$cv_loc), sd = sd(peaks_obs$cv_loc)),
  sigma_rt  = list(meanlog = mean(log(peaks_obs$sigma_rt)),
                   sdlog   = sd(log(peaks_obs$sigma_rt))),
  sigma_cv  = list(meanlog = mean(log(peaks_obs$sigma_cv)),
                   sdlog   = sd(log(peaks_obs$sigma_cv))),
  intensity = list(meanlog = mean(log(peaks_obs$intensity)),
                   sdlog   = sd(log(peaks_obs$intensity))),
  noise     = list(mean = 0.005, sd = 0.005),
  rt_loc_raw    = peaks_obs$rt_loc,
  cv_loc_raw    = peaks_obs$cv_loc,
  sigma_rt_raw  = peaks_obs$sigma_rt,
  sigma_cv_raw  = peaks_obs$sigma_cv,
  intensity_raw = peaks_obs$intensity
)
H <- 100L; W <- 30L

scenario <- biomarkers_level1(toy_params, seed = 42L)
sim <- simulate_case_control_cohort(
  scenario$background_peak_params, scenario$biomarkers,
  n_cases = 15L, n_controls = 15L,
  H = H, W = W, seed = 7L
)
Z_list <- lapply(sim$samples, `[[`, "Z")
y <- sim$y

pretrain_result <- pretrain_denoising_online(
  scenario$background_peak_params,
  H = H, W = W,
  steps_per_epoch = 30L, epochs = 5L, batch_size = 16L,
  noise_scale = 3.0, val_n = 32L, device = "cpu", seed = 1L
)
encoder <- pretrain_result$encoder

elnet <- cv_encoder_elastic_net(encoder, Z_list, y,
                                 k = 5L, alpha = 0.5, seed = 1L)
sprintf("Pretrain done. Case/control AUC = %.3f, %d non-zero coefs",
        elnet$auc, sum(elnet$coefs != 0))

# Biomarker's target location for reference
b_r <- scenario$biomarkers$rt_loc * H
b_c <- scenario$biomarkers$cv_loc * W
sprintf("Biomarker at RT pixel %.0f, CV pixel %.0f", b_r, b_c)
```

At this tiny scale (30 samples, 5 pretrain epochs, 100×30 image) the classifier signal is weak but this is enough to see the mechanism.

# 2. Under the hood: one sample's CAM

`aggregate_saliency_masked` calls `compute_saliency_map` per sample. Let's do it explicitly for one case:

```{r one-cam, fig.height = 4}
case_idx <- which(y == 2L & sim$per_sample_firing[, 1])[1]
Z_case <- Z_list[[case_idx]]

sal_one <- compute_saliency_map(encoder, Z_case, coefs = elnet$coefs,
                                 device = "cpu")
str(sal_one, max.level = 1)
```

Two versions come back:

```{r one-cam-plot, fig.height = 4}
gridExtra::grid.arrange(
  plot_mat_diverging(sal_one$raw_cam,
                     sprintf("raw_cam: encoder-resolution (%dx%d)",
                             nrow(sal_one$raw_cam), ncol(sal_one$raw_cam))),
  plot_mat_diverging(sal_one$upsampled_cam,
                     sprintf("upsampled_cam: input-resolution (%dx%d)",
                             nrow(sal_one$upsampled_cam),
                             ncol(sal_one$upsampled_cam))),
  ncol = 2
)
```

- **`raw_cam`** is at *encoder resolution*, which is smaller than the input because the encoder downsamples (stem strides). For a 100×30 input with stem strides 4×1, the raw CAM is about 25×30.
- **`upsampled_cam`** is bilinearly interpolated to the input size (100×30). Uses `torch::nnf_interpolate` internally. Necessary for aligning saliency with input-space coordinates.

Red = classifier is pushing toward case at that region; blue = toward control; white = neutral or zero. On a single sample the map is noisy — even if there's a real biomarker signal, one sample's activation carries a lot of per-sample idiosyncrasy.

# 3. Averaging across cases — this is what aggregation does

The point of `aggregate_saliency_masked` is: **the classifier signal is systematic (same location across cases) while per-sample noise is random**. Averaging over many cases cancels the noise and reinforces the signal.

```{r aggregation, fig.height = 4}
case_indices    <- which(y == 2L)
control_indices <- which(y == 1L)
orig_dims <- lapply(sim$samples,
                     function(s) c(H = nrow(s$Z), W = ncol(s$Z)))

sal_cases    <- aggregate_saliency_masked(encoder, Z_list,
                                            coefs = elnet$coefs,
                                            sample_indices = case_indices,
                                            orig_dims = orig_dims)
sal_controls <- aggregate_saliency_masked(encoder, Z_list,
                                            coefs = elnet$coefs,
                                            sample_indices = control_indices,
                                            orig_dims = orig_dims)

mark <- function(p, r, c)
  p + annotate("point", x = c, y = r, shape = 21, color = "cyan",
               fill = NA, size = 8, stroke = 1.2)

lim <- c(-1, 1) * max(abs(c(sal_cases, sal_controls)))
gridExtra::grid.arrange(
  mark(plot_mat_diverging(sal_cases,    "aggregated saliency: cases",
                          limits = lim), b_r, b_c),
  mark(plot_mat_diverging(sal_controls, "aggregated saliency: controls",
                          limits = lim), b_r, b_c),
  ncol = 2
)
```

**How to read this on a real run:**

- **Cases panel**: red regions are where the classifier consistently says "this looks like a case". The biomarker location (cyan circle) should show up red if the pipeline is working.
- **Controls panel**: the same regions should show *blue* — where controls are systematically less like cases at those coordinates. Alternatively they might be near zero because the biomarker peak is simply absent (no signal to weight).
- **Consistent red-in-cases + blue-in-controls at the biomarker location** = the classifier learned the injected signal AND localized it correctly.

At this tiny scale (5 pretrain epochs) the maps may look mostly gray — that's expected. On a real run with meaningful pretraining, hotspots at biomarker locations are usually clear.

# 4. The masking mechanism — why `orig_dims` matters

Every sample enters as an `H × W` padded matrix, but the actual data occupies only `(orig_H, orig_W)` in the top-left. `aggregate_saliency_masked` uses `orig_dims` to build a **per-sample mask**:

```r
# From R/interpretability.R, inside the aggregation loop
mask <- matrix(0, H, W)
mask[seq_len(orig_H), seq_len(orig_W)] <- 1
sum_mat   <- sum_mat + cam * mask       # only real-data pixels contribute
count_mat <- count_mat + mask            # per-pixel divisor
```

Without this, aggregation would treat padded (zero-input) regions as "the classifier said neutral here", which artificially deflates the average toward zero at pixels that are padding for even one sample.

To see the effect explicitly, compare padded and pre-padded coordinate systems:

```{r pad-viz, fig.height = 3}
# In this toy setup all samples share H x W so masking is a no-op.
# On a real cohort orig_dims varies per sample and masking matters.
tbl <- tibble(
  sample = seq_along(orig_dims),
  orig_H = sapply(orig_dims, `[`, "H"),
  orig_W = sapply(orig_dims, `[`, "W")
)
head(tbl)
sprintf("Padded H x W: %d x %d. Orig dims range: H %d..%d, W %d..%d",
        H, W, min(tbl$orig_H), max(tbl$orig_H),
        min(tbl$orig_W), max(tbl$orig_W))
```

In this tutorial all samples share dimensions because we generated them at `H × W`, so masking is a no-op. On a real cohort where samples were trimmed to slightly different sizes before padding, the mask ensures each pixel's average only counts samples that had real data there.

# 5. Using the map — what to do with a saliency hotspot

Once you have `sal_cases`, the practical follow-ups:

- **Find local maxima** — `top_local_maxima(sal_cases, top_k = 10)` gives the K brightest hotspots as `(RT_pixel, CV_pixel, value)`. That's your candidate biomarker locations.
- **Map back to physical units** — the RT and CV axes in your original files (`samples[[i]]$time` and `samples[[i]]$cv`) tell you what those pixel coordinates mean in seconds and volts. That's what you'd cross-reference with a compound library.
- **Compare with controls** — if you also compute `sal_controls`, subtracting gives you a *class-difference saliency* that highlights regions where the signal is specific to cases rather than shared.
- **Evaluate against ground truth** (in a simulation): `evaluate_biomarker_recovery(sal_cases, biomarkers, ...)` returns per-pixel AUC and top-K precision/recall.

# Practical tips

- **Always use `sample_indices` = the case group** for the primary saliency map. Aggregating over all samples smears the signal because control saliencies contribute opposite-sign contributions.
- **Averaging is a stability tool, not a magic filter** — if the underlying classifier AUC is chance-level, no amount of aggregation will produce a meaningful map. Check `cv_encoder_elastic_net$auc` first; if it's ≤ 0.60 the saliency isn't reliable.
- **`raw_cam` vs `upsampled_cam`** — you almost always want the upsampled version for interpretation. The raw version is useful for debugging (checking that encoder channels are firing at all).
- **`coefs` from `cv_encoder_elastic_net`** are averaged across CV folds. Alternative: run `cv_encoder_elastic_net` yourself with `k = 1` for a single-fold classifier if you want no averaging, or aggregate coefs by median instead of mean.
