---
title: "Walkthrough: generate_one_synthetic step by step"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.5, dpi = 96)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)

# Small dimensions so heatmaps stay readable
H <- 100L   # RT rows
W <- 30L    # CV cols

# Consistent heatmap helper — used everywhere below.
plot_mat <- function(mat, title = "", low = "white",
                     high = "#B2182B", limits = NULL) {
  H <- nrow(mat); W <- ncol(mat)
  df <- expand.grid(rt = seq_len(H), cv = seq_len(W))
  df$val <- as.vector(mat)
  ggplot(df, aes(x = cv, y = rt, fill = val)) +
    geom_tile() +
    scale_fill_gradient(low = low, high = high, limits = limits,
                        name = "intensity") +
    scale_y_reverse() +
    coord_fixed(ratio = 1) +
    labs(x = "CV pixel", y = "RT pixel", title = title) +
    theme_minimal(base_size = 11) +
    theme(panel.grid = element_blank())
}

set.seed(1)
```

# Summary

This walkthrough follows `generate_one_synthetic()` end to end (`location_mode = "empirical"`, `attribute_mode = "joint"`) on a small canvas so every step is visible. Sections:

1. **The building block — one 2D Gaussian "splotch"** — Constructs a single 2D Gaussian from an outer product and shows the effect of sigma and amplitude.
2. **Where do the numbers come from? — `peak_params`** — Introduces raw vs fitted fields of `peak_params` and plots the 7 toy observed peaks.
3. **Draw ONE peak using the default pipeline** — Manually reproduces the empirical + joint bootstrap plus location and size jitter to place one peak.
4. **Stack N peaks into a full sample** — Samples the peak count, pre-samples indices, and loops peak accumulation while snapshotting the canvas.
5. **Add noise — folded Normal** — Adds i.i.d. folded-Normal noise, plots the texture on its own scale, and compares clean vs noisy at the peak scale.
6. **Sanity check: call the real function** — Confirms `generate_one_synthetic()` produces the same character of output.
7. **Appendix: distribution reference** — Table of every random draw with its distribution and parameter source.

This walkthrough follows `generate_one_synthetic()` in its **default** configuration (`location_mode = "empirical"`, `attribute_mode = "joint"`) — the recommended pipeline as of the most recent commits. Every step visualizes what is happening on a small `r H` × `r W` canvas so plots stay readable.

## Cheat sheet: parametric vs empirical

Every random draw in the function is one of two types:

| Step | Distribution | Type |
|---|---|---|
| Number of peaks | `Normal(n_peaks$mean, sd)` | **parametric** |
| Row index `j` into observed peaks | `sample.int(n_obs)` uniform on `1..n_obs` | **empirical** (bootstrap) |
| Peak center location on axis | `rt_loc_raw[j], cv_loc_raw[j]` + `Normal(0, jitter)` | **empirical** + parametric jitter |
| Peak morphology (sigma_rt, sigma_cv, intensity) | `(sigma_rt_raw[j], sigma_cv_raw[j], intensity_raw[j])` | **empirical** joint triple |
| Per-peak scale factor | `log-Normal(0, size_jitter²)` | **parametric** |
| Background noise | `folded Normal(noise$mean, noise$sd)` per pixel | **parametric** |

There is **no multivariate Normal anywhere.** Joint distributions are handled by resampling observed rows, not by fitting an MVN.

---

# 1. The building block — one 2D Gaussian "splotch"

A single synthetic peak is a 2D Gaussian on the image grid. It has 5 numbers:

- `mu_rt`, `mu_cv` — center coordinates (in pixels)
- `sig_rt`, `sig_cv` — width along each axis (in pixels, one standard deviation)
- `amp` — peak height

## The math

A 2D Gaussian aligned to the RT and CV axes is *separable*:

$$G(r, c) = \exp\!\left(-\frac{1}{2}\left(\frac{r - \mu_{RT}}{\sigma_{RT}}\right)^2\right) \cdot \exp\!\left(-\frac{1}{2}\left(\frac{c - \mu_{CV}}{\sigma_{CV}}\right)^2\right)$$

So we compute two 1D Gaussians — one along RT, one along CV — and take their outer product to get the 2D shape.

```{r one-splotch}
mu_rt <- 40; mu_cv <- 15
sig_rt <- 4; sig_cv <- 1.5
amp <- 1.0

rows <- seq_len(H)                          # 1..H, positions along RT axis
cols <- seq_len(W)                          # 1..W, positions along CV axis

g_rt <- exp(-0.5 * ((rows - mu_rt) / sig_rt)^2)   # length H, 1D Gaussian along RT
g_cv <- exp(-0.5 * ((cols - mu_cv) / sig_cv)^2)   # length W, 1D Gaussian along CV

splotch <- amp * outer(g_rt, g_cv)          # H x W matrix, the 2D Gaussian
plot_mat(splotch, title = "One synthetic peak (mu_rt=40, mu_cv=15, sig_rt=4, sig_cv=1.5, amp=1)")
```

The 1D Gaussians themselves are worth seeing on their own — the outer product just multiplies them element-wise:

```{r one-splotch-marginals, fig.height = 2.5}
mar_df <- rbind(
  tibble(axis = "g_rt (along RT)", pos = rows, val = g_rt),
  tibble(axis = "g_cv (along CV)", pos = cols, val = g_cv)
)
ggplot(mar_df, aes(x = pos, y = val)) +
  geom_line() + facet_wrap(~ axis, scales = "free") +
  labs(x = "pixel", y = "value") + theme_minimal()
```

## Effect of sigma

Change `sig_rt`, `sig_cv` while holding everything else fixed:

```{r splotch-sigma, fig.height = 3}
mkone <- function(sr, sc) {
  amp * outer(exp(-0.5 * ((rows - mu_rt) / sr)^2),
              exp(-0.5 * ((cols - mu_cv) / sc)^2))
}
gridExtra::grid.arrange(
  plot_mat(mkone(2, 1),    title = "sig_rt=2, sig_cv=1"),
  plot_mat(mkone(8, 1),    title = "sig_rt=8, sig_cv=1 (tall)"),
  plot_mat(mkone(4, 3),    title = "sig_rt=4, sig_cv=3 (wide)"),
  ncol = 3
)
```

## Effect of amplitude

Amplitude just rescales — same shape, brighter or dimmer. Because the 1D Gaussians are unnormalized (`exp(-0.5·z²)` with no `1/(σ√(2π))` in front), the peak-center value is exactly `amp`.

```{r splotch-amp, fig.height = 3}
gridExtra::grid.arrange(
  plot_mat(0.5 * splotch, title = "amp=0.5", limits = c(0, 1.5)),
  plot_mat(splotch,       title = "amp=1.0", limits = c(0, 1.5)),
  plot_mat(1.5 * splotch, title = "amp=1.5", limits = c(0, 1.5)),
  ncol = 3
)
```

This is the atom the whole generator is built from. Everything else in `generate_one_synthetic` is machinery for choosing the numbers `(mu_rt, mu_cv, sig_rt, sig_cv, amp)` and repeating.

---

# 2. Where do the numbers come from? — `peak_params`

Upstream, `estimate_peak_params()` walks every real sample, runs the peak detector, and pools every detected peak's coordinates and morphology into big parallel vectors. Two families of information are stored:

- **Raw observations** — one entry per detected peak: `rt_loc_raw`, `cv_loc_raw`, `sigma_rt_raw`, `sigma_cv_raw`, `intensity_raw`. Position `j` in every one of these vectors is the same real peak.
- **Fitted summary distributions** — one number per statistic: `n_peaks$mean/sd`, `rt_loc$mean/sd`, `sigma_rt$meanlog/sdlog`, etc. Used only when you opt into `"marginal"` mode.

For this walkthrough we skip the real cohort and hand-build a tiny toy `params` object with 7 imaginary observed peaks:

```{r toy-params}
peaks_obs <- tibble(
  rt_loc    = c(0.20, 0.35, 0.50, 0.50, 0.70, 0.80, 0.85),
  cv_loc    = c(0.30, 0.50, 0.60, 0.40, 0.70, 0.40, 0.55),
  sigma_rt  = c(3.0,  6.0,  2.0,  4.0,  5.0,  3.0,  7.0),
  sigma_cv  = c(1.5,  2.0,  1.0,  1.5,  2.0,  1.0,  2.5),
  intensity = c(0.5,  1.2,  0.3,  0.8,  1.5,  0.4,  1.8)
)
peaks_obs

toy_params <- list(
  n_peaks_detected = nrow(peaks_obs),
  n_peaks   = list(mean = 8, sd = 2),
  rt_loc    = list(mean = mean(peaks_obs$rt_loc), sd = sd(peaks_obs$rt_loc)),
  cv_loc    = list(mean = mean(peaks_obs$cv_loc), sd = sd(peaks_obs$cv_loc)),
  sigma_rt  = list(meanlog = mean(log(peaks_obs$sigma_rt)),
                   sdlog   = sd(log(peaks_obs$sigma_rt))),
  sigma_cv  = list(meanlog = mean(log(peaks_obs$sigma_cv)),
                   sdlog   = sd(log(peaks_obs$sigma_cv))),
  intensity = list(meanlog = mean(log(peaks_obs$intensity)),
                   sdlog   = sd(log(peaks_obs$intensity))),
  noise     = list(mean = 0.005, sd = 0.005),
  rt_loc_raw    = peaks_obs$rt_loc,
  cv_loc_raw    = peaks_obs$cv_loc,
  sigma_rt_raw  = peaks_obs$sigma_rt,
  sigma_cv_raw  = peaks_obs$sigma_cv,
  intensity_raw = peaks_obs$intensity
)
```

Visualize the 7 observed peak positions on our `r H` × `r W` canvas:

```{r toy-params-plot, fig.height = 3.2}
peaks_obs %>%
  mutate(
    rt_pixel = rt_loc * H,
    cv_pixel = cv_loc * W,
    label    = as.character(seq_len(n()))
  ) %>%
  ggplot(aes(x = cv_pixel, y = rt_pixel, size = intensity, label = label)) +
  geom_point(alpha = 0.6, color = "#B2182B") +
  geom_text(size = 3, nudge_x = 1, hjust = 0) +
  scale_y_reverse(limits = c(H, 0)) +
  xlim(0, W) + coord_fixed() +
  scale_size_continuous(range = c(3, 10)) +
  labs(x = "CV pixel", y = "RT pixel",
       title = "The 7 observed peaks in toy_params",
       size = "observed intensity") +
  theme_minimal()
```

**Locations are stored as fractions of image dimensions** (0–1). That way one `peak_params` object can be used to generate synthetic samples at any target `(H, W)` — the position in pixels is computed as `rt_loc * H` at generation time.

---

# 3. Draw ONE peak using the default pipeline

Now we mimic exactly what `generate_one_synthetic` does for a single peak in `location_mode = "empirical"`, `attribute_mode = "joint"`.

## 3.1 Pick an index into the observed pool — *empirical*

The atomic operation is: draw a row index `j` uniformly at random from `1..n_obs`. All 5 attributes of the synthetic peak will be inherited from that row.

```{r pick-j}
n_obs <- length(toy_params$rt_loc_raw)   # 7
j <- sample.int(n_obs, size = 1)
j
peaks_obs[j, ]
```

Under the default (both empirical location and joint attributes), the **same** `j` supplies both the location and the morphology. That single choice is the entire empirical draw.

## 3.2 Location — empirical bootstrap plus a small parametric jitter

```{r location-draw}
location_jitter_rt <- 2  # SD in pixels
location_jitter_cv <- 1

mu_rt <- toy_params$rt_loc_raw[j] * H + rnorm(1, 0, location_jitter_rt)
mu_cv <- toy_params$cv_loc_raw[j] * W + rnorm(1, 0, location_jitter_cv)
mu_rt <- max(1, min(H, mu_rt))    # clamp to canvas
mu_cv <- max(1, min(W, mu_cv))
c(mu_rt = mu_rt, mu_cv = mu_cv)
```

The bootstrap picks one of the 7 known hotspots. The **jitter** — a `Normal(0, σ)` displacement of a couple of pixels — models the fact that even the same compound doesn't land at pixel-exact positions between real runs. Jitter is parametric with a tunable SD; the location itself is empirical.

## 3.3 Morphology — empirical joint triple

Because we're in joint mode, the widths and amplitude all come from the same row `j`:

```{r morph-draw}
sig_rt   <- toy_params$sigma_rt_raw[j]
sig_cv   <- toy_params$sigma_cv_raw[j]
base_amp <- toy_params$intensity_raw[j]
c(sig_rt = sig_rt, sig_cv = sig_cv, base_amp = base_amp)
```

**Why it matters that this is joint**: in real GC-DMS, wider peaks tend to be brighter (more compound → both wider RT footprint and higher signal). By resampling all three together we preserve that correlation exactly — no MVN fitting required. If we had chosen `attribute_mode = "marginal"`, `sig_rt`, `sig_cv`, and `base_amp` would each be drawn from independent log-Normals fitted to the marginals of the pool, destroying the joint structure.

## 3.4 Per-peak size jitter — always-on parametric log-Normal

Even in joint mode, a small multiplicative perturbation is applied so we don't just produce exact copies of the 7 observed peaks:

```{r size-jitter}
size_jitter <- 0.15
set.seed(3)   # so scale_factor lands near 1.0 for pedagogy
scale_factor <- exp(rnorm(1, mean = 0, sd = size_jitter))  # log-Normal(0, size_jitter²)
sig_rt <- max(sig_rt * scale_factor, 0.8)
sig_cv <- max(sig_cv * scale_factor, 0.8)
amp    <- base_amp * (0.5 + 0.5 * scale_factor)
c(scale_factor = scale_factor, sig_rt = sig_rt, sig_cv = sig_cv, amp = amp)
```

- **Log-Normal** because widths and intensity are positive multiplicative quantities.
- The `0.8` floor on sigmas prevents pathologically thin single-pixel peaks.
- Amplitude is scaled *half* as aggressively as sigma (`0.5 + 0.5 * scale_factor`) because empirically peak height doesn't grow proportionally with peak width.

> **Note on `size_jitter` magnitude.** The *function default* is `size_jitter = 0.6`, which gives multiplicative variation of roughly ×1.8 or ÷1.8 (one SD in log-space of 0.6). That is intentionally loose — good for regularization when the observed pool is small — but produces synthetic peaks noticeably bigger or smaller than any observed peak, which can look wrong. The `getting-started` vignette tunes it down to **`size_jitter = 0.15`** (SD in log-space of 0.15 → multiplicative ×1.16 or ÷1.16), which matches real GC-DMS peak-size variability much better and is what should be used in production. We follow the vignette here so that `scale_factor ≈ 1` and the sampled peak stays close to the observed morphology.

## 3.5 Rasterize the 2D Gaussian

Same math as Section 1, but now with the sampled parameters:

```{r one-full-peak}
g_rt <- exp(-0.5 * ((rows - mu_rt) / sig_rt)^2)
g_cv <- exp(-0.5 * ((cols - mu_cv) / sig_cv)^2)
peak_k <- amp * outer(g_rt, g_cv)

plot_mat(peak_k,
         title = sprintf("One sampled peak: j=%d, mu=(%.1f, %.1f), sig=(%.2f, %.2f), amp=%.2f",
                         j, mu_rt, mu_cv, sig_rt, sig_cv, amp))
```

That is one splotch, drawn end-to-end using the default pipeline. Compare to the entry in `peaks_obs[j, ]` above — the center is near that observed hotspot, the shape is derived from that observed peak's morphology, and the size_jitter has stretched/shrunk it slightly.

---

# 4. Stack N peaks into a full sample

## 4.1 Decide how many peaks — parametric Normal

```{r n-peaks-draw}
n_peaks <- max(1L, round(rnorm(1,
  mean = toy_params$n_peaks$mean,
  sd   = max(toy_params$n_peaks$sd,
             toy_params$n_peaks$mean * 0.3))))
n_peaks
```

The `sd` has a `0.3 * mean` floor — if the observed cohort happened to have unusually tight peak counts, we deliberately widen the distribution. This keeps the synthetic set from all having exactly the same peak count.

## 4.2 Pre-sample all N indices at once

```{r preindex}
loc_idx  <- sample.int(n_obs, size = n_peaks, replace = TRUE)
attr_idx <- loc_idx    # joint mode + empirical location share the same index
loc_idx
```

## 4.3 Loop and accumulate

Additive superposition — each new peak's 2D Gaussian is *added* to the running canvas. Overlapping peaks sum. We record the running canvas after every peak so we can watch the sample build up.

```{r stack, fig.height = 6}
set.seed(7)   # a seed that spreads picks across observed peaks
clean <- matrix(0, nrow = H, ncol = W)
snapshots <- list()

for (k in seq_len(n_peaks)) {
  j <- loc_idx[k]
  mu_rt <- toy_params$rt_loc_raw[j] * H + rnorm(1, 0, location_jitter_rt)
  mu_cv <- toy_params$cv_loc_raw[j] * W + rnorm(1, 0, location_jitter_cv)
  mu_rt <- max(1, min(H, mu_rt)); mu_cv <- max(1, min(W, mu_cv))

  ja <- attr_idx[k]
  sig_rt   <- toy_params$sigma_rt_raw[ja]
  sig_cv   <- toy_params$sigma_cv_raw[ja]
  base_amp <- toy_params$intensity_raw[ja]

  scale_factor <- exp(rnorm(1, 0, size_jitter))
  sig_rt <- max(sig_rt * scale_factor, 0.8)
  sig_cv <- max(sig_cv * scale_factor, 0.8)
  amp    <- base_amp * (0.5 + 0.5 * scale_factor)

  g_rt <- exp(-0.5 * ((rows - mu_rt) / sig_rt)^2)
  g_cv <- exp(-0.5 * ((cols - mu_cv) / sig_cv)^2)
  clean <- clean + amp * outer(g_rt, g_cv)

  snapshots[[k]] <- clean
}

# Fix the color scale across ALL panels so intensity accumulation
# is visible (otherwise ggplot auto-renormalizes each panel and every
# stage looks the same).
final_max <- max(clean)
ks <- if (n_peaks >= 4L) {
  unique(round(seq(1, n_peaks, length.out = 6L)))
} else {
  seq_len(n_peaks)
}
plots <- lapply(ks, function(k)
  plot_mat(snapshots[[k]],
           title = sprintf("After %d of %d peaks", k, n_peaks),
           limits = c(0, final_max)))
do.call(gridExtra::grid.arrange, c(plots, ncol = 3))
```

**How to read this**: the color scale is fixed across all panels at `[0, max(clean)]`. Bright cells that only appear in later panels mean new peaks landed in previously-quiet regions. Bright cells that get brighter across panels mean peaks are stacking on top of each other (overlapping peaks *add*, so the same location can accumulate intensity from multiple draws).

If you saw this at auto-scale (each panel independently rescaled) every panel would look similar because the ratio between the brightest and dimmest bins stays roughly constant. Fixing the scale is what lets you see the growth.

The final panel is the **clean** synthetic sample. Now we add noise.

---

# 5. Add noise — folded Normal

```{r noise, fig.height = 3}
noise_sd_use <- max(1e-8, toy_params$noise$sd)
noise <- matrix(
  abs(rnorm(H * W, mean = toy_params$noise$mean, sd = noise_sd_use)),
  nrow = H, ncol = W
)
noisy <- clean + noise
```

- Each pixel is an **independent** draw from `Normal(noise$mean, noise$sd)`.
- The `abs()` folds the negative half onto the positive half (a *folded Normal* distribution) — real ion counts can't be negative.
- Noise is i.i.d. per pixel; there is no spatial correlation modeled. Fine as a regularizer for denoising-autoencoder training (this is Vincent et al.'s argument), but a limitation compared to real instrument noise which has row-correlated drift.

### Noise at its own scale — the texture

Because real GC-DMS noise is only ~0.5–1% of typical peak amplitudes (`noise$sd = 0.005` vs peak amps ~0.5–1.5 in this toy example), plotting noise on the same scale as peaks makes it invisible. Give the noise panel its own auto scale first so the per-pixel texture is legible:

```{r noise-alone, fig.height = 3.3}
plot_mat(noise, title = sprintf(
  "noise only, own color scale (max = %.4f)", max(noise)))
```

That's the **folded Normal** at work — each pixel is an independent draw, so there is no spatial structure, just fine speckle. The color-bar max is ~0.02, less than 1% of any peak.

### Clean vs noisy at the shared peak scale

```{r noise-plots, fig.height = 3.5}
gridExtra::grid.arrange(
  plot_mat(clean, title = "clean (peaks only)",
           limits = c(0, max(noisy))),
  plot_mat(noisy, title = "noisy = clean + noise (encoder input)",
           limits = c(0, max(noisy))),
  ncol = 2
)
```

At the peak-intensity scale, `clean` and `noisy` look nearly identical — which is exactly the point. Real GC-DMS noise is small relative to peaks; the **denoising** pretraining task is not "erase a lot of noise," it's a regularization trick that forces the encoder to route information through its bottleneck rather than learn the trivial identity mapping (Vincent et al. 2008). If noise dominated the image visually, real peaks would be unrecoverable, and pretraining would fail for a different reason.

`generate_one_synthetic()` returns `list(clean = clean, noisy = noisy)`. During denoising pretraining the encoder gets `noisy` as input and is asked to reconstruct `clean` — that's the denoising task.

---

# 6. Sanity check: call the real function

We built each piece manually. Now confirm the real `generate_one_synthetic()` produces the same kind of output. Random seeds will differ so the exact peaks won't match, but the character of the image should be identical.

```{r end-to-end, eval = requireNamespace("classydms", quietly = TRUE), fig.height = 3.5}
library(classydms)
set.seed(42)
result <- generate_one_synthetic(
  toy_params, H = H, W = W,
  add_noise = TRUE,
  size_jitter = 0.15,                                   # match production, not function default 0.6
  location_mode = "empirical",
  location_jitter_rt = location_jitter_rt,
  location_jitter_cv = location_jitter_cv,
  attribute_mode = "joint"
)

gridExtra::grid.arrange(
  plot_mat(result$clean, title = "generate_one_synthetic()$clean",
           limits = c(0, max(result$noisy))),
  plot_mat(result$noisy, title = "generate_one_synthetic()$noisy",
           limits = c(0, max(result$noisy))),
  ncol = 2
)
```

# Appendix: distribution reference

Every random draw in `generate_one_synthetic`, with the parameter that governs it and its role:

| Random draw | Distribution | Parameter source | Purpose |
|---|---|---|---|
| Number of peaks per sample | `Normal(μ, max(σ, 0.3·μ))` | `params$n_peaks$mean/sd` | how "busy" this sample is |
| Row index `j` | uniform on `1..n_obs` | `length(params$rt_loc_raw)` | pick one real observed peak to imitate |
| Per-peak location jitter | `Normal(0, σ_loc)` per axis | `location_jitter_rt`, `location_jitter_cv` | instrument reproducibility around each hotspot |
| Per-peak size factor | `log-Normal(0, size_jitter²)` | `size_jitter` | regularization; avoid exact copies |
| Per-pixel background | folded `Normal(μ, σ)` (i.i.d.) | `params$noise$mean/sd` | denoising target |
| **Location** `(rt, cv)` | `(rt_loc_raw[j], cv_loc_raw[j])` | empirical pool | joint spatial distribution |
| **Morphology** `(σ_rt, σ_cv, amp)` | `(sigma_rt_raw[j], sigma_cv_raw[j], intensity_raw[j])` | empirical pool | joint width–intensity coupling |

**No multivariate Normal is fit anywhere.** Joint distributions (`(rt, cv)` and `(σ_rt, σ_cv, amp)`) are preserved by resampling observed rows rather than by fitting an MVN. Both have the same joint index `j` in the default configuration — that's what couples location to morphology when you resample.
