---
title: "`r sprintf('Peak-parameters report — %s', params$cohort_name)`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
params:
  peak_params: NA
  y: NA
  batch: NA
  class_labels: NA
  cohort_name: "cohort"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 96)
library(ggplot2)
library(dplyr)
library(tibble)

pp <- params$peak_params
y  <- params$y
batch <- params$batch

# --- Per-peak long table -----------------------------------------------
peaks <- tibble::tibble(
  sample_idx = pp$sample_idx_raw,
  rt_loc     = pp$rt_loc_raw,
  cv_loc     = pp$cv_loc_raw,
  sigma_rt   = pp$sigma_rt_raw,
  sigma_cv   = pp$sigma_cv_raw,
  intensity  = pp$intensity_raw
)

# --- Per-sample summary ------------------------------------------------
sample_names <- if (!is.null(pp$sample_names)) {
  pp$sample_names
} else {
  as.character(seq_len(pp$n_samples))
}
if (length(sample_names) < pp$n_samples) {
  sample_names <- c(sample_names,
                    as.character(seq(length(sample_names) + 1L,
                                     pp$n_samples)))
}

samples_df <- tibble::tibble(
  sample_idx  = seq_len(pp$n_samples),
  sample_name = sample_names
)

# Attach class labels
has_y <- !is.null(y) && length(y) > 0
if (has_y) {
  # Accept either names(y) matching sample_names or a positional vector
  if (!is.null(names(y))) {
    lookup <- y[samples_df$sample_name]
  } else if (length(y) == pp$n_samples) {
    lookup <- y
  } else {
    warning("y provided but neither named-with-sample-names nor length-",
            pp$n_samples, "; ignoring y for class comparisons.")
    lookup <- rep(NA, pp$n_samples)
    has_y <- FALSE
  }
  # Optional remap of numeric / string class values to display labels.
  cls_char <- as.character(lookup)
  cl <- params$class_labels
  if (!is.null(cl) && length(cl) > 0 && !is.null(names(cl))) {
    remap <- cl[cls_char]
    cls_char <- ifelse(is.na(remap), cls_char, unname(remap))
  }
  samples_df$class <- factor(cls_char)
}

has_batch <- !is.null(batch) && length(batch) > 0
if (has_batch) {
  if (!is.null(names(batch))) {
    b_lookup <- batch[samples_df$sample_name]
  } else if (length(batch) == pp$n_samples) {
    b_lookup <- batch
  } else {
    warning("batch provided but neither named-with-sample-names nor length-",
            pp$n_samples, "; ignoring batch for coloring.")
    b_lookup <- rep(NA, pp$n_samples)
    has_batch <- FALSE
  }
  samples_df$batch <- factor(b_lookup)
}

# Per-sample peak count
n_peaks_per_sample <- as.integer(table(factor(peaks$sample_idx,
                                              levels = seq_len(pp$n_samples))))
samples_df$n_peaks <- n_peaks_per_sample

# Per-sample centroid (mean rt_loc, mean cv_loc)
centroid_df <- peaks |>
  dplyr::group_by(sample_idx) |>
  dplyr::summarise(rt_centroid = mean(rt_loc),
                   cv_centroid = mean(cv_loc),
                   .groups = "drop")
samples_df <- dplyr::left_join(samples_df, centroid_df, by = "sample_idx")

# --- p-value collector for BH-FDR at the end ---------------------------
p_records <- list()
record_p <- function(label, p, test) {
  p_records[[length(p_records) + 1L]] <<-
    tibble::tibble(test = test, comparison = label, p_raw = p)
}

# Safe wilcox that returns NA when a group is empty / degenerate
safe_wilcox <- function(x, g) {
  ok <- is.finite(x) & !is.na(g)
  x <- x[ok]; g <- g[ok]
  if (length(unique(g)) < 2L) return(NA_real_)
  if (length(x) < 3L) return(NA_real_)
  tryCatch(stats::wilcox.test(x ~ g)$p.value, error = function(e) NA_real_)
}
```

# 1. Summary

```{r sec1-summary}
summary_tbl <- tibble::tibble(
  Field  = c("Cohort", "N samples", "N peaks detected",
             "Mean peaks / sample", "Median peaks / sample",
             "Noise mean (fitted)", "Noise SD (fitted)"),
  Value  = c(params$cohort_name,
             format(pp$n_samples),
             format(pp$n_peaks_detected),
             sprintf("%.1f", mean(samples_df$n_peaks)),
             sprintf("%.1f", stats::median(samples_df$n_peaks)),
             signif(pp$noise$mean, 3),
             signif(pp$noise$sd, 3))
)
knitr::kable(summary_tbl)

if (has_y) {
  class_tbl <- samples_df |>
    dplyr::count(class, name = "n") |>
    dplyr::mutate(pct = sprintf("%.1f%%", 100 * n / sum(n)))
  cat("\n\n**Class balance:**\n\n")
  knitr::kable(class_tbl)
}
```

# 2. Per-sample peak count

```{r sec2-peakcount}
p_count <- ggplot(samples_df, aes(x = n_peaks)) +
  geom_histogram(bins = 30, fill = "grey60", color = "white") +
  labs(x = "peaks per sample", y = "n samples",
       title = "Distribution of per-sample peak counts") +
  theme_minimal()
if (has_y) {
  p_count <- ggplot(samples_df, aes(x = n_peaks, fill = class,
                                    color = class)) +
    geom_density(alpha = 0.3) +
    labs(x = "peaks per sample", y = "density",
         title = "Peak count by class") +
    theme_minimal()
}
print(p_count)

if (has_y) {
  p_val <- safe_wilcox(samples_df$n_peaks, samples_df$class)
  record_p("peak count", p_val, "Wilcoxon rank-sum")
  cat("\n\nWilcoxon rank-sum on peak counts by class: p =",
      signif(p_val, 3), " (BH-adjusted values in Section 9)\n")
}
```

# 3. Peak location hotspots

Hotspot maps use the raw `(rt_loc, cv_loc)` coordinates (fractions of the padded image). Bin counts reflect where peaks land in image space; higher-count bins are cross-sample recurring locations.

```{r sec3-hotspots, fig.height = 5}
# Coarse grid used both for the hotspot map and the sample-level density
# features in Section 4. Chosen to give ~40 RT bins and ~15 CV bins.
GRID_RT <- 40L
GRID_CV <- 15L
rt_breaks <- seq(0, 1, length.out = GRID_RT + 1L)
cv_breaks <- seq(0, 1, length.out = GRID_CV + 1L)

peaks$rt_bin <- cut(peaks$rt_loc, rt_breaks,
                    include.lowest = TRUE, labels = FALSE)
peaks$cv_bin <- cut(peaks$cv_loc, cv_breaks,
                    include.lowest = TRUE, labels = FALSE)

p_hot_all <- ggplot(peaks, aes(x = cv_loc, y = rt_loc)) +
  geom_bin2d(bins = c(GRID_CV, GRID_RT)) +
  scale_fill_viridis_c(option = "magma") +
  labs(x = "CV (fraction of image)", y = "RT (fraction of image)",
       title = "All peaks (pooled across samples)") +
  theme_minimal()
print(p_hot_all)
```

```{r sec3-hotspots-by-class, eval = has_y, fig.height = 4}
peaks_c <- dplyr::left_join(peaks,
                            samples_df |> dplyr::select(sample_idx, class),
                            by = "sample_idx")

p_by_class <- ggplot(peaks_c, aes(x = cv_loc, y = rt_loc)) +
  geom_bin2d(bins = c(GRID_CV, GRID_RT)) +
  scale_fill_viridis_c(option = "magma") +
  facet_wrap(~ class) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Peaks by class") +
  theme_minimal()
print(p_by_class)
```

```{r sec3-diff-map, eval = has_y, fig.height = 4}
# Difference map: normalize each class to relative density, then subtract.
lv <- levels(samples_df$class)
if (length(lv) == 2L) {
  bin_counts <- peaks_c |>
    dplyr::count(class, rt_bin, cv_bin, name = "n") |>
    dplyr::group_by(class) |>
    dplyr::mutate(rel = n / sum(n)) |>
    dplyr::ungroup()

  wide <- tidyr::pivot_wider(bin_counts |> dplyr::select(-n),
                             names_from = class, values_from = rel,
                             values_fill = 0)
  wide$diff <- wide[[lv[2]]] - wide[[lv[1]]]

  p_diff <- ggplot(wide,
                   aes(x = cv_bin, y = rt_bin, fill = diff)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0,
                         name = sprintf("%s - %s", lv[2], lv[1])) +
    labs(x = "CV bin", y = "RT bin",
         title = sprintf("Class-difference density: %s - %s",
                         lv[2], lv[1])) +
    theme_minimal()
  print(p_diff)

  cat("\n\nRed cells indicate bins with higher relative peak density in ",
      lv[2], ". Blue cells indicate higher density in ", lv[1],
      ". Values are class-normalized relative frequencies, so differences ",
      "cancel to zero across the grid.")
} else {
  cat("Class difference map only shown for exactly 2 classes.")
}
```

# 4. Sample-level clustering and shift analysis

Two questions:

1. **Do samples cluster by class?** — PCA on a per-sample spatial-density feature vector (peaks binned into `r GRID_RT` × `r GRID_CV` cells), colored by class and (optionally) batch. Wilcoxon on PC1.
2. **Are some samples globally shifted?** — Each sample has a spatial centroid `(mean rt_loc, mean cv_loc)`. A sample's shift is its centroid minus the cohort-median centroid. Large shifts suggest instrument drift or run-specific offsets that global alignment can undo.

Global alignment applied here: `aligned_(rt,cv) = raw_(rt,cv) - (sample_shift)`. The before/after hotspot maps show whether this concentrates the cross-sample peak population.

```{r sec4-pca, fig.height = 5}
# Build sample x bin feature matrix
bin_id <- (peaks$cv_bin - 1L) * GRID_RT + peaks$rt_bin
feat <- matrix(0L, nrow = pp$n_samples, ncol = GRID_RT * GRID_CV)
tab <- table(peaks$sample_idx, bin_id)
feat[cbind(as.integer(rownames(tab)),
           as.integer(colnames(tab)))] <- 0L  # ensure init
for (r in seq_len(nrow(tab))) {
  si <- as.integer(rownames(tab)[r])
  cols <- as.integer(colnames(tab))
  feat[si, cols] <- as.integer(tab[r, ])
}
# Drop zero-variance columns and standardize
var_cols <- apply(feat, 2, var)
keep_cols <- which(var_cols > 0)
feat_use <- feat[, keep_cols, drop = FALSE]
if (ncol(feat_use) >= 2L && nrow(feat_use) >= 3L) {
  pc <- stats::prcomp(feat_use, scale. = TRUE)
  pca_df <- tibble::tibble(sample_idx = seq_len(pp$n_samples),
                           PC1 = pc$x[, 1],
                           PC2 = pc$x[, 2])
  pca_df <- dplyr::left_join(pca_df, samples_df, by = "sample_idx")
  var_expl <- pc$sdev^2 / sum(pc$sdev^2)

  aes_map <- aes(x = PC1, y = PC2, label = sample_name)
  if (has_y)     aes_map$colour <- rlang::sym("class")
  if (has_batch) aes_map$shape  <- rlang::sym("batch")

  p_pca <- ggplot(pca_df, aes_map) +
    geom_point(size = 2.5, alpha = 0.85) +
    labs(x = sprintf("PC1 (%.1f%%)", 100 * var_expl[1]),
         y = sprintf("PC2 (%.1f%%)", 100 * var_expl[2]),
         title = "Sample PCA on spatial-density features") +
    theme_minimal()
  print(p_pca)

  if (has_y) {
    p_val <- safe_wilcox(pca_df$PC1, pca_df$class)
    record_p("PC1 by class", p_val, "Wilcoxon rank-sum")
    cat("\n\nWilcoxon on PC1 by class: p =", signif(p_val, 3),
        " (BH-adjusted values in Section 9)\n")
  }
} else {
  cat("Not enough samples or nonzero bins to run PCA.")
}
```

```{r sec4-shifts, fig.height = 4}
cohort_med_rt <- stats::median(samples_df$rt_centroid, na.rm = TRUE)
cohort_med_cv <- stats::median(samples_df$cv_centroid, na.rm = TRUE)
samples_df$shift_rt <- samples_df$rt_centroid - cohort_med_rt
samples_df$shift_cv <- samples_df$cv_centroid - cohort_med_cv
samples_df$shift_mag <- sqrt(samples_df$shift_rt^2 + samples_df$shift_cv^2)

top_shift <- samples_df |>
  dplyr::arrange(dplyr::desc(shift_mag)) |>
  dplyr::select(sample_name, shift_rt, shift_cv, shift_mag) |>
  utils::head(15)
cat("**Top 15 samples by shift magnitude** (fractions of image):\n\n")
knitr::kable(top_shift, digits = 4)

shift_hist <- ggplot(samples_df, aes(x = shift_mag)) +
  geom_histogram(bins = 30, fill = "grey60", color = "white") +
  labs(x = "shift magnitude", y = "n samples",
       title = "Per-sample shift magnitude") +
  theme_minimal()
print(shift_hist)
```

```{r sec4-before-after, fig.height = 4}
# Apply per-sample shift to each peak
peaks_shift <- peaks |>
  dplyr::left_join(samples_df |>
                     dplyr::select(sample_idx, shift_rt, shift_cv),
                   by = "sample_idx") |>
  dplyr::mutate(rt_aligned = rt_loc - shift_rt,
                cv_aligned = cv_loc - shift_cv)

p_before <- ggplot(peaks_shift, aes(x = cv_loc, y = rt_loc)) +
  geom_bin2d(bins = c(GRID_CV, GRID_RT)) +
  scale_fill_viridis_c(option = "magma") +
  labs(x = "CV (fraction)", y = "RT (fraction)", title = "Before alignment") +
  theme_minimal()
p_after <- ggplot(peaks_shift, aes(x = cv_aligned, y = rt_aligned)) +
  geom_bin2d(bins = c(GRID_CV, GRID_RT)) +
  scale_fill_viridis_c(option = "magma") +
  labs(x = "CV (fraction)", y = "RT (fraction)", title = "After global alignment") +
  theme_minimal()
if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(p_before, p_after, ncol = 2)
} else {
  print(p_before); print(p_after)
}
```

<!--
Follow-up work (deferred): cluster samples on their (shift_rt, shift_cv)
vectors (e.g. k-means with several k) to detect unlabeled batch
structure. Track down whether samples with the largest |shift_mag|
share a processing date or run number that isn't currently captured in
the batch metadata.
-->

# 5. Peak widths

```{r sec5-widths, fig.height = 4}
widths_long <- tibble::tibble(
  sample_idx = peaks$sample_idx,
  sigma_rt = peaks$sigma_rt,
  sigma_cv = peaks$sigma_cv
) |>
  tidyr::pivot_longer(cols = c(sigma_rt, sigma_cv),
                      names_to = "axis", values_to = "sigma")

if (has_y) {
  widths_long <- dplyr::left_join(widths_long,
                                  samples_df |> dplyr::select(sample_idx, class),
                                  by = "sample_idx")
  p_w <- ggplot(widths_long, aes(x = sigma, color = class, fill = class)) +
    geom_density(alpha = 0.3) +
    facet_wrap(~ axis, scales = "free") +
    labs(x = "sigma (pixels)", y = "density",
         title = "Peak-width densities by axis and class") +
    theme_minimal()
  print(p_w)

  # Per-sample median widths for testing
  per_samp_w <- widths_long |>
    dplyr::group_by(sample_idx, axis, class) |>
    dplyr::summarise(sigma_med = stats::median(sigma), .groups = "drop")

  for (a in c("sigma_rt", "sigma_cv")) {
    sub <- per_samp_w[per_samp_w$axis == a, ]
    p_val <- safe_wilcox(sub$sigma_med, sub$class)
    record_p(paste0(a, " (per-sample median)"), p_val,
             "Wilcoxon rank-sum")
    cat("\n\nWilcoxon on per-sample median ", a,
        " by class: p =", signif(p_val, 3), "\n")
  }
} else {
  p_w <- ggplot(widths_long, aes(x = sigma)) +
    geom_density(fill = "grey60", alpha = 0.5) +
    facet_wrap(~ axis, scales = "free") +
    labs(x = "sigma (pixels)", y = "density",
         title = "Peak-width densities") +
    theme_minimal()
  print(p_w)
}
```

# 6. Peak intensities

```{r sec6-intensity, fig.height = 4}
peaks$log_intensity <- log(pmax(peaks$intensity, .Machine$double.eps))

if (has_y) {
  peaks_c <- dplyr::left_join(peaks,
                              samples_df |> dplyr::select(sample_idx, class),
                              by = "sample_idx")
  p_i <- ggplot(peaks_c, aes(x = log_intensity, color = class, fill = class)) +
    geom_density(alpha = 0.3) +
    labs(x = "log(intensity)", y = "density",
         title = "Log-intensity densities by class") +
    theme_minimal()
  print(p_i)

  per_samp_i <- peaks_c |>
    dplyr::group_by(sample_idx, class) |>
    dplyr::summarise(log_i_med = stats::median(log_intensity),
                     .groups = "drop")
  p_val <- safe_wilcox(per_samp_i$log_i_med, per_samp_i$class)
  record_p("per-sample median log(intensity)", p_val,
           "Wilcoxon rank-sum")
  cat("\n\nWilcoxon on per-sample median log(intensity) by class: p =",
      signif(p_val, 3), "\n")
} else {
  p_i <- ggplot(peaks, aes(x = log_intensity)) +
    geom_density(fill = "grey60", alpha = 0.5) +
    labs(x = "log(intensity)", y = "density",
         title = "Log-intensity density") +
    theme_minimal()
  print(p_i)
}
```

# 7. Noise

```{r sec7-noise}
knitr::kable(tibble::tibble(
  Field = c("Fitted noise mean",
            "Fitted noise SD",
            "Estimated `robust_eps` (0.01 * Q0.95)"),
  Value = c(
    signif(pp$noise$mean, 3),
    signif(pp$noise$sd, 3),
    if (!is.null(pp$intensity_raw) && length(pp$intensity_raw)) {
      signif(as.numeric(stats::quantile(pp$intensity_raw, 0.95)) * 0.01, 3)
    } else {
      NA
    }
  )
))
cat("\nCompare noise SD to a candidate hybrid eps of ",
    signif(3 * pp$noise$sd, 3),
    " (3 * noise SD). If this is much larger than the empirical eps, ",
    "consider tightening eps for future peak detection.\n")
```

# 8. Diagnostic flags

Three per-sample flags are computed to surface samples worth eyeballing:

- **`low_peak_count`** — the sample's total number of detected peaks is
  below Q1 - 1.5 * IQR relative to the cohort's peak-count distribution
  (the standard boxplot outlier rule, low side). Usually indicates
  under-saturated signal, a failed run, or a sample-prep issue that
  suppressed volatile release.

- **`high_peak_count`** — the sample's peak count is above Q3 + 1.5 * IQR
  (the same rule, high side). Usually indicates contamination,
  oversaturation, or a run whose atypical background got mis-classified
  as peaks.

- **`large_shift`** — the sample's peak-cloud *centroid* is farther from
  the cohort's median centroid than 95% of samples. Concretely: each
  sample has a spatial centroid `(mean RT of its peaks, mean CV of its
  peaks)`, and its shift is `(sample centroid) - (cohort median
  centroid)`. `shift_mag` is the Euclidean length of that vector. Large
  shifts usually mean the sample's peaks systematically land at
  different (RT, CV) coordinates than the rest of the cohort — for
  example instrument drift for that specific run, a batch offset, or an
  RT/CV calibration issue. Section 4's before / after alignment plots
  show whether subtracting the per-sample shift concentrates the cohort
  peak cloud.

Samples flagged *only* for `large_shift` (with normal peak count) are
the most interesting: they likely have real signal that just landed in
a shifted location — global alignment can rescue them. Samples flagged
for both `low_peak_count` and `large_shift` are more concerning and
warrant looking at the raw spectrum.

```{r sec8-flags}
q <- stats::quantile(samples_df$n_peaks, c(0.25, 0.75), na.rm = TRUE)
iqr <- q[2] - q[1]
low <- samples_df$n_peaks < q[1] - 1.5 * iqr
hi  <- samples_df$n_peaks > q[2] + 1.5 * iqr

shift_q <- stats::quantile(samples_df$shift_mag, 0.95, na.rm = TRUE)
big_shift <- samples_df$shift_mag > shift_q

flags <- samples_df |>
  dplyr::mutate(
    low_peak_count  = low,
    high_peak_count = hi,
    large_shift     = big_shift
  ) |>
  dplyr::filter(low_peak_count | high_peak_count | large_shift) |>
  dplyr::select(sample_name, n_peaks, shift_mag,
                low_peak_count, high_peak_count, large_shift)
if (nrow(flags) == 0L) {
  cat("No samples flagged on peak-count IQR or top-5% shift criteria.\n")
} else {
  knitr::kable(flags, digits = 4)
}
```

# 9. Multiple-testing summary

Benjamini-Hochberg-adjusted p-values across every class comparison reported above.

```{r sec9-fdr}
if (length(p_records) == 0L) {
  cat("No class comparisons ran (y not provided or single-class).\n")
} else {
  p_df <- dplyr::bind_rows(p_records)
  p_df$p_bh <- stats::p.adjust(p_df$p_raw, method = "BH")
  knitr::kable(p_df, digits = 4)
}
```

# Deferred (Phase 2)

Once alignment findings from Section 4 are interpreted, follow-up work will add:

- `build_peak_catalog()` to cluster shifted peaks across the cohort
- Anchor-peak analysis (peaks present in ≥ 90% of samples) with per-class comparison and volcano plot
- Sporadic-peak analysis (peaks present in < 30% of samples) — group-specific signals
- k-means on per-sample shift vectors to detect unlabeled batch structure
