---
title: "Cross-cohort catalog comparison: AKI, CHF, asthma, respiratory"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    fig_width: 8
    fig_height: 4.5
params:
  output_root: "output"
  eps_rt: 0.003
  eps_cv: 0.015
  min_support_frac: 0.10
  do_alignment: true             # if TRUE, run intra-cohort alignment
  do_cross_cohort_alignment: true # if TRUE, also do cross-cohort alignment (recommended)
  align_eps_rt_loose: 0.03
  align_eps_cv_loose: 0.08
  cross_align_eps_rt_loose: 0.04
  cross_align_eps_cv_loose: 0.10
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dpi = 96)
library(classydms)
library(dplyr)
library(tibble)
library(ggplot2)
library(tidyr)
library(purrr)
```

# Purpose

You have peak_params from four cohorts (AKI, CHF, asthma, respiratory) with disease/control labels.

> **Note on "control" labels in this study.** In each cohort the "control" group consists of subjects who presented with **similar symptoms as cases but were later found not to have the specific disease** being studied. Controls are therefore themselves symptomatic and produce disease-like volatile profiles from whatever else is going on; they are NOT healthy volunteers. This means:
>
> - Case-vs-control comparisons within a cohort are testing "disease X vs a mixed similarly-sick reference population", not "disease X vs healthy baseline"
> - The disease-specific signal is expected to be small — a modest perturbation on a large shared "sick person" volatile background
> - High case-control overlap and small prevalence deltas are expected, not evidence that the pipeline is broken
> - The `only_a` (disease-unique) compounds are the strongest biomarker candidates precisely because they distinguish the specific disease from other similar illnesses

This report:

1. Loads and builds a catalog per condition (9 total: 4 diseases + 4 controls + respiratory's "other")
2. Reports a summary table for each catalog
3. **Within-disease**: compares each disease to its matched control — biomarker candidates via prevalence delta and unique compounds
4. **Cross-condition controls**: compares the 4 control catalogs. **Important:** in this study "controls" are NOT healthy — they are subjects with similar symptoms who were later found NOT to have the specific disease being studied. So the controls are themselves a heterogeneous "similarly-symptomatic" population, and the comparison asks whether these similarly-sick reference populations look alike across cohorts, not whether the healthy-human volatile signature is stable.
5. **Cross-condition diseases**: compares diseases to each other — shared vs disease-specific compounds
6. **Universal catalog**: `merge_catalogs()` across everything, with per-source prevalence tracking

Renders can take a while (each catalog is ~500k pool peaks). Cache is enabled on the expensive chunks so re-renders are fast.

# 1. Load data and build catalogs

```{r load-build, cache = TRUE}
load_and_backfill <- function(cohort_name, root = params$output_root) {
  proc <- readRDS(file.path(root, cohort_name, "processed.rds"))
  pp   <- readRDS(file.path(root, cohort_name, "peak_params.rds"))
  # Backfill if missing sample_idx_raw
  if (is.null(pp$sample_idx_raw)) {
    pp <- backfill_sample_idx_raw(pp, sample_names = proc$ids)
  }
  list(proc = proc, pp = pp)
}

# Load
d_aki  <- load_and_backfill("aki")
d_chf  <- load_and_backfill("chf")
d_asth <- load_and_backfill("asthma")
d_resp <- load_and_backfill("respiratory")

# Subset by class and build catalogs
build_from <- function(pp) build_peak_catalog(
  pp,
  eps_rt = params$eps_rt,
  eps_cv = params$eps_cv,
  min_support_frac = params$min_support_frac
)

cat_aki           <- build_from(subset_peak_params(d_aki$pp,  which(d_aki$proc$y  == 2)))
cat_aki_control   <- build_from(subset_peak_params(d_aki$pp,  which(d_aki$proc$y  == 1)))
cat_chf           <- build_from(subset_peak_params(d_chf$pp,  which(d_chf$proc$y  == 2)))
cat_chf_control   <- build_from(subset_peak_params(d_chf$pp,  which(d_chf$proc$y  == 1)))
cat_asthma        <- build_from(subset_peak_params(d_asth$pp, which(d_asth$proc$y == 2)))
cat_asthma_control<- build_from(subset_peak_params(d_asth$pp, which(d_asth$proc$y == 1)))
cat_resp_flu      <- build_from(subset_peak_params(d_resp$pp, which(d_resp$proc$y == 2)))
cat_resp_other    <- build_from(subset_peak_params(d_resp$pp, which(d_resp$proc$y == 3)))
cat_resp_control  <- build_from(subset_peak_params(d_resp$pp, which(d_resp$proc$y == 1)))

all_cats <- list(
  aki             = cat_aki,             aki_control    = cat_aki_control,
  chf             = cat_chf,             chf_control    = cat_chf_control,
  asthma          = cat_asthma,          asthma_control = cat_asthma_control,
  resp_flu        = cat_resp_flu,        resp_other     = cat_resp_other,
  resp_control    = cat_resp_control
)
```

# 1b. Alignment (optional) and its effect on catalog compactness

Systematic per-sample drift in instrument-detected peak positions can fragment true underlying compounds into multiple catalog entries. We use `align_via_catalog()` to correct drift per cohort before catalog building, then rebuild catalogs on aligned data and compare against the un-aligned baseline.

**Method** (from `align_via_catalog`): build a preliminary catalog with loose eps, use each anchor compound's cluster centroid + member-peak offsets to derive per-sample shifts, iterate.

```{r alignment, cache = TRUE, eval = params$do_alignment}
align_one <- function(pp, name, tag = "") {
  message(sprintf("Aligning %s%s", name, tag))
  align_via_catalog(pp,
    eps_rt_loose = params$align_eps_rt_loose,
    eps_cv_loose = params$align_eps_cv_loose,
    min_support_frac = params$min_support_frac,
    iterate = TRUE, max_iter = 3L, verbose = FALSE)
}

# Align each cohort's full peak_params, then subset by class
d_aki$pp_aligned  <- align_one(d_aki$pp,  "aki")
d_chf$pp_aligned  <- align_one(d_chf$pp,  "chf")
d_asth$pp_aligned <- align_one(d_asth$pp, "asthma")
d_resp$pp_aligned <- align_one(d_resp$pp, "respiratory")

# Rebuild catalogs on aligned peak_params
cat_aki_aligned            <- build_from(subset_peak_params(d_aki$pp_aligned,  which(d_aki$proc$y  == 2)))
cat_aki_control_aligned    <- build_from(subset_peak_params(d_aki$pp_aligned,  which(d_aki$proc$y  == 1)))
cat_chf_aligned            <- build_from(subset_peak_params(d_chf$pp_aligned,  which(d_chf$proc$y  == 2)))
cat_chf_control_aligned    <- build_from(subset_peak_params(d_chf$pp_aligned,  which(d_chf$proc$y  == 1)))
cat_asthma_aligned         <- build_from(subset_peak_params(d_asth$pp_aligned, which(d_asth$proc$y == 2)))
cat_asthma_control_aligned <- build_from(subset_peak_params(d_asth$pp_aligned, which(d_asth$proc$y == 1)))
cat_resp_flu_aligned       <- build_from(subset_peak_params(d_resp$pp_aligned, which(d_resp$proc$y == 2)))
cat_resp_other_aligned     <- build_from(subset_peak_params(d_resp$pp_aligned, which(d_resp$proc$y == 3)))
cat_resp_control_aligned   <- build_from(subset_peak_params(d_resp$pp_aligned, which(d_resp$proc$y == 1)))

all_cats_aligned <- list(
  aki             = cat_aki_aligned,             aki_control    = cat_aki_control_aligned,
  chf             = cat_chf_aligned,             chf_control    = cat_chf_control_aligned,
  asthma          = cat_asthma_aligned,          asthma_control = cat_asthma_control_aligned,
  resp_flu        = cat_resp_flu_aligned,        resp_other     = cat_resp_other_aligned,
  resp_control    = cat_resp_control_aligned
)
```

## Per-sample shift magnitudes across cohorts

Distribution of the RT and CV shifts each sample received from alignment. Bimodal or long-tailed distributions suggest date/batch structure. A tight peak near 0 means alignment didn't detect meaningful drift.

```{r shift-histograms, eval = params$do_alignment, fig.height = 6}
shifts_df <- bind_rows(
  data.frame(cohort = "aki",         shift_rt = d_aki$pp_aligned$alignment$shifts_rt,
                                     shift_cv = d_aki$pp_aligned$alignment$shifts_cv),
  data.frame(cohort = "chf",         shift_rt = d_chf$pp_aligned$alignment$shifts_rt,
                                     shift_cv = d_chf$pp_aligned$alignment$shifts_cv),
  data.frame(cohort = "asthma",      shift_rt = d_asth$pp_aligned$alignment$shifts_rt,
                                     shift_cv = d_asth$pp_aligned$alignment$shifts_cv),
  data.frame(cohort = "respiratory", shift_rt = d_resp$pp_aligned$alignment$shifts_rt,
                                     shift_cv = d_resp$pp_aligned$alignment$shifts_cv)
)
shifts_long <- shifts_df %>%
  tidyr::pivot_longer(cols = c(shift_rt, shift_cv),
                       names_to = "axis", values_to = "shift")
ggplot(shifts_long, aes(x = shift)) +
  geom_histogram(bins = 40, fill = "#B2182B") +
  facet_grid(cohort ~ axis, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "shift (fraction of image)", y = "n samples",
       title = "Per-sample alignment shifts by cohort and axis") +
  theme_minimal()
```

## Catalog compactness: unaligned vs aligned

Six key metrics per cohort. Alignment is helping when: n_compounds decreases, n_singletons decreases, median_prevalence increases, n_anchor_90pct increases, median_obs_per_sample stays near 1.

```{r compactness-compare, eval = params$do_alignment}
compact_row <- function(cat, name, method) {
  c <- catalog_compactness(cat)
  tibble(
    catalog = name, method = method,
    n_compounds = c$n_compounds,
    n_singletons = nrow(cat$singletons),
    median_obs_per_sample = round(c$median_obs_per_sample, 3),
    median_loc_sd_rt_px = round(c$median_loc_sd_rt_px, 2),
    median_prevalence = round(median(cat$compounds$prevalence), 3),
    n_anchor_90pct = sum(cat$compounds$prevalence >= 0.9)
  )
}
compactness_before <- purrr::imap_dfr(all_cats,          ~compact_row(.x, .y, "unaligned"))
compactness_after  <- purrr::imap_dfr(all_cats_aligned,  ~compact_row(.x, .y, "aligned"))
compactness_compare <- bind_rows(compactness_before, compactness_after) %>%
  arrange(catalog, method)
knitr::kable(compactness_compare)
```

## Delta table: is alignment helping per cohort?

```{r delta-table, eval = params$do_alignment}
delta_df <- inner_join(
  compactness_before %>% rename_with(~paste0(., "_before"), -c(catalog, method)) %>%
    select(-method),
  compactness_after %>% rename_with(~paste0(., "_after"),  -c(catalog, method)) %>%
    select(-method),
  by = "catalog"
) %>%
  mutate(
    delta_n_compounds     = n_compounds_after - n_compounds_before,
    delta_n_singletons    = n_singletons_after - n_singletons_before,
    delta_median_prev     = median_prevalence_after - median_prevalence_before,
    delta_n_anchor_90pct  = n_anchor_90pct_after - n_anchor_90pct_before
  ) %>%
  select(catalog, delta_n_compounds, delta_n_singletons,
         delta_median_prev, delta_n_anchor_90pct)
knitr::kable(delta_df,
             caption = "Change from alignment: negative delta on compounds/singletons = better; positive delta on prevalence/anchor = better")
```

**Interpretation guide:**

- **All deltas in expected direction across cohorts** → alignment is helping generally; use aligned catalogs going forward
- **Some cohorts help, some don't** → drift is cohort-specific; consider aligning only the cohorts that benefit
- **All deltas near zero** → cohorts have minimal drift; alignment is safe to skip
- **Deltas in wrong direction** → alignment is over-correcting; loosen `eps_rt_loose` or increase `min_samples_per_anchor`

## Which catalog to use downstream

```{r select-catalogs, eval = params$do_alignment}
# Intra-cohort aligned catalogs — may be overridden by cross-cohort
# alignment in the next section.
all_cats <- all_cats_aligned
cat_aki             <- cat_aki_aligned
cat_aki_control     <- cat_aki_control_aligned
cat_chf             <- cat_chf_aligned
cat_chf_control     <- cat_chf_control_aligned
cat_asthma          <- cat_asthma_aligned
cat_asthma_control  <- cat_asthma_control_aligned
cat_resp_flu        <- cat_resp_flu_aligned
cat_resp_other      <- cat_resp_other_aligned
cat_resp_control    <- cat_resp_control_aligned
```

# 1c. Cross-cohort alignment

Intra-cohort alignment (Section 1b) corrects drift **within** each cohort — samples run close together in time drifting together. But if different cohorts were run months apart on the same instrument, or on different service periods, the cohorts themselves can sit at systematically-different `(rt_loc, cv_loc)` positions from each other. Intra-cohort alignment can't detect that.

This section runs `align_across_cohorts()` — merges all cohorts' peak_params into one super-cohort, aligns via cluster-membership, then splits back. Each cohort's samples get shifted to a shared reference so their compound positions line up.

**Diagnostic value**: if intra-cohort shifts were tiny but the universal catalog still shows vertical stripes with many "1-source-only" compounds, cross-cohort drift is the likely culprit and this section will fix it.

```{r cross-alignment, cache = TRUE, eval = params$do_cross_cohort_alignment}
# Use intra-cohort aligned peak_params (if available) as input
input_pp <- list(
  aki         = if (params$do_alignment) d_aki$pp_aligned  else d_aki$pp,
  chf         = if (params$do_alignment) d_chf$pp_aligned  else d_chf$pp,
  asthma      = if (params$do_alignment) d_asth$pp_aligned else d_asth$pp,
  respiratory = if (params$do_alignment) d_resp$pp_aligned else d_resp$pp
)

cross_aligned <- align_across_cohorts(input_pp,
  eps_rt_loose = params$cross_align_eps_rt_loose,
  eps_cv_loose = params$cross_align_eps_cv_loose,
  min_support_frac = params$min_support_frac,
  iterate = TRUE, max_iter = 3L, verbose = TRUE)

# Rebuild catalogs on cross-aligned peak_params
cat_aki_cross            <- build_from(subset_peak_params(cross_aligned$aki,         which(d_aki$proc$y  == 2)))
cat_aki_control_cross    <- build_from(subset_peak_params(cross_aligned$aki,         which(d_aki$proc$y  == 1)))
cat_chf_cross            <- build_from(subset_peak_params(cross_aligned$chf,         which(d_chf$proc$y  == 2)))
cat_chf_control_cross    <- build_from(subset_peak_params(cross_aligned$chf,         which(d_chf$proc$y  == 1)))
cat_asthma_cross         <- build_from(subset_peak_params(cross_aligned$asthma,      which(d_asth$proc$y == 2)))
cat_asthma_control_cross <- build_from(subset_peak_params(cross_aligned$asthma,      which(d_asth$proc$y == 1)))
cat_resp_flu_cross       <- build_from(subset_peak_params(cross_aligned$respiratory, which(d_resp$proc$y == 2)))
cat_resp_other_cross     <- build_from(subset_peak_params(cross_aligned$respiratory, which(d_resp$proc$y == 3)))
cat_resp_control_cross   <- build_from(subset_peak_params(cross_aligned$respiratory, which(d_resp$proc$y == 1)))

all_cats_cross <- list(
  aki             = cat_aki_cross,             aki_control    = cat_aki_control_cross,
  chf             = cat_chf_cross,             chf_control    = cat_chf_control_cross,
  asthma          = cat_asthma_cross,          asthma_control = cat_asthma_control_cross,
  resp_flu        = cat_resp_flu_cross,        resp_other     = cat_resp_other_cross,
  resp_control    = cat_resp_control_cross
)
```

## Cross-cohort shifts by cohort

Distribution of shifts each sample received from the cross-cohort alignment. If intra-cohort alignment already happened, these are ADDITIONAL shifts on top. Look for cohort-level offsets — if all samples in one cohort shifted by ~+0.005 and all in another by ~-0.005, the cross-cohort drift signal is real.

```{r cross-shift-histograms, eval = params$do_cross_cohort_alignment, fig.height = 6}
cross_shifts_df <- bind_rows(
  data.frame(cohort = "aki",         shift_rt = cross_aligned$aki$alignment$shifts_rt,
                                     shift_cv = cross_aligned$aki$alignment$shifts_cv),
  data.frame(cohort = "chf",         shift_rt = cross_aligned$chf$alignment$shifts_rt,
                                     shift_cv = cross_aligned$chf$alignment$shifts_cv),
  data.frame(cohort = "asthma",      shift_rt = cross_aligned$asthma$alignment$shifts_rt,
                                     shift_cv = cross_aligned$asthma$alignment$shifts_cv),
  data.frame(cohort = "respiratory", shift_rt = cross_aligned$respiratory$alignment$shifts_rt,
                                     shift_cv = cross_aligned$respiratory$alignment$shifts_cv)
)
cross_shifts_long <- cross_shifts_df %>%
  tidyr::pivot_longer(cols = c(shift_rt, shift_cv),
                       names_to = "axis", values_to = "shift")
ggplot(cross_shifts_long, aes(x = shift, fill = cohort)) +
  geom_histogram(bins = 40, position = "identity", alpha = 0.5) +
  facet_wrap(~ axis, scales = "free", ncol = 2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "shift (fraction of image)", y = "n samples",
       title = "Cross-cohort shifts (with cohort overlay)",
       subtitle = "If cohort colors separate along the x-axis, cross-cohort drift is real") +
  theme_minimal()
```

## Median shift by cohort — quick summary

```{r cross-cohort-shift-summary, eval = params$do_cross_cohort_alignment}
cross_shifts_df %>%
  group_by(cohort) %>%
  summarise(
    median_shift_rt = round(median(shift_rt), 5),
    median_shift_cv = round(median(shift_cv), 5),
    max_abs_rt      = round(max(abs(shift_rt)), 5),
    max_abs_cv      = round(max(abs(shift_cv)), 5),
    n_samples       = n(),
    .groups = "drop"
  ) %>% knitr::kable(caption = "Cross-cohort shift magnitudes per cohort")
```

**How to read this table:**

- **Median shifts near 0** with **max_abs small** → cohorts are already well-aligned to each other; cross-cohort alignment isn't needed.
- **Median shifts differ between cohorts** (e.g., one cohort's median is +0.005, another's is -0.005) → real cross-cohort drift; alignment is fixing it.
- **Large max_abs but median near 0** → some individual samples had large shifts (potential outliers) but most cohort samples were fine.

## Compactness: unaligned vs intra-cohort aligned vs cross-cohort aligned

```{r cross-compactness-compare, eval = params$do_cross_cohort_alignment}
compact_row_cross <- function(cat, name, method) {
  c <- catalog_compactness(cat)
  tibble(
    catalog = name, method = method,
    n_compounds = c$n_compounds,
    n_singletons = nrow(cat$singletons),
    median_prevalence = round(median(cat$compounds$prevalence), 3),
    n_anchor_90pct = sum(cat$compounds$prevalence >= 0.9)
  )
}
compactness_cross <- purrr::imap_dfr(all_cats_cross, ~compact_row_cross(.x, .y, "cross_aligned"))

three_way <- bind_rows(
  purrr::imap_dfr(all_cats,         ~compact_row_cross(.x, .y, "aligned_intra")),
  compactness_cross
) %>% arrange(catalog, method)
knitr::kable(three_way,
             caption = "Compactness after intra-cohort vs cross-cohort alignment")
```

## Universal catalog spatial check — is the vertical-stripe pattern reduced?

The direct test of whether cross-cohort alignment worked. Build **two** universal catalogs — one from un-cross-aligned peak_params, one from cross-aligned — using `build_universal_catalog()` which pools ALL cohorts' samples and runs clustering once. This is the correct comparison: if drift was fragmenting compounds across cohorts, alignment should produce fewer, tighter clusters.

```{r universal-cross-preview, cache = TRUE, eval = params$do_cross_cohort_alignment}
# Un-cross-aligned universal (built from cohort-aligned OR original pp)
pp_source_before <- list(
  aki = subset_peak_params(input_pp$aki,         seq_len(input_pp$aki$n_samples)),
  aki_ctrl = NULL,                                # placeholder, dropped below
  chf = subset_peak_params(input_pp$chf,         seq_len(input_pp$chf$n_samples)),
  chf_ctrl = NULL,
  asthma = subset_peak_params(input_pp$asthma,   seq_len(input_pp$asthma$n_samples)),
  asthma_ctrl = NULL,
  resp = subset_peak_params(input_pp$respiratory, seq_len(input_pp$respiratory$n_samples)),
  resp_ctrl = NULL
)
# Actually simpler: just pass the 4 whole-cohort pps (they already contain all classes)
preview_pp_before <- list(
  aki         = input_pp$aki,
  chf         = input_pp$chf,
  asthma      = input_pp$asthma,
  respiratory = input_pp$respiratory
)
preview_pp_after <- list(
  aki         = cross_aligned$aki,
  chf         = cross_aligned$chf,
  asthma      = cross_aligned$asthma,
  respiratory = cross_aligned$respiratory
)

universal_before_cross <- build_universal_catalog(preview_pp_before,
  eps_rt = params$eps_rt, eps_cv = params$eps_cv,
  min_support_frac = params$min_support_frac)
universal_after_cross  <- build_universal_catalog(preview_pp_after,
  eps_rt = params$eps_rt, eps_cv = params$eps_cv,
  min_support_frac = params$min_support_frac)

sprintf("Universal (before cross-align): %d compounds",
        nrow(universal_before_cross$compounds))
sprintf("Universal (after  cross-align): %d compounds",
        nrow(universal_after_cross$compounds))
```

```{r universal-cross-spatial, eval = params$do_cross_cohort_alignment, fig.height = 6, fig.width = 10}
render_universal_map <- function(uni, title_suffix) {
  n_cohorts <- length(uni$parameters$source_names)
  per_src <- do.call(rbind, uni$compounds$per_source_prevalence) |>
    as.data.frame()
  colnames(per_src) <- uni$parameters$source_names
  n_sources <- rowSums(per_src > 0)
  map_df <- tibble(
    rt_loc = uni$compounds$rt_loc,
    cv_loc = uni$compounds$cv_loc,
    n_sources = n_sources,
    prevalence = uni$compounds$prevalence
  ) |> mutate(coverage_bin = case_when(
    n_sources == n_cohorts ~ sprintf("%d sources (universal)", n_cohorts),
    n_sources >= max(2, n_cohorts - 1) ~ sprintf(">= %d sources",
                                                    max(2, n_cohorts - 1)),
    n_sources >= 2 ~ "2 sources",
    TRUE ~ "1 source only"
  ))
  map_df$coverage_bin <- factor(map_df$coverage_bin,
    levels = c("1 source only", "2 sources",
                sprintf(">= %d sources", max(2, n_cohorts - 1)),
                sprintf("%d sources (universal)", n_cohorts)))
  ggplot(map_df, aes(x = cv_loc, y = rt_loc, color = coverage_bin,
                     size = prevalence)) +
    geom_point(alpha = 0.55) +
    scale_color_manual(values = c(
      "1 source only" = "#B2182B", "2 sources" = "#EF8A62"),
      limits = levels(map_df$coverage_bin),
      guide = guide_legend(override.aes = list(size = 3))) +
    scale_size_continuous(range = c(0.3, 3), guide = "none") +
    scale_y_reverse() +
    labs(x = "CV (fraction)", y = "RT (fraction)",
         title = paste("Universal catalog —", title_suffix),
         subtitle = sprintf("%d compounds", nrow(uni$compounds)),
         color = "source coverage") +
    theme_minimal()
}
gridExtra::grid.arrange(
  render_universal_map(universal_before_cross, "before cross-align"),
  render_universal_map(universal_after_cross,  "after cross-align"),
  ncol = 2
)
```

**How to read this side-by-side:**

- **Fewer red points (1-source-only) after alignment** → compounds that were previously split across cohorts merged into cross-cohort ones. Good.
- **More dark-blue points (all-source) after alignment** → previously-split compounds now count as universal. Good.
- **Vertical-stripe pattern reduced after alignment** → the RT drift between cohorts is being corrected.
- **Same or worse pattern** → drift isn't the main story; other factors (real per-cohort chemistry differences, or clustering artifacts) may be at play.

## Use cross-aligned catalogs for downstream analysis?

```{r select-cross-catalogs, eval = params$do_cross_cohort_alignment}
# Swap in the cross-aligned catalogs. Comment these lines out to
# revert to intra-cohort aligned if the cross-alignment doesn't help.
all_cats <- all_cats_cross
cat_aki             <- cat_aki_cross
cat_aki_control     <- cat_aki_control_cross
cat_chf             <- cat_chf_cross
cat_chf_control     <- cat_chf_control_cross
cat_asthma          <- cat_asthma_cross
cat_asthma_control  <- cat_asthma_control_cross
cat_resp_flu        <- cat_resp_flu_cross
cat_resp_other      <- cat_resp_other_cross
cat_resp_control    <- cat_resp_control_cross
```

# 2. Summary table across all catalogs

```{r summary-table}
summary_row <- function(cat, name) {
  s <- catalog_summary(cat, print = FALSE)
  tibble(
    catalog     = name,
    n_samples   = s$n_samples,
    n_pool_peaks= s$n_pool_peaks,
    n_compounds = s$n_compounds,
    n_singletons= s$n_singletons,
    frac_singleton = round(s$fraction_singleton, 3),
    median_prev = round(median(cat$compounds$prevalence), 3),
    mean_prev   = round(mean(cat$compounds$prevalence), 3),
    n_anchor_90pct = sum(cat$compounds$prevalence >= 0.90)
  )
}
summary_df <- purrr::imap_dfr(all_cats, ~summary_row(.x, .y))
knitr::kable(summary_df)
```

**How to read this table:**

- Similar `n_pool_peaks / n_samples` across cohorts → similar peak detection settings and instrument behavior (good baseline)
- Very different `n_compounds` → could reflect real differences OR that eps needs tuning for one cohort
- `frac_singleton` — high (>20%) suggests eps is too tight or many compounds are truly rare; low (<10%) suggests eps is reasonable

# 3. Within-disease: disease vs control per cohort

For each of the 4 diseases, compare disease catalog to its matched control catalog. Focuses biomarker discovery: which compounds appear in disease but not control (or at different prevalence)?

```{r within-disease-compare}
within_pairs <- list(
  aki    = list(disease = cat_aki,       control = cat_aki_control),
  chf    = list(disease = cat_chf,       control = cat_chf_control),
  asthma = list(disease = cat_asthma,    control = cat_asthma_control),
  resp_flu   = list(disease = cat_resp_flu,   control = cat_resp_control),
  resp_other = list(disease = cat_resp_other, control = cat_resp_control)
)

within_comparisons <- purrr::imap(within_pairs, function(pair, name) {
  compare_catalogs(pair$disease, pair$control,
                    name_a = "disease", name_b = "control")
})

within_summary <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  cbind(comparison = name, comp$summary)
})
knitr::kable(within_summary)
```

## Prevalence-delta distribution per disease

```{r within-delta, fig.height = 6}
delta_df <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  tibble(comparison = name, prevalence_delta = comp$shared$prevalence_delta)
})
ggplot(delta_df, aes(x = prevalence_delta)) +
  geom_histogram(bins = 40, fill = "#B2182B") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ comparison, ncol = 2, scales = "free_y") +
  labs(x = "prevalence delta (control − disease)",
       y = "n shared compounds",
       title = "Prevalence delta for shared compounds within each disease pair") +
  theme_minimal()
```

**Reading**: bimodal or long-tailed distributions on either side indicate compounds systematically expressed differently between disease and control — the biomarker candidates.

## Top biomarker candidates per disease

For each disease, the shared compounds with the largest `|prevalence_delta|` are candidates: same (RT, CV) coordinates, very different prevalence between the two groups.

```{r top-biomarkers}
top_biom <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  comp$shared |>
    dplyr::arrange(dplyr::desc(abs(prevalence_delta))) |>
    head(10) |>
    mutate(comparison = name) |>
    select(comparison, rt_loc_disease, cv_loc_disease,
           prevalence_disease, prevalence_control, prevalence_delta,
           spatial_distance)
})
knitr::kable(top_biom, digits = 3)
```

Negative `prevalence_delta` = compound MORE prevalent in disease (candidate biomarker of the disease itself). Positive = LESS prevalent in disease (candidate for "protective" or "control-associated" compounds).

## Only-in-disease and only-in-control compounds

```{r unique-counts}
unique_counts <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  tibble(
    comparison = name,
    only_disease = nrow(comp$only_a),
    only_control = nrow(comp$only_b),
    disease_top_prevalence = max(comp$only_a$prevalence, na.rm = TRUE),
    control_top_prevalence = max(comp$only_b$prevalence, na.rm = TRUE)
  )
})
knitr::kable(unique_counts)
```

Compounds only in disease that also have high prevalence within disease are the strongest biomarker candidates — same physical position doesn't exist in controls at all, AND it's consistently expressed in disease samples.

## Spatial view of unique compounds per disease

Each panel shows one disease pair. **Grey dots** = compounds shared between disease and control (background context). **Red dots** = compounds only in disease. **Blue dots** = compounds only in control. Point size = compound prevalence within its catalog. Cluster patterns of the colored dots often reveal spatial regions of the (RT, CV) space that differentiate the classes.

```{r unique-spatial, fig.height = 9, fig.width = 10}
unique_spatial_df <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  bind_rows(
    if (nrow(comp$shared) > 0)
      tibble(comparison = name, group = "shared",
             rt_loc = comp$shared$rt_loc_disease,
             cv_loc = comp$shared$cv_loc_disease,
             prevalence = (comp$shared$prevalence_disease +
                             comp$shared$prevalence_control) / 2)
    else tibble(comparison = character(0), group = character(0),
                rt_loc = numeric(0), cv_loc = numeric(0),
                prevalence = numeric(0)),
    if (nrow(comp$only_a) > 0)
      tibble(comparison = name, group = "only disease",
             rt_loc = comp$only_a$rt_loc,
             cv_loc = comp$only_a$cv_loc,
             prevalence = comp$only_a$prevalence)
    else tibble(comparison = character(0), group = character(0),
                rt_loc = numeric(0), cv_loc = numeric(0),
                prevalence = numeric(0)),
    if (nrow(comp$only_b) > 0)
      tibble(comparison = name, group = "only control",
             rt_loc = comp$only_b$rt_loc,
             cv_loc = comp$only_b$cv_loc,
             prevalence = comp$only_b$prevalence)
    else tibble(comparison = character(0), group = character(0),
                rt_loc = numeric(0), cv_loc = numeric(0),
                prevalence = numeric(0))
  )
})
unique_spatial_df$group <- factor(unique_spatial_df$group,
  levels = c("shared", "only disease", "only control"))

ggplot(unique_spatial_df,
       aes(x = cv_loc, y = rt_loc,
           color = group, size = prevalence, alpha = group)) +
  geom_point() +
  scale_color_manual(values = c("shared"       = "grey60",
                                  "only disease" = "#B2182B",
                                  "only control" = "#2166AC")) +
  scale_alpha_manual(values = c("shared" = 0.2, "only disease" = 0.85,
                                  "only control" = 0.85), guide = "none") +
  scale_size_continuous(range = c(0.5, 3.5)) +
  scale_y_reverse() +
  facet_wrap(~ comparison, ncol = 2) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Spatial distribution of unique vs shared compounds",
       color = "compound group") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Top only-in-disease compounds table

Highest-prevalence disease-only compounds per cohort — the concrete biomarker candidates from the unique-compound analysis:

```{r top-only-disease}
top_only_disease <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  if (nrow(comp$only_a) == 0) return(NULL)
  comp$only_a |>
    arrange(desc(prevalence)) |>
    head(10) |>
    mutate(comparison = name) |>
    select(comparison, compound_id, rt_loc, cv_loc,
           prevalence, n_observations, n_samples)
})
knitr::kable(top_only_disease, digits = 3)
```

# 4. Cross-condition controls: how consistent are the "study-defined control" volatile signatures?

**Note on the "control" label:** in this study, controls are NOT healthy volunteers. They are subjects who presented with similar symptoms as the case group but were later found not to have the specific disease of interest. Controls therefore have their own (heterogeneous) underlying illnesses producing volatiles — the comparison here asks whether these similarly-symptomatic reference populations look alike across the 4 studies, not whether a healthy baseline reproduces.

If the 4 control catalogs share most compounds despite being drawn from different studies of different diseases, that tells us the "similarly-symptomatic ill" population has a recognizable common volatile profile — which is useful for interpreting biomarker signal in the case-vs-control comparisons of Section 3, because it means the "control" background is fairly stable across studies. Big disagreements between controls could reflect either genuine differences in the underlying illness mix per study, or batch/site/protocol differences.

```{r control-vs-control}
control_cats <- list(
  aki_control    = cat_aki_control,
  chf_control    = cat_chf_control,
  asthma_control = cat_asthma_control,
  resp_control   = cat_resp_control
)

# All pairwise combinations of controls
control_pairs <- t(combn(names(control_cats), 2))
control_comparisons <- lapply(seq_len(nrow(control_pairs)), function(i) {
  a <- control_pairs[i, 1]; b <- control_pairs[i, 2]
  cmp <- compare_catalogs(control_cats[[a]], control_cats[[b]],
                           name_a = a, name_b = b)
  list(pair = paste(a, "vs", b), summary = cmp$summary,
       shared = cmp$shared)
})
ctrl_summary <- do.call(rbind, lapply(control_comparisons, function(x)
  cbind(pair = x$pair, x$summary)))
knitr::kable(ctrl_summary)
```

**High `frac_shared_a` and `frac_shared_b` on all pairs (say > 0.7)** → controls are consistent; the "similarly-symptomatic-but-disease-negative" reference population has a reproducible volatile profile across studies. Low fractions suggest cohort-specific artifacts (batch effect, different equipment, different sample handling) OR genuine differences in the underlying illnesses in each study's control arm — worth investigating either way before trusting cross-cohort analysis.

## Prevalence-delta on controls

Prevalence-delta for shared compounds between controls SHOULD be small if the study-defined control populations are drawn from similar underlying-illness distributions. Large deltas indicate cohort-level differences in the reference-population volatile profile — could be batch effects OR real differences in the case-mix in each study's control arm.

```{r control-delta-summary}
ctrl_delta <- lapply(control_comparisons, function(x)
  tibble(pair = x$pair, prevalence_delta = x$shared$prevalence_delta))
ctrl_delta_df <- do.call(rbind, ctrl_delta)
ctrl_delta_df |>
  group_by(pair) |>
  summarise(
    median_abs_delta = median(abs(prevalence_delta)),
    q75_abs_delta    = quantile(abs(prevalence_delta), 0.75),
    max_abs_delta    = max(abs(prevalence_delta)),
    n_delta_gt_0.20  = sum(abs(prevalence_delta) > 0.20)
  ) |>
  knitr::kable(digits = 3)
```

# 5. Cross-condition diseases: shared vs unique

Same analysis but comparing diseases to each other. Compounds shared across multiple diseases at similar prevalence are "generic pathology" candidates; compounds unique to one disease are disease-specific candidates.

```{r disease-vs-disease}
disease_cats <- list(
  aki      = cat_aki,
  chf      = cat_chf,
  asthma   = cat_asthma,
  resp_flu = cat_resp_flu,
  resp_other = cat_resp_other
)

disease_pairs <- t(combn(names(disease_cats), 2))
disease_comparisons <- lapply(seq_len(nrow(disease_pairs)), function(i) {
  a <- disease_pairs[i, 1]; b <- disease_pairs[i, 2]
  cmp <- compare_catalogs(disease_cats[[a]], disease_cats[[b]],
                           name_a = a, name_b = b)
  list(pair = paste(a, "vs", b), summary = cmp$summary,
       shared = cmp$shared)
})
dis_summary <- do.call(rbind, lapply(disease_comparisons, function(x)
  cbind(pair = x$pair, x$summary)))
knitr::kable(dis_summary)
```

## Which disease pairs have the most compound overlap?

```{r disease-overlap-heatmap, fig.height = 4}
overlap_mat <- matrix(NA_real_, nrow = length(disease_cats),
                       ncol = length(disease_cats),
                       dimnames = list(names(disease_cats),
                                        names(disease_cats)))
diag(overlap_mat) <- 1
for (i in seq_len(nrow(disease_pairs))) {
  a <- disease_pairs[i, 1]; b <- disease_pairs[i, 2]
  frac_a <- disease_comparisons[[i]]$summary$frac_shared_a
  frac_b <- disease_comparisons[[i]]$summary$frac_shared_b
  overlap_mat[a, b] <- frac_a
  overlap_mat[b, a] <- frac_b
}
overlap_df <- as_tibble(overlap_mat, rownames = "row") |>
  pivot_longer(-row, names_to = "col", values_to = "frac_shared")

ggplot(overlap_df, aes(x = col, y = row, fill = frac_shared)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", frac_shared)), size = 3) +
  scale_fill_gradient(low = "#F7FBFF", high = "#08306B",
                       name = "frac shared") +
  labs(x = "col catalog (denominator)", y = "row catalog",
       title = "Fraction of row-catalog compounds found in col-catalog") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Read the matrix: element `[i, j]` = fraction of compounds in disease `i` that also appear in disease `j`. Note asymmetry — smaller catalogs have higher fractions.

# 5b. Catalog spatial heatmaps

Faceted (RT × CV) hexbin density of compounds per catalog. Where the catalog concentrates its compounds spatially — and how much the various cohorts overlap in coverage — is easier to see side by side than one catalog at a time.

## Density heatmap for each catalog

```{r catalog-heatmap, fig.height = 10, fig.width = 10}
all_compounds_df <- purrr::imap_dfr(all_cats, function(cat, name) {
  tibble(
    catalog    = name,
    rt_loc     = cat$compounds$rt_loc,
    cv_loc     = cat$compounds$cv_loc,
    prevalence = cat$compounds$prevalence
  )
})
# Consistent facet order: diseases first, controls second
all_compounds_df$catalog <- factor(all_compounds_df$catalog,
  levels = c("aki", "chf", "asthma", "resp_flu", "resp_other",
              "aki_control", "chf_control", "asthma_control",
              "resp_control"))

ggplot(all_compounds_df, aes(x = cv_loc, y = rt_loc)) +
  geom_hex(bins = c(20, 40)) +
  scale_fill_gradient(low = "#F7FBFF", high = "#08306B",
                       name = "n compounds") +
  scale_y_reverse() +
  facet_wrap(~ catalog, ncol = 3) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Compound density heatmap per catalog") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

**Reading:** hotspots that appear in every panel are conserved compound regions (universal volatile signature). Panels with different hotspot patterns from the rest are potential batch effects or genuine cohort-specific chemistry.

## Prevalence-weighted spatial map per catalog

Same layout but each compound sized by its prevalence — highlights the anchor compounds visually.

```{r catalog-prevalence-map, fig.height = 10, fig.width = 10}
ggplot(all_compounds_df, aes(x = cv_loc, y = rt_loc,
                              size = prevalence, color = prevalence)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "#F7FBFF", high = "#B2182B",
                        name = "prevalence") +
  scale_size_continuous(range = c(0.3, 2.5), guide = "none") +
  scale_y_reverse() +
  facet_wrap(~ catalog, ncol = 3) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Compound prevalence per catalog") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Bright red points (high prevalence, in most samples) that align across panels are the anchor compounds present in every cohort.

## Disease-only overlay across catalogs

Overlay compounds that are unique to each disease (from within-disease comparisons) on a single plot — reveals whether disease-only compounds cluster in different regions per disease.

```{r disease-only-overlay, fig.height = 6, fig.width = 8}
disease_only_all <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  if (nrow(comp$only_a) == 0) return(NULL)
  tibble(disease = name,
          rt_loc = comp$only_a$rt_loc,
          cv_loc = comp$only_a$cv_loc,
          prevalence = comp$only_a$prevalence)
})

ggplot(disease_only_all,
       aes(x = cv_loc, y = rt_loc, color = disease, size = prevalence)) +
  geom_point(alpha = 0.7) +
  scale_y_reverse() +
  scale_size_continuous(range = c(0.5, 3.5)) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "All disease-only compounds overlaid",
       color = "disease") +
  theme_minimal()
```

Regions of the plot where colors mix are areas multiple diseases have unique compounds — could reflect "generic pathology" positions. Regions dominated by one color are disease-specific spatial signatures.

# 6. Universal catalog across everything

Uses `build_universal_catalog()` to pool ALL samples from all cohorts and cluster once, rather than merging pre-built catalogs. This is the correct approach because it lets systematic per-cohort offsets get absorbed into cluster centroids instead of producing duplicate entries.

The nine "sources" here are the 9 disease/control groups (5 disease groups + 4 controls). Each compound's `per_source_prevalence` is the fraction of that source's samples in which the compound appeared.

```{r universal, cache = TRUE}
# Build per-condition pp objects to serve as sources for the universal
uni_pp_sources <- list(
  aki            = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$aki
                                      else if (params$do_alignment)         d_aki$pp_aligned
                                      else                                    d_aki$pp,
                                      which(d_aki$proc$y == 2)),
  aki_ctrl       = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$aki
                                      else if (params$do_alignment)         d_aki$pp_aligned
                                      else                                    d_aki$pp,
                                      which(d_aki$proc$y == 1)),
  chf            = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$chf
                                      else if (params$do_alignment)         d_chf$pp_aligned
                                      else                                    d_chf$pp,
                                      which(d_chf$proc$y == 2)),
  chf_ctrl       = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$chf
                                      else if (params$do_alignment)         d_chf$pp_aligned
                                      else                                    d_chf$pp,
                                      which(d_chf$proc$y == 1)),
  asthma         = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$asthma
                                      else if (params$do_alignment)         d_asth$pp_aligned
                                      else                                    d_asth$pp,
                                      which(d_asth$proc$y == 2)),
  asthma_ctrl    = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$asthma
                                      else if (params$do_alignment)         d_asth$pp_aligned
                                      else                                    d_asth$pp,
                                      which(d_asth$proc$y == 1)),
  resp_flu       = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$respiratory
                                      else if (params$do_alignment)         d_resp$pp_aligned
                                      else                                    d_resp$pp,
                                      which(d_resp$proc$y == 2)),
  resp_other     = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$respiratory
                                      else if (params$do_alignment)         d_resp$pp_aligned
                                      else                                    d_resp$pp,
                                      which(d_resp$proc$y == 3)),
  resp_ctrl      = subset_peak_params(if (params$do_cross_cohort_alignment) cross_aligned$respiratory
                                      else if (params$do_alignment)         d_resp$pp_aligned
                                      else                                    d_resp$pp,
                                      which(d_resp$proc$y == 1))
)

universal <- build_universal_catalog(uni_pp_sources,
  eps_rt = params$eps_rt, eps_cv = params$eps_cv,
  min_support_frac = params$min_support_frac)
sprintf("Universal catalog: %d compounds", nrow(universal$compounds))
```

## Per-source presence pattern

For each merged compound, how many source catalogs contribute?

```{r per-source}
per_source_df <- do.call(rbind, universal$compounds$per_source_prevalence) |>
  as.data.frame()
colnames(per_source_df) <- universal$parameters$source_names
per_source_df$n_sources_present <-
  rowSums(per_source_df[, universal$parameters$source_names] > 0)
per_source_df$compound_id <- universal$compounds$compound_id

sprintf("Compounds present in ALL 9 sources: %d",
        sum(per_source_df$n_sources_present == 9))
sprintf("Compounds present in >= 5 sources:  %d",
        sum(per_source_df$n_sources_present >= 5))
sprintf("Compounds present in only 1 source: %d",
        sum(per_source_df$n_sources_present == 1))

ggplot(per_source_df, aes(x = n_sources_present)) +
  geom_bar(fill = "#B2182B") +
  scale_x_continuous(breaks = 1:9) +
  labs(x = "n sources compound is present in",
       y = "n compounds",
       title = "Distribution of compound source-coverage in universal catalog") +
  theme_minimal()
```

## Spatial map of universal catalog by source coverage

Where in (RT, CV) space do universally-present compounds sit vs disease-specific compounds?

```{r universal-spatial, fig.height = 5.5, fig.width = 8}
uni_map <- tibble(
  rt_loc = universal$compounds$rt_loc,
  cv_loc = universal$compounds$cv_loc,
  n_sources = per_source_df$n_sources_present,
  prevalence = universal$compounds$prevalence
) |> mutate(coverage_bin = case_when(
  n_sources == 9 ~ "9 sources (universal)",
  n_sources >= 5 ~ "5-8 sources",
  n_sources >= 2 ~ "2-4 sources",
  TRUE ~ "1 source only"
))
uni_map$coverage_bin <- factor(uni_map$coverage_bin,
  levels = c("1 source only", "2-4 sources",
              "5-8 sources", "9 sources (universal)"))

ggplot(uni_map, aes(x = cv_loc, y = rt_loc, color = coverage_bin,
                     size = prevalence)) +
  geom_point(alpha = 0.55) +
  scale_color_manual(values = c(
    "1 source only"          = "#B2182B",
    "2-4 sources"            = "#EF8A62",
    "5-8 sources"            = "#67A9CF",
    "9 sources (universal)"  = "#08306B")) +
  scale_size_continuous(range = c(0.3, 3), guide = "none") +
  scale_y_reverse() +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Universal catalog compounds by source-coverage",
       color = "source coverage") +
  theme_minimal()
```

Dark-blue points (universal) show the "shared backbone" position across all cohorts. Red points (single-source) are cohort-specific compound positions — a filter for cohort-effect diagnosis: if red points cluster in the (RT, CV) neighborhood of universal compounds, they might be jitter-affected duplicates; if they sit far from any universal compound, they're distinctively new positions.

Compounds at `n_sources_present = 9` are candidates for "universal human volatile signature" — appear across every cohort we tested.

## Compounds unique to disease groups

Compounds present in disease sources but NOT in any control source:

```{r disease-only-compounds}
disease_sources <- c("aki", "chf", "asthma", "resp_flu", "resp_other")
control_sources <- c("aki_ctrl", "chf_ctrl", "asthma_ctrl", "resp_ctrl")

per_source_df$disease_hits <- rowSums(per_source_df[, disease_sources] > 0)
per_source_df$control_hits <- rowSums(per_source_df[, control_sources] > 0)

disease_only <- per_source_df[per_source_df$disease_hits > 0 &
                                  per_source_df$control_hits == 0, ]
control_only <- per_source_df[per_source_df$control_hits > 0 &
                                  per_source_df$disease_hits == 0, ]

sprintf("Compounds in >=1 disease source AND 0 control sources: %d",
        nrow(disease_only))
sprintf("Compounds in >=1 control source AND 0 disease sources: %d",
        nrow(control_only))
```

Cross-reference these against the within-disease biomarker candidates from Section 3 — compounds hitting both lists are the strongest signal.

## Multi-disease shared but control-absent compounds

Compounds in ≥ 2 disease sources but 0 control sources: these are candidates for "generic pathology" biomarkers.

```{r multi-disease-shared}
multi_disease <- per_source_df[per_source_df$disease_hits >= 2 &
                                  per_source_df$control_hits == 0, ]
sprintf("Compounds in >= 2 disease sources but 0 controls: %d",
        nrow(multi_disease))
if (nrow(multi_disease) > 0) {
  # Add compound rt/cv for reference
  multi_disease_details <- universal$compounds[
    match(multi_disease$compound_id, universal$compounds$compound_id),
    c("compound_id", "rt_loc", "cv_loc", "prevalence")]
  multi_disease_details$n_diseases_present <- multi_disease$disease_hits
  print(knitr::kable(head(multi_disease_details, 20), digits = 3))
}
```

# 7. Save outputs

Save the universal catalog and comparison objects so downstream steps (biomarker candidate ranking, cross-cohort encoder pretraining) can pick up without recomputing.

```{r save-outputs}
outdir <- "catalog_comparison_outputs"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
saveRDS(all_cats,             file.path(outdir, "all_catalogs.rds"))
saveRDS(within_comparisons,   file.path(outdir, "within_disease_comparisons.rds"))
saveRDS(control_comparisons,  file.path(outdir, "control_vs_control_comparisons.rds"))
saveRDS(disease_comparisons,  file.path(outdir, "disease_vs_disease_comparisons.rds"))
saveRDS(universal,            file.path(outdir, "universal_catalog.rds"))
sprintf("All comparison objects saved to %s/", outdir)
```

# What to do next

- **Rank biomarker candidates**: combine the "only in disease" list (Section 3) with the "high `|prevalence_delta|`" list (Section 3) and the "cross-disease shared but control-absent" list (Section 6). Compounds hitting all three are your strongest candidates.
- **Universal encoder pretraining**: `pretrain_denoising_online(catalog = universal, ...)` gives you an encoder trained on volatiles from all cohorts. Test whether it outperforms per-cohort encoders on any specific disease's classification task.
- **Check the summary table (Section 2)** for anomalies: if one cohort has 10x fewer pool peaks per sample, or 10x more singletons, that cohort might have a preprocessing or peak-detection issue that should be addressed before trusting the cross-cohort analysis.
