---
title: "Cross-cohort catalog comparison: AKI, CHF, asthma, respiratory"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    fig_width: 8
    fig_height: 4.5
params:
  output_root: "output"          # directory containing aki/, chf/, asthma/, respiratory/ subdirs
  eps_rt: 0.003
  eps_cv: 0.015
  min_support_frac: 0.10
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dpi = 96)
library(classydms)
library(dplyr)
library(tibble)
library(ggplot2)
library(tidyr)
library(purrr)
```

# Purpose

You have peak_params from four cohorts (AKI, CHF, asthma, respiratory) with disease/control labels. This report:

1. Loads and builds a catalog per condition (9 total: 4 diseases + 4 controls + respiratory's "other")
2. Reports a summary table for each catalog
3. **Within-disease**: compares each disease to its matched control — biomarker candidates via prevalence delta and unique compounds
4. **Cross-condition controls**: compares the 4 control catalogs — are "healthy human volatile signatures" consistent across cohorts?
5. **Cross-condition diseases**: compares diseases to each other — shared vs disease-specific compounds
6. **Universal catalog**: `merge_catalogs()` across everything, with per-source prevalence tracking

Renders can take a while (each catalog is ~500k pool peaks). Cache is enabled on the expensive chunks so re-renders are fast.

# 1. Load data and build catalogs

```{r load-build, cache = TRUE}
load_and_backfill <- function(cohort_name, root = params$output_root) {
  proc <- readRDS(file.path(root, cohort_name, "processed.rds"))
  pp   <- readRDS(file.path(root, cohort_name, "peak_params.rds"))
  # Backfill if missing sample_idx_raw
  if (is.null(pp$sample_idx_raw)) {
    pp <- backfill_sample_idx_raw(pp, sample_names = proc$ids)
  }
  list(proc = proc, pp = pp)
}

# Load
d_aki  <- load_and_backfill("aki")
d_chf  <- load_and_backfill("chf")
d_asth <- load_and_backfill("asthma")
d_resp <- load_and_backfill("respiratory")

# Subset by class and build catalogs
build_from <- function(pp) build_peak_catalog(
  pp,
  eps_rt = params$eps_rt,
  eps_cv = params$eps_cv,
  min_support_frac = params$min_support_frac
)

cat_aki           <- build_from(subset_peak_params(d_aki$pp,  which(d_aki$proc$y  == 2)))
cat_aki_control   <- build_from(subset_peak_params(d_aki$pp,  which(d_aki$proc$y  == 1)))
cat_chf           <- build_from(subset_peak_params(d_chf$pp,  which(d_chf$proc$y  == 2)))
cat_chf_control   <- build_from(subset_peak_params(d_chf$pp,  which(d_chf$proc$y  == 1)))
cat_asthma        <- build_from(subset_peak_params(d_asth$pp, which(d_asth$proc$y == 2)))
cat_asthma_control<- build_from(subset_peak_params(d_asth$pp, which(d_asth$proc$y == 1)))
cat_resp_flu      <- build_from(subset_peak_params(d_resp$pp, which(d_resp$proc$y == 2)))
cat_resp_other    <- build_from(subset_peak_params(d_resp$pp, which(d_resp$proc$y == 3)))
cat_resp_control  <- build_from(subset_peak_params(d_resp$pp, which(d_resp$proc$y == 1)))

all_cats <- list(
  aki             = cat_aki,             aki_control    = cat_aki_control,
  chf             = cat_chf,             chf_control    = cat_chf_control,
  asthma          = cat_asthma,          asthma_control = cat_asthma_control,
  resp_flu        = cat_resp_flu,        resp_other     = cat_resp_other,
  resp_control    = cat_resp_control
)
```

# 2. Summary table across all catalogs

```{r summary-table}
summary_row <- function(cat, name) {
  s <- catalog_summary(cat, print = FALSE)
  tibble(
    catalog     = name,
    n_samples   = s$n_samples,
    n_pool_peaks= s$n_pool_peaks,
    n_compounds = s$n_compounds,
    n_singletons= s$n_singletons,
    frac_singleton = round(s$fraction_singleton, 3),
    median_prev = round(median(cat$compounds$prevalence), 3),
    mean_prev   = round(mean(cat$compounds$prevalence), 3),
    n_anchor_90pct = sum(cat$compounds$prevalence >= 0.90)
  )
}
summary_df <- purrr::imap_dfr(all_cats, ~summary_row(.x, .y))
knitr::kable(summary_df)
```

**How to read this table:**

- Similar `n_pool_peaks / n_samples` across cohorts → similar peak detection settings and instrument behavior (good baseline)
- Very different `n_compounds` → could reflect real differences OR that eps needs tuning for one cohort
- `frac_singleton` — high (>20%) suggests eps is too tight or many compounds are truly rare; low (<10%) suggests eps is reasonable

# 3. Within-disease: disease vs control per cohort

For each of the 4 diseases, compare disease catalog to its matched control catalog. Focuses biomarker discovery: which compounds appear in disease but not control (or at different prevalence)?

```{r within-disease-compare}
within_pairs <- list(
  aki    = list(disease = cat_aki,       control = cat_aki_control),
  chf    = list(disease = cat_chf,       control = cat_chf_control),
  asthma = list(disease = cat_asthma,    control = cat_asthma_control),
  resp_flu   = list(disease = cat_resp_flu,   control = cat_resp_control),
  resp_other = list(disease = cat_resp_other, control = cat_resp_control)
)

within_comparisons <- purrr::imap(within_pairs, function(pair, name) {
  compare_catalogs(pair$disease, pair$control,
                    name_a = "disease", name_b = "control")
})

within_summary <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  cbind(comparison = name, comp$summary)
})
knitr::kable(within_summary)
```

## Prevalence-delta distribution per disease

```{r within-delta, fig.height = 6}
delta_df <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  tibble(comparison = name, prevalence_delta = comp$shared$prevalence_delta)
})
ggplot(delta_df, aes(x = prevalence_delta)) +
  geom_histogram(bins = 40, fill = "#B2182B") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ comparison, ncol = 2, scales = "free_y") +
  labs(x = "prevalence delta (control − disease)",
       y = "n shared compounds",
       title = "Prevalence delta for shared compounds within each disease pair") +
  theme_minimal()
```

**Reading**: bimodal or long-tailed distributions on either side indicate compounds systematically expressed differently between disease and control — the biomarker candidates.

## Top biomarker candidates per disease

For each disease, the shared compounds with the largest `|prevalence_delta|` are candidates: same (RT, CV) coordinates, very different prevalence between the two groups.

```{r top-biomarkers}
top_biom <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  comp$shared |>
    dplyr::arrange(dplyr::desc(abs(prevalence_delta))) |>
    head(10) |>
    mutate(comparison = name) |>
    select(comparison, rt_loc_disease, cv_loc_disease,
           prevalence_disease, prevalence_control, prevalence_delta,
           spatial_distance)
})
knitr::kable(top_biom, digits = 3)
```

Negative `prevalence_delta` = compound MORE prevalent in disease (candidate biomarker of the disease itself). Positive = LESS prevalent in disease (candidate for "protective" or "control-associated" compounds).

## Only-in-disease and only-in-control compounds

```{r unique-counts}
unique_counts <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  tibble(
    comparison = name,
    only_disease = nrow(comp$only_a),
    only_control = nrow(comp$only_b),
    disease_top_prevalence = max(comp$only_a$prevalence, na.rm = TRUE),
    control_top_prevalence = max(comp$only_b$prevalence, na.rm = TRUE)
  )
})
knitr::kable(unique_counts)
```

Compounds only in disease that also have high prevalence within disease are the strongest biomarker candidates — same physical position doesn't exist in controls at all, AND it's consistently expressed in disease samples.

## Spatial view of unique compounds per disease

Each panel shows one disease pair. **Grey dots** = compounds shared between disease and control (background context). **Red dots** = compounds only in disease. **Blue dots** = compounds only in control. Point size = compound prevalence within its catalog. Cluster patterns of the colored dots often reveal spatial regions of the (RT, CV) space that differentiate the classes.

```{r unique-spatial, fig.height = 9, fig.width = 10}
unique_spatial_df <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  bind_rows(
    if (nrow(comp$shared) > 0)
      tibble(comparison = name, group = "shared",
             rt_loc = comp$shared$rt_loc_disease,
             cv_loc = comp$shared$cv_loc_disease,
             prevalence = (comp$shared$prevalence_disease +
                             comp$shared$prevalence_control) / 2)
    else tibble(comparison = character(0), group = character(0),
                rt_loc = numeric(0), cv_loc = numeric(0),
                prevalence = numeric(0)),
    if (nrow(comp$only_a) > 0)
      tibble(comparison = name, group = "only disease",
             rt_loc = comp$only_a$rt_loc,
             cv_loc = comp$only_a$cv_loc,
             prevalence = comp$only_a$prevalence)
    else tibble(comparison = character(0), group = character(0),
                rt_loc = numeric(0), cv_loc = numeric(0),
                prevalence = numeric(0)),
    if (nrow(comp$only_b) > 0)
      tibble(comparison = name, group = "only control",
             rt_loc = comp$only_b$rt_loc,
             cv_loc = comp$only_b$cv_loc,
             prevalence = comp$only_b$prevalence)
    else tibble(comparison = character(0), group = character(0),
                rt_loc = numeric(0), cv_loc = numeric(0),
                prevalence = numeric(0))
  )
})
unique_spatial_df$group <- factor(unique_spatial_df$group,
  levels = c("shared", "only disease", "only control"))

ggplot(unique_spatial_df,
       aes(x = cv_loc, y = rt_loc,
           color = group, size = prevalence, alpha = group)) +
  geom_point() +
  scale_color_manual(values = c("shared"       = "grey60",
                                  "only disease" = "#B2182B",
                                  "only control" = "#2166AC")) +
  scale_alpha_manual(values = c("shared" = 0.2, "only disease" = 0.85,
                                  "only control" = 0.85), guide = "none") +
  scale_size_continuous(range = c(0.5, 3.5)) +
  scale_y_reverse() +
  facet_wrap(~ comparison, ncol = 2) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Spatial distribution of unique vs shared compounds",
       color = "compound group") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Top only-in-disease compounds table

Highest-prevalence disease-only compounds per cohort — the concrete biomarker candidates from the unique-compound analysis:

```{r top-only-disease}
top_only_disease <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  if (nrow(comp$only_a) == 0) return(NULL)
  comp$only_a |>
    arrange(desc(prevalence)) |>
    head(10) |>
    mutate(comparison = name) |>
    select(comparison, compound_id, rt_loc, cv_loc,
           prevalence, n_observations, n_samples)
})
knitr::kable(top_only_disease, digits = 3)
```

# 4. Cross-condition controls: how consistent are "healthy" volatile signatures?

If the 4 control catalogs represent healthy human volatile emissions from similar sample-collection protocols, we'd expect them to share most compounds. Big disagreements between controls could reflect batch/site/protocol differences.

```{r control-vs-control}
control_cats <- list(
  aki_control    = cat_aki_control,
  chf_control    = cat_chf_control,
  asthma_control = cat_asthma_control,
  resp_control   = cat_resp_control
)

# All pairwise combinations of controls
control_pairs <- t(combn(names(control_cats), 2))
control_comparisons <- lapply(seq_len(nrow(control_pairs)), function(i) {
  a <- control_pairs[i, 1]; b <- control_pairs[i, 2]
  cmp <- compare_catalogs(control_cats[[a]], control_cats[[b]],
                           name_a = a, name_b = b)
  list(pair = paste(a, "vs", b), summary = cmp$summary,
       shared = cmp$shared)
})
ctrl_summary <- do.call(rbind, lapply(control_comparisons, function(x)
  cbind(pair = x$pair, x$summary)))
knitr::kable(ctrl_summary)
```

**High `frac_shared_a` and `frac_shared_b` on all pairs (say > 0.7)** → controls are consistent; healthy signature reproduces across cohorts. Low fractions suggest cohort-specific artifacts (batch effect, different equipment, different sample handling) — a red flag for cross-cohort analysis.

## Prevalence-delta on controls

Prevalence-delta for shared compounds between controls SHOULD be small if the cohorts are drawn from similar populations. Large deltas indicate cohort-level differences even in "baseline" volatiles.

```{r control-delta-summary}
ctrl_delta <- lapply(control_comparisons, function(x)
  tibble(pair = x$pair, prevalence_delta = x$shared$prevalence_delta))
ctrl_delta_df <- do.call(rbind, ctrl_delta)
ctrl_delta_df |>
  group_by(pair) |>
  summarise(
    median_abs_delta = median(abs(prevalence_delta)),
    q75_abs_delta    = quantile(abs(prevalence_delta), 0.75),
    max_abs_delta    = max(abs(prevalence_delta)),
    n_delta_gt_0.20  = sum(abs(prevalence_delta) > 0.20)
  ) |>
  knitr::kable(digits = 3)
```

# 5. Cross-condition diseases: shared vs unique

Same analysis but comparing diseases to each other. Compounds shared across multiple diseases at similar prevalence are "generic pathology" candidates; compounds unique to one disease are disease-specific candidates.

```{r disease-vs-disease}
disease_cats <- list(
  aki      = cat_aki,
  chf      = cat_chf,
  asthma   = cat_asthma,
  resp_flu = cat_resp_flu,
  resp_other = cat_resp_other
)

disease_pairs <- t(combn(names(disease_cats), 2))
disease_comparisons <- lapply(seq_len(nrow(disease_pairs)), function(i) {
  a <- disease_pairs[i, 1]; b <- disease_pairs[i, 2]
  cmp <- compare_catalogs(disease_cats[[a]], disease_cats[[b]],
                           name_a = a, name_b = b)
  list(pair = paste(a, "vs", b), summary = cmp$summary,
       shared = cmp$shared)
})
dis_summary <- do.call(rbind, lapply(disease_comparisons, function(x)
  cbind(pair = x$pair, x$summary)))
knitr::kable(dis_summary)
```

## Which disease pairs have the most compound overlap?

```{r disease-overlap-heatmap, fig.height = 4}
overlap_mat <- matrix(NA_real_, nrow = length(disease_cats),
                       ncol = length(disease_cats),
                       dimnames = list(names(disease_cats),
                                        names(disease_cats)))
diag(overlap_mat) <- 1
for (i in seq_len(nrow(disease_pairs))) {
  a <- disease_pairs[i, 1]; b <- disease_pairs[i, 2]
  frac_a <- disease_comparisons[[i]]$summary$frac_shared_a
  frac_b <- disease_comparisons[[i]]$summary$frac_shared_b
  overlap_mat[a, b] <- frac_a
  overlap_mat[b, a] <- frac_b
}
overlap_df <- as_tibble(overlap_mat, rownames = "row") |>
  pivot_longer(-row, names_to = "col", values_to = "frac_shared")

ggplot(overlap_df, aes(x = col, y = row, fill = frac_shared)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", frac_shared)), size = 3) +
  scale_fill_gradient(low = "#F7FBFF", high = "#08306B",
                       name = "frac shared") +
  labs(x = "col catalog (denominator)", y = "row catalog",
       title = "Fraction of row-catalog compounds found in col-catalog") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Read the matrix: element `[i, j]` = fraction of compounds in disease `i` that also appear in disease `j`. Note asymmetry — smaller catalogs have higher fractions.

# 5b. Catalog spatial heatmaps

Faceted (RT × CV) hexbin density of compounds per catalog. Where the catalog concentrates its compounds spatially — and how much the various cohorts overlap in coverage — is easier to see side by side than one catalog at a time.

## Density heatmap for each catalog

```{r catalog-heatmap, fig.height = 10, fig.width = 10}
all_compounds_df <- purrr::imap_dfr(all_cats, function(cat, name) {
  tibble(
    catalog    = name,
    rt_loc     = cat$compounds$rt_loc,
    cv_loc     = cat$compounds$cv_loc,
    prevalence = cat$compounds$prevalence
  )
})
# Consistent facet order: diseases first, controls second
all_compounds_df$catalog <- factor(all_compounds_df$catalog,
  levels = c("aki", "chf", "asthma", "resp_flu", "resp_other",
              "aki_control", "chf_control", "asthma_control",
              "resp_control"))

ggplot(all_compounds_df, aes(x = cv_loc, y = rt_loc)) +
  geom_hex(bins = c(20, 40)) +
  scale_fill_gradient(low = "#F7FBFF", high = "#08306B",
                       name = "n compounds") +
  scale_y_reverse() +
  facet_wrap(~ catalog, ncol = 3) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Compound density heatmap per catalog") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

**Reading:** hotspots that appear in every panel are conserved compound regions (universal volatile signature). Panels with different hotspot patterns from the rest are potential batch effects or genuine cohort-specific chemistry.

## Prevalence-weighted spatial map per catalog

Same layout but each compound sized by its prevalence — highlights the anchor compounds visually.

```{r catalog-prevalence-map, fig.height = 10, fig.width = 10}
ggplot(all_compounds_df, aes(x = cv_loc, y = rt_loc,
                              size = prevalence, color = prevalence)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "#F7FBFF", high = "#B2182B",
                        name = "prevalence") +
  scale_size_continuous(range = c(0.3, 2.5), guide = "none") +
  scale_y_reverse() +
  facet_wrap(~ catalog, ncol = 3) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Compound prevalence per catalog") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Bright red points (high prevalence, in most samples) that align across panels are the anchor compounds present in every cohort.

## Disease-only overlay across catalogs

Overlay compounds that are unique to each disease (from within-disease comparisons) on a single plot — reveals whether disease-only compounds cluster in different regions per disease.

```{r disease-only-overlay, fig.height = 6, fig.width = 8}
disease_only_all <- purrr::imap_dfr(within_comparisons, function(comp, name) {
  if (nrow(comp$only_a) == 0) return(NULL)
  tibble(disease = name,
          rt_loc = comp$only_a$rt_loc,
          cv_loc = comp$only_a$cv_loc,
          prevalence = comp$only_a$prevalence)
})

ggplot(disease_only_all,
       aes(x = cv_loc, y = rt_loc, color = disease, size = prevalence)) +
  geom_point(alpha = 0.7) +
  scale_y_reverse() +
  scale_size_continuous(range = c(0.5, 3.5)) +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "All disease-only compounds overlaid",
       color = "disease") +
  theme_minimal()
```

Regions of the plot where colors mix are areas multiple diseases have unique compounds — could reflect "generic pathology" positions. Regions dominated by one color are disease-specific spatial signatures.

# 6. Universal catalog across everything

```{r universal, cache = TRUE}
universal <- merge_catalogs(
  list(cat_aki, cat_aki_control,
       cat_chf, cat_chf_control,
       cat_asthma, cat_asthma_control,
       cat_resp_flu, cat_resp_other, cat_resp_control),
  names = c("aki", "aki_ctrl",
             "chf", "chf_ctrl",
             "asthma", "asthma_ctrl",
             "resp_flu", "resp_other", "resp_ctrl")
)
sprintf("Universal catalog: %d compounds",
        nrow(universal$compounds))
```

## Per-source presence pattern

For each merged compound, how many source catalogs contribute?

```{r per-source}
per_source_df <- do.call(rbind, universal$compounds$per_source_prevalence) |>
  as.data.frame()
colnames(per_source_df) <- universal$parameters$source_names
per_source_df$n_sources_present <-
  rowSums(per_source_df[, universal$parameters$source_names] > 0)
per_source_df$compound_id <- universal$compounds$compound_id

sprintf("Compounds present in ALL 9 sources: %d",
        sum(per_source_df$n_sources_present == 9))
sprintf("Compounds present in >= 5 sources:  %d",
        sum(per_source_df$n_sources_present >= 5))
sprintf("Compounds present in only 1 source: %d",
        sum(per_source_df$n_sources_present == 1))

ggplot(per_source_df, aes(x = n_sources_present)) +
  geom_bar(fill = "#B2182B") +
  scale_x_continuous(breaks = 1:9) +
  labs(x = "n sources compound is present in",
       y = "n compounds",
       title = "Distribution of compound source-coverage in universal catalog") +
  theme_minimal()
```

## Spatial map of universal catalog by source coverage

Where in (RT, CV) space do universally-present compounds sit vs disease-specific compounds?

```{r universal-spatial, fig.height = 5.5, fig.width = 8}
uni_map <- tibble(
  rt_loc = universal$compounds$rt_loc,
  cv_loc = universal$compounds$cv_loc,
  n_sources = per_source_df$n_sources_present,
  prevalence = universal$compounds$prevalence
) |> mutate(coverage_bin = case_when(
  n_sources == 9 ~ "9 sources (universal)",
  n_sources >= 5 ~ "5-8 sources",
  n_sources >= 2 ~ "2-4 sources",
  TRUE ~ "1 source only"
))
uni_map$coverage_bin <- factor(uni_map$coverage_bin,
  levels = c("1 source only", "2-4 sources",
              "5-8 sources", "9 sources (universal)"))

ggplot(uni_map, aes(x = cv_loc, y = rt_loc, color = coverage_bin,
                     size = prevalence)) +
  geom_point(alpha = 0.55) +
  scale_color_manual(values = c(
    "1 source only"          = "#B2182B",
    "2-4 sources"            = "#EF8A62",
    "5-8 sources"            = "#67A9CF",
    "9 sources (universal)"  = "#08306B")) +
  scale_size_continuous(range = c(0.3, 3), guide = "none") +
  scale_y_reverse() +
  labs(x = "CV (fraction)", y = "RT (fraction)",
       title = "Universal catalog compounds by source-coverage",
       color = "source coverage") +
  theme_minimal()
```

Dark-blue points (universal) show the "shared backbone" position across all cohorts. Red points (single-source) are cohort-specific compound positions — a filter for cohort-effect diagnosis: if red points cluster in the (RT, CV) neighborhood of universal compounds, they might be jitter-affected duplicates; if they sit far from any universal compound, they're distinctively new positions.

Compounds at `n_sources_present = 9` are candidates for "universal human volatile signature" — appear across every cohort we tested.

## Compounds unique to disease groups

Compounds present in disease sources but NOT in any control source:

```{r disease-only-compounds}
disease_sources <- c("aki", "chf", "asthma", "resp_flu", "resp_other")
control_sources <- c("aki_ctrl", "chf_ctrl", "asthma_ctrl", "resp_ctrl")

per_source_df$disease_hits <- rowSums(per_source_df[, disease_sources] > 0)
per_source_df$control_hits <- rowSums(per_source_df[, control_sources] > 0)

disease_only <- per_source_df[per_source_df$disease_hits > 0 &
                                  per_source_df$control_hits == 0, ]
control_only <- per_source_df[per_source_df$control_hits > 0 &
                                  per_source_df$disease_hits == 0, ]

sprintf("Compounds in >=1 disease source AND 0 control sources: %d",
        nrow(disease_only))
sprintf("Compounds in >=1 control source AND 0 disease sources: %d",
        nrow(control_only))
```

Cross-reference these against the within-disease biomarker candidates from Section 3 — compounds hitting both lists are the strongest signal.

## Multi-disease shared but control-absent compounds

Compounds in ≥ 2 disease sources but 0 control sources: these are candidates for "generic pathology" biomarkers.

```{r multi-disease-shared}
multi_disease <- per_source_df[per_source_df$disease_hits >= 2 &
                                  per_source_df$control_hits == 0, ]
sprintf("Compounds in >= 2 disease sources but 0 controls: %d",
        nrow(multi_disease))
if (nrow(multi_disease) > 0) {
  # Add compound rt/cv for reference
  multi_disease_details <- universal$compounds[
    match(multi_disease$compound_id, universal$compounds$compound_id),
    c("compound_id", "rt_loc", "cv_loc", "prevalence")]
  multi_disease_details$n_diseases_present <- multi_disease$disease_hits
  print(knitr::kable(head(multi_disease_details, 20), digits = 3))
}
```

# 7. Save outputs

Save the universal catalog and comparison objects so downstream steps (biomarker candidate ranking, cross-cohort encoder pretraining) can pick up without recomputing.

```{r save-outputs}
outdir <- "catalog_comparison_outputs"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
saveRDS(all_cats,             file.path(outdir, "all_catalogs.rds"))
saveRDS(within_comparisons,   file.path(outdir, "within_disease_comparisons.rds"))
saveRDS(control_comparisons,  file.path(outdir, "control_vs_control_comparisons.rds"))
saveRDS(disease_comparisons,  file.path(outdir, "disease_vs_disease_comparisons.rds"))
saveRDS(universal,            file.path(outdir, "universal_catalog.rds"))
sprintf("All comparison objects saved to %s/", outdir)
```

# What to do next

- **Rank biomarker candidates**: combine the "only in disease" list (Section 3) with the "high `|prevalence_delta|`" list (Section 3) and the "cross-disease shared but control-absent" list (Section 6). Compounds hitting all three are your strongest candidates.
- **Universal encoder pretraining**: `pretrain_denoising_online(catalog = universal, ...)` gives you an encoder trained on volatiles from all cohorts. Test whether it outperforms per-cohort encoders on any specific disease's classification task.
- **Check the summary table (Section 2)** for anomalies: if one cohort has 10x fewer pool peaks per sample, or 10x more singletons, that cohort might have a preprocessing or peak-detection issue that should be addressed before trusting the cross-cohort analysis.
