---
title: "Stage 1: Prevalence ladder"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
params:
  catalog_path: "catalog_comparison_outputs_physical/all_catalogs.rds"
  catalog_name: "aki"
  H: 1200
  W: 40
  # Fixed simulation knobs (from Stage 5 realistic-preset defaults)
  n_biomarkers: 5
  anchor_prevalence_threshold: 0.9
  min_intensity_quantile: 0.95
  case_intensity_mult: 1.2
  n_per_class: 200
  n_seeds: 5
  # Prevalence ladder: comma-separated case_prevalence values;
  # control_prevalence = 1 - case_prevalence.
  case_prevalence_grid: "0.7,0.8,0.9,1.0"
  # Pretraining (shared across all rungs — encoder is signal-agnostic)
  n_pretrain_samples: 400
  pretrain_epochs: 20
  # Classification — Model to run at each rung. Fastest is encoder_enet.
  model: "encoder_enet"       # one of: pixel_enet, amp_enet, encoder_enet, pretrained_cnn
  cv_k: 5
  cnn_epochs: 15
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(classydms)
library(dplyr)
library(tibble)
library(ggplot2)
library(purrr)
library(pROC)
library(glmnet)
library(torch)
```

# Purpose

Fix everything except case/control prevalence. Sweep `case_prevalence` and set `control_prevalence = 1 - case_prevalence` (symmetric ladder). Record AUC per condition, averaged across seeds.

**Answers:** what is the smallest prevalence delta at which the pipeline reliably detects the signal?

```{r knobs-echo, results = "asis"}
cp_grid <- as.numeric(strsplit(params$case_prevalence_grid, ",")[[1]])
n_seeds <- params$n_seeds
cat(sprintf(
"| knob | value |\n|---|---|\n| catalog | %s |\n| n_biomarkers | %d |\n| min_intensity_quantile | %.2f |\n| case_intensity_mult | %.2f |\n| n_per_class | %d |\n| n_seeds | %d |\n| ladder (case_prevalence values) | %s |\n| model | %s |\n",
  params$catalog_name, params$n_biomarkers, params$min_intensity_quantile,
  params$case_intensity_mult, params$n_per_class, n_seeds,
  paste(cp_grid, collapse = ", "), params$model))
```

# 1. Load catalog and pretrain encoder

```{r load-catalog}
all_cats <- readRDS(params$catalog_path)
cat_use <- all_cats[[params$catalog_name]]
sprintf("Loaded %s catalog: %d compounds",
        params$catalog_name, nrow(cat_use$compounds))
```

```{r pretrain, cache = TRUE, eval = params$model %in% c("encoder_enet","pretrained_cnn")}
enc_res <- pretrain_autoencoder_from_catalog(cat_use,
  H = params$H, W = params$W,
  n_synthetic = params$n_pretrain_samples,
  epochs = params$pretrain_epochs,
  batch_size = 16, device = "cpu", seed = 1)
encoder <- enc_res$encoder
sprintf("Encoder pretrained: final MSE = %.5f",
        tail(enc_res$loss_history, 1))
```

# 2. Run the ladder

```{r ladder-helper}
run_one <- function(sim, catalog, encoder, model, cv_k, cnn_epochs, seed) {
  Z_list <- lapply(sim$samples, `[[`, "Z")
  y <- sim$y
  set.seed(seed)
  if (model == "pixel_enet") {
    X <- do.call(rbind, lapply(Z_list, as.vector))
    fold_id <- sample(rep(seq_len(cv_k), length.out = length(y)))
    oof <- numeric(length(y))
    for (k in seq_len(cv_k)) {
      tr <- fold_id != k; te <- !tr
      fit <- glmnet::cv.glmnet(X[tr,], y[tr], family="binomial",
                                alpha=0.5, nfolds=3)
      oof[te] <- as.numeric(predict(fit, newx=X[te,], s="lambda.min",
                                       type="response"))
    }
    return(as.numeric(pROC::auc(pROC::roc(y, oof, quiet=TRUE))))
  }
  if (model == "amp_enet") {
    cmp <- catalog$compounds
    rt_use <- if ("rt_frac" %in% names(cmp)) cmp$rt_frac else cmp$rt_loc
    cv_use <- if ("cv_frac" %in% names(cmp)) cmp$cv_frac else cmp$cv_loc
    H <- nrow(Z_list[[1]]); W <- ncol(Z_list[[1]])
    row_range <- function(rt) {
      i0 <- max(1, floor(rt * H) - 3); i1 <- min(H, floor(rt * H) + 3)
      seq(i0, i1)
    }
    col_range <- function(cv) {
      i0 <- max(1, floor(cv * W) - 2); i1 <- min(W, floor(cv * W) + 2)
      seq(i0, i1)
    }
    X <- do.call(rbind, lapply(Z_list, function(Z) {
      vapply(seq_len(nrow(cmp)), function(i) {
        sum(Z[row_range(rt_use[i]), col_range(cv_use[i])])
      }, numeric(1))
    }))
    fold_id <- sample(rep(seq_len(cv_k), length.out = length(y)))
    oof <- numeric(length(y))
    for (k in seq_len(cv_k)) {
      tr <- fold_id != k; te <- !tr
      fit <- glmnet::cv.glmnet(X[tr,], y[tr], family="binomial",
                                alpha=0.5, nfolds=3)
      oof[te] <- as.numeric(predict(fit, newx=X[te,], s="lambda.min",
                                       type="response"))
    }
    return(as.numeric(pROC::auc(pROC::roc(y, oof, quiet=TRUE))))
  }
  if (model == "encoder_enet") {
    return(cv_encoder_elastic_net(encoder, Z_list, y,
                                    k=cv_k, alpha=0.5, seed=seed)$auc)
  }
  if (model == "pretrained_cnn") {
    return(cv_pretrained_oof(Z_list, y, encoder,
                              k=cv_k, epochs=cnn_epochs, seed=seed)$auc)
  }
  stop("Unknown model: ", model)
}
```

```{r ladder-run, cache = TRUE}
grid <- expand.grid(
  case_prevalence = cp_grid,
  sim_seed = seq_len(n_seeds),
  KEEP.OUT.ATTRS = FALSE
)

results <- purrr::pmap_dfr(grid, function(case_prevalence, sim_seed) {
  message(sprintf("Rung: case_prev=%.2f, seed=%d",
                  case_prevalence, sim_seed))
  spec <- catalog_biomarkers_universal_backdrop(cat_use,
    n_biomarkers = params$n_biomarkers,
    anchor_prevalence_threshold = params$anchor_prevalence_threshold,
    min_intensity_quantile = params$min_intensity_quantile,
    case_prevalence = case_prevalence,
    control_prevalence = 1 - case_prevalence,
    case_intensity_mult = params$case_intensity_mult,
    seed = sim_seed * 1000 + 7)
  sim <- simulate_case_control_from_catalog(cat_use, spec,
    n_cases = params$n_per_class,
    n_controls = params$n_per_class,
    H = params$H, W = params$W,
    seed = sim_seed * 1000 + 8)
  auc <- run_one(sim, cat_use,
                  if (params$model %in% c("encoder_enet","pretrained_cnn"))
                    encoder else NULL,
                  params$model, params$cv_k, params$cnn_epochs, sim_seed)
  tibble(case_prevalence = case_prevalence,
          delta = 2 * case_prevalence - 1,
          sim_seed = sim_seed,
          auc = auc)
})
results
```

# 3. AUC vs prevalence delta

```{r plot-ladder, fig.width = 8, fig.height = 5}
summary_df <- results |>
  group_by(case_prevalence, delta) |>
  summarise(auc_mean = mean(auc, na.rm = TRUE),
             auc_sd   = sd(auc,   na.rm = TRUE),
             n_reps   = sum(!is.na(auc)),
             .groups = "drop")

ggplot(summary_df, aes(x = delta, y = auc_mean)) +
  geom_ribbon(aes(ymin = pmax(0, auc_mean - auc_sd),
                   ymax = pmin(1, auc_mean + auc_sd)),
               alpha = 0.25, fill = "#B2182B") +
  geom_line(color = "#B2182B", linewidth = 1) +
  geom_point(size = 3, color = "#B2182B") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0.7, linetype = "dotted", color = "grey60") +
  scale_x_continuous(
    breaks = summary_df$delta,
    labels = sprintf("%.2f/%.2f (Δ=%.2f)",
                      summary_df$case_prevalence,
                      1 - summary_df$case_prevalence,
                      summary_df$delta)) +
  coord_cartesian(ylim = c(0.4, 1.0)) +
  labs(x = "case_prevalence / control_prevalence (delta)",
       y = "OOF AUC (mean ± SD)",
       title = sprintf("Prevalence-ladder AUC (%s, %s)",
                        params$catalog_name, params$model),
       subtitle = sprintf(
         "n=%d/%d, intensity_mult=%.2fx, %d biomarkers, %d seeds",
         params$n_per_class, params$n_per_class,
         params$case_intensity_mult, params$n_biomarkers, n_seeds)) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

# 4. Individual seed strip plot

```{r strip, fig.width = 7, fig.height = 4}
ggplot(results, aes(x = factor(case_prevalence), y = auc)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  geom_jitter(width = 0.15, height = 0, size = 2.5, alpha = 0.8,
               color = "#B2182B") +
  coord_cartesian(ylim = c(0.4, 1.0)) +
  labs(x = "case_prevalence",
       y = "OOF AUC (individual seeds)",
       title = "Per-seed AUC — spread indicates within-condition variability") +
  theme_minimal(base_size = 12)
```

# 5. Numeric table

```{r summary-table}
knitr::kable(summary_df |>
  mutate(across(c(auc_mean, auc_sd), ~round(., 3))),
  caption = sprintf("AUC by rung, %s model, n=%d per class",
                     params$model, params$n_per_class))
```

# Reading

- **AUC monotonically rises with delta** → the pipeline is signal-limited; larger effects are easier.
- **AUC saturates at some rung** → beyond that delta the model has learned the compound positions.
- **AUC drops below 0.7 at some rung** → practical detection limit for this cohort size / catalog / model combination.
- **Wide SD ribbon** → seed-sensitivity; increase `n_seeds`.

Once you have the ladder, sweep `n_per_class` (100 vs 200) or `n_biomarkers` (3 vs 5 vs 10) to see how sensitivity depends on sample count and signal spread.
